---
title: "PCA diagnostic loadings"
---

```{r}
#| include: false
source("../../../../R/core/style.R")
source("../../../../R/core/spatial.R")
source("../../../../R/domains/pathways.R")
source("../../../../R/figures/results/05-trajectory-robustness.R")
list2env(prepare_pca_data(), envir = environment())
```

## Lower-variance temporal modes

PC4 and PC5 are retained to document the first five fitted PCA modes. They
account for 5.7% and 4.6% of represented cell-trajectory variance,
respectively. Their loadings are not used to define the primary pathway
geometry because their omission stability is weaker than that of PC1 and the
PC2--PC3 subspace (Results 05).

> 本页保留 PC4–PC5 的时间载荷，便于核查前五个 PCA 模式。它们不进入主路径结论。

```{r}
#| label: fig-supp-pca-pc45-loadings
#| fig-cap: "Temporal loadings for PC4 and PC5. These lower-variance modes are retained for diagnostic completeness and are not primary pathway axes."
#| fig-width: 10
#| fig-height: 3.8
pc45_data <- loading_plot_data |> filter(component %in% levels(component)[4:5])
ggplot(pc45_data, aes(year, loading)) +
  geom_hline(yintercept = 0, colour = col_axis, linewidth = wid_axis) +
  geom_col(aes(fill = is_positive), width = .78, show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = pal_direction[["warming"]], "FALSE" = pal_direction[["cooling"]])) +
  facet_wrap(~component, nrow = 1) +
  scale_x_continuous(breaks = seq(1980, 2020, 5)) +
  labs(x = NULL, y = "Loading") +
  theme_base(grid = TRUE) + theme_facet(base_size = 9)
```

The figure is descriptive. PC labels and signs can change under equivalent
refits; the retained main-text inference concerns the stable leading axis and
the PC2--PC3 subspace, rather than a mechanism assigned to either lower mode.

> 图件为诊断性材料。主文仅对 PC1 与 PC2–PC3 平面作稳健性限定下的描述，不将低阶模式解释为机制。
