---
title: "Seasonal and ice-state context"
aliases:
  - "/explorations/warming-acceleration/draft/02-seasonal-ice-context.html"
  - "/explorations/warming-temporal-pathways/draft/02-seasonal-ice-context.html"
  - "/explorations/warming-temporal-pathways/manuscript/results/02-seasonal-ice-context.html"
---

```{r}
#| include: false
source("../../../../R/core/style.R")
source("../../../../R/domains/seasonal-ice.R")
seasonal_ice <- prepare_seasonal_ice_data()
```

Annual warming can follow different seasonal pathways. We compare endpoint-aligned trailing 10-year Sen rates for annual LSWT, seasonal LSWT, extreme-temperature and ice-day series. These are directional co-variation diagnostics, not additive contributions to annual warming.

> 年均增温可由不同季节路径构成。这里比较年均、四季、极端温度和冰日序列的端点对齐 10 年 Sen rate；它们描述方向性共变，不能相加为年均增温贡献。

```{r}
#| label: fig-results-seasonal-ice-rate-alignment
#| fig-cap: "Distribution of lake-level Spearman alignment between annual and diagnostic trailing 10-year Sen rates. Points are a deterministic visual sample of about one per 100 lakes; this is descriptive co-variation, not additive seasonal attribution."
ggplot(filter(seasonal_ice$alignment_long, is.finite(spearman_alignment)),
  aes(series, spearman_alignment)) +
  geom_hline(yintercept = 0, colour = col_grid, linewidth = wid_grid) +
  geom_violin(fill = scales::alpha(col_inchart_fill, .35), colour = col_inchart_line,
    linewidth = wid_inchart_line, trim = FALSE) +
  geom_boxplot(width = .15, outlier.shape = NA, fill = "white",
    colour = col_inchart_line, linewidth = wid_inchart_line) +
  geom_point(data = seasonal_ice$alignment_visual_sample,
    position = position_jitter(width = .13, height = 0, seed = 20260720),
    colour = col_text_soft, alpha = .16, size = .45) +
  coord_flip() +
  labs(x = NULL, y = "Lake Spearman alignment with annual rate") +
  theme_base(grid = TRUE)
```

Warm-season trends are most available and normally align more closely with annual local-rate histories. Winter and spring must be interpreted with ice state: repeated 0 °C days are a finite frozen state, not ordinary liquid-water observations. Therefore low cold-season alignment is an information boundary, not evidence that cold seasons are unimportant.

> 暖季趋势最完整，通常与年均局部速度历史更一致。冬春必须结合冰状态：反复出现的 0 °C 是冻结状态，不是普通液态温度。冷季对齐较弱主要是信息边界，不能解读为冷季不重要。

Ice duration supplies a second state diagnostic. Its association with annual local rates varies among lakes and regions. It does not establish glacier-meltwater forcing, nor解释每条降温轨迹。

> 冰期长度提供第二种状态诊断。它与年局部 rate 的关系因湖泊和地区而异；不证明冰川融水驱动，也不解释每条降温轨迹。

Detailed sign-conditioned comparisons and frozen-state limitations remain in [Seasonal and Ice Diagnostics](../../investigations/seasonal-ice-diagnostics.qmd).

> 按增温/降温分支的比较与冻结状态限制见 Seasonal and Ice Diagnostics。
