---
title: "Candidate Results tables"
aliases:
  - "/explorations/warming-temporal-pathways/manuscript/results/07-candidate-tables.html"
---

```{r}
#| include: false
source("../../../../R/core/style.R")
source("../../../../R/figures/results/climate-link-candidate-tables.R")
list2env(prepare_results_candidate_tables(), envir = environment())
emit_list_table <- function(frame, digits = 2) {
  frame <- frame |> mutate(across(where(is.numeric), ~formatC(.x, format = "f", digits = digits)))
  lines <- apply(rbind(names(frame), as.matrix(frame)), 1, function(row) paste0("- ", paste(paste0("- ", row), collapse = "\n  ")))
  knitr::asis_output(paste(c("::: {.list-table}", lines, ":::"), collapse = "\n"))
}
```

These tables are candidate supplementary material. They report descriptive
sample composition and PCA stability. They do not contain trend-significance,
formal group tests, or temporal-block-omission inference because those analyses
have not yet been defined and run.

> 本页是候选补充表，报告描述性样本组成和 PCA 稳健性。不含趋势显著性、正式组间检验或时间块遗漏推断，因为这些分析尚未定义并运行。

## Lake warming by continent

```{r}
#| results: asis
emit_list_table(continent_summary |> arrange(desc(n)))
```

## PCA variance spectrum

```{r}
#| results: asis
emit_list_table(pca_variance |> select(pc, explained_pct, cumulative_pct))
```

## LOCO subspace stability

```{r}
#| results: asis
emit_list_table(loco |> select(omitted_continent, subspace, min_cosine, max_angle_deg, mean_angle_deg, occupied_cells))
```
