---
title: "Association-pole trajectory composites"
---

```{r}
#| include: false
source("../../../../R/core/style.R")
source("../../../../R/domains/climate-links-seasonal.R")
source("../../../../R/domains/climate-links-selected.R")
tele_spatial <- prepare_selected_tele_spatial_fields()
tele_composites <- prepare_selected_tele_trajectory_composites(spatial = tele_spatial)
```

## Local-rate histories at the retained association poles

The comparison groups below are defined from the retained NAO/AO sensitivity
fields. They visualise contrasting trajectory histories after candidate
screening; they are not an independent replication or a circulation-to-lake
mechanism test.

> 本页以已保留的 NAO/AO 敏感性场定义两端群体，展示其轨迹历史。它不提供独立复现或环流到湖泊的机制证据。

```{r}
#| label: fig-supp-tele-speed-poles
#| fig-cap: "Trailing 10-year annual and JJA Sen-speed composites for negative and positive NAO/AO association poles. Ribbons show the within-pole interquartile range, not an uncertainty interval."
#| fig-width: 12
#| fig-height: 7
ggplot(tele_composites$speeds, aes(year, speed_mean, colour = sensitivity_pole, fill = sensitivity_pole)) +
  geom_hline(yintercept = 0, colour = col_grid, linewidth = wid_grid) +
  geom_ribbon(aes(ymin = speed_q25, ymax = speed_q75), alpha = .16, colour = NA) +
  geom_line(linewidth = wid_inchart_line) +
  facet_grid(representation ~ tele_index, scales = "free_y") +
  scale_colour_manual(values = c(
    "Negative-association pole" = pal_direction[["cooling"]],
    "Positive-association pole" = "#00A6B2"
  )) +
  scale_fill_manual(values = c(
    "Negative-association pole" = pal_direction[["cooling"]],
    "Positive-association pole" = "#00A6B2"
  )) +
  labs(x = "Endpoint year", y = "Trailing 10-year Sen speed (°C yr⁻¹)", colour = NULL, fill = NULL) +
  theme_base(grid = TRUE) +
  theme_facet(base_size = 9) +
  theme(legend.position = "bottom")
```

By 2020, the positive pole was close to 0.069 °C yr$^{-1}$, while the negative
poles were close to −0.014 °C yr$^{-1}$ for NAO and −0.018 °C yr$^{-1}$ for
AO. These numerical contrasts describe the retained association family and
should be read with its screening and stability boundary.

> 2020 年正关联极点约为 0.069 °C yr⁻¹；NAO 和 AO 的负关联极点约为 −0.014、−0.018 °C yr⁻¹。数值用于描述保留关联族，需与筛查和稳定性边界一起阅读。
