---
title: "Teleconnection candidate atlas"
---

```{r}
#| include: false
source("../../../../R/core/style.R")
source("../../../../R/domains/climate-links-lag-surface.R")
source("../../../../R/domains/climate-links-seasonal.R")
source("../../../../R/domains/climate-links-selected.R")
list2env(prepare_pca_seasonal_lag_surface(), envir = environment())
selected_tele_base <- prepare_selected_seasonal_teleconnection_data()
selected_tele_increment_base <- selected_tele_increment(selected_tele_base$cv_results) |>
  left_join(selected_tele_base$candidates, by = "response_season") |>
  mutate(candidate_label = paste0(response, " LSWT | ", predictor, " ", tele_index, " | lag ", lag_quarters, " quarters"))
```

## Complete descriptive lag screen

This atlas retains every planned seasonal and annual lag display before
manuscript selection. Cell-level median absolute correlation is a landscape
description. It does not select an optimal physical delay and does not provide
causal attribution.

> 本图集保留所有计划中的季节和年尺度 lag 图，供后续筛选。格网中位绝对相关只描述关联图景，不选择物理最优延迟，也不提供因果归因。

```{r}
#| label: fig-results-seasonal-lag-heatmap
#| fig-cap: "Median absolute equal-area-cell correlation across quarterly lags, by LSWT response season and climate index."
#| fig-width: 12
#| fig-height: 8
ggplot(seasonal_heatmap, aes(lag_label, index, fill = median_abs_r)) +
  geom_tile() + facet_wrap(~response_series, ncol = 1) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1, name = "Median |r|") +
  labs(x = "Quarter lag", y = NULL) + theme_base(grid = FALSE) + theme_facet(base_size = 8)
```

```{r}
#| label: fig-results-annual-lag-heatmap
#| fig-cap: "Median absolute equal-area-cell correlation of annual LSWT with seasonal climate indices at 0–3-year lags."
#| fig-width: 10
#| fig-height: 7
ggplot(annual_heatmap, aes(lag_label, predictor_season, fill = median_abs_r)) +
  geom_tile() + facet_wrap(~index, ncol = 2) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1, name = "Median |r|") +
  labs(x = "Year lag", y = "Index season") + theme_base(grid = FALSE) + theme_facet(base_size = 8)
```

## Spatially varying descriptive peak lag

Peak-lag colour indicates the lag with largest absolute cell association;
opacity indicates magnitude. Low-opacity colours have no clear dominant lag.
These maps test the descriptive proposition that lag structure differs across
space, without calling each colour a robust delayed response.

> 峰值 lag 颜色表示每个格网绝对关联最大处，透明度表示强度。低透明度颜色没有明确主导 lag。地图检验 lag 结构是否空间不同，不把每种颜色都称作稳健延迟响应。

```{r}
#| label: fig-results-tele-best-lag-maps
#| fig-cap: "Descriptive peak quarterly lag by response season and climate index. Opacity is maximum absolute cell correlation."
#| fig-width: 12
#| fig-height: 12
ggplot(best_lag) +
  geom_rect(aes(xmin = lon_min, xmax = lon_max, ymin = lat_min, ymax = lat_max, fill = best_lag, alpha = max_abs_r), colour = NA) +
  the_geom_coastline() + facet_grid(index ~ response_series) +
  scale_fill_viridis_d(option = "turbo", name = "Peak quarter lag") +
  scale_alpha_continuous(range = c(.12, 1), name = "Maximum |r|") +
  coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 85), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") + theme_map_grid(base_size = 7) + theme(legend.position = "bottom")
```

## PCA-coordinate additions across fixed candidates

The following comparison is a candidate screen, not a discovery replication:
candidates were chosen after inspecting the lag landscape. It nevertheless
shows explicitly whether PC1, PC2, PC3, their joint PC2–PC3 plane, or their
full combination adds held-out spatial prediction beyond geography and lake
morphology.

> 下图是候选筛查，不是独立发现重复，因为候选来自 lag 图景之后。它仍明确比较 PC1、PC2、PC3、PC2–PC3 联合平面及全组合，是否在地理和湖泊形态之外增加空间留出预测。

```{r}
#| label: fig-results-tele-pca-additions-all-candidates
#| fig-cap: "Held-out spatial-prediction increments for PCA-coordinate additions across all fixed seasonal climate candidates."
#| fig-width: 12
#| fig-height: 7
selected_tele_increment_base |>
  filter(addition %in% c("pc1_increment", "pc2_increment", "pc3_increment", "pc23_increment", "pc123_increment")) |>
  mutate(addition = recode(addition, pc1_increment = "+ PC1", pc2_increment = "+ PC2", pc3_increment = "+ PC3", pc23_increment = "+ PC2–PC3", pc123_increment = "+ PC1–PC3")) |>
  ggplot(aes(candidate_label, heldout_r2_increment, fill = addition)) +
  geom_hline(yintercept = 0, colour = col_grid, linewidth = wid_grid) +
  geom_col(position = position_dodge(width = .8), width = .72) + coord_flip() +
  labs(x = NULL, y = "Held-out R² increment", fill = NULL) + theme_base(grid = TRUE) + theme(legend.position = "bottom")
```
