---
title: "Widespread but heterogeneous lake warming"
---

```{r}
#| include: false
source("../../../../R/core/style.R")
source("../../../../R/core/spatial.R")
source("../../../../R/domains/kinematics.R")
source("../../../../R/figures/results/01-global-kinematics.R")
list2env(prepare_kinematics_data(), envir = environment())
list2env(prepare_results_kinematics_data(), envir = environment())
fmt_num <- function(x, digits = 2) formatC(x, format = "f", digits = digits)
```

## Long-term warming is widespread

:::{.callout-tip}
- Stress the spatial heterogeneity, since the warming states were already described in @tong2023.
- Here we use lake-equal mean, focusing on the lake individuals,
  but @tong2023 used area-weighted
  (2 layers:
  every grid is lake-area-weighted, and global is grid-area-weighted), focusing
  on a reasonable global state.
  So it is natural we are stressing the heterogeneity, by showing the range,
  quantiles and so on.
- Here we use Sen slope, but @tong2023 used linear slope.
  In supplementary we can diff the results to check which regions are
  "not that stable".
:::

Lake warming was widespread during 1981--2020,
although the magnitude of warming varied markedly among individual lakes
(@fig-long-term-warming-distribution).
Across all 92,245 lake records, 92.8% exhibited positive warming,
and 33.8% of the trends were significant at $p < 0.05$.
The median warming rate was 0.18 °C decade^-1^,
with an interquartile range of 0.09--0.31 °C decade^-1^;
whereas the lake-equal mean was 0.22 ± 0.19 °C decade^-1^.
The distribution has a positive high-warming tail,
in which a relatively small subset of rapidly warming lakes
contributes disproportionately to the lake-equal mean.

```{r}
#| label: fig-long-term-warming-distribution
#| fig-cap: "Long-term lake warming, 1981–2020. Top: Lake-equal mean 40-year Sen slope in each occupied 1° grid cell. Diagonal hatching shows cells where at least half of lakes display a significant (p < 0.05, Mann–Kendall) trend. Bottom: Comparison of the distribution of individual-lake Sen slopes (lake-equal, black) and a two-step area-weighted distribution, where annual temperatures are first lake-area weighted within each grid cell and cell slopes are then weighted by grid-cell area (blue)."
#| fig-width: 8
#| fig-height: 4
#| column: page

hatch_segments <- fig1_grid |>
  filter(hatch) |>
  transmute(
    x = lon_cell,
    y = lat_cell,
    xend = lon_cell + 1,
    yend = lat_cell + 1
  )

fig1_map <- ggplot(fig1_grid) +
  geom_tile(
    aes(lon_cell + .5, lat_cell + .5, fill = lake_equal_sen_decade),
    width = 1,
    height = 1
  ) +
  geom_segment(
    data = hatch_segments,
    aes(x = x, y = y, xend = xend, yend = yend),
    inherit.aes = FALSE,
    colour = "grey15",
    linewidth = .17,
    alpha = .7
  ) +
  the_geom_coastline() +
  scale_fill_gradient2(
    low = pal_direction[["cooling"]],
    mid = "grey92",
    high = pal_direction[["warming"]],
    limits = c(-.5, .5),
    oob = scales::squish,
    name = "Sen slope\n(°C decade⁻¹)"
  ) +
  scale_x_continuous(breaks = seq(-180, 180, 60), expand = c(0, 0)) +
  scale_y_continuous(breaks = seq(-60, 90, 30), expand = c(0, 0)) +
  coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 90), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") +
  theme_map_grid()

lake_stats <- quantile(
  long_term_distribution$warming,
  c(.25, .5, .75),
  na.rm = TRUE
)
area_stats <- weighted_summary
fig1_distribution <- ggplot() +
  geom_density(
    data = long_term_distribution,
    aes(warming),
    colour = col_text,
    linewidth = wid_inchart_line,
    adjust = 1
  ) +
  geom_density(
    data = weighted_cell_slopes,
    aes(cell_sen_slope_decade, weight = cell_area_weight),
    colour = col_text,
    linetype = "dashed",
    linewidth = wid_inchart_line,
    adjust = 1
  ) +
  geom_vline(
    xintercept = lake_stats[[1]],
    colour = "#2166AC",
    linewidth = .45
  ) +
  geom_vline(
    xintercept = lake_stats[[3]],
    colour = "#B2182B",
    linewidth = .45
  ) +
  geom_vline(
    xintercept = area_stats[[1]],
    colour = "#2166AC",
    linetype = "dashed",
    linewidth = .45
  ) +
  geom_vline(
    xintercept = area_stats[[3]],
    colour = "#B2182B",
    linetype = "dashed",
    linewidth = .45
  ) +
  annotate(
    "text",
    x = .78,
    y = 4.35,
    label = sprintf(
      "Lake-equal (solid): skewness %.2f",
      distribution_skewness[["lake_equal"]]
    ),
    hjust = 1,
    colour = col_text,
    size = 2.15
  ) +
  annotate(
    "text",
    x = .78,
    y = 4.05,
    label = sprintf(
      "Area-weighted (dashed): skewness %.2f",
      distribution_skewness[["area_weighted"]]
    ),
    hjust = 1,
    colour = col_text,
    size = 2.15
  ) +
  coord_cartesian(xlim = c(-.5, .8)) +
  labs(x = "Sen slope (°C decade⁻¹)", y = NULL) + theme_base(grid = FALSE) +
  theme(
    panel.background = element_rect(fill = "white", colour = NA),
  )

draw_fixed_aspect(
  ggdraw(fig1_map) +
  draw_plot(
    fig1_distribution,
    x = .04,
    y = .15,
    width = .25,
    height = .35
  ),
  aspect = 2
)
```

Long-term warming rates were spatially organized.
Relatively high warming rates occurred across Arctic Siberia,
particularly central and eastern Siberia (approximately 100--140°E),
with additional high-rate areas in northern Canada and Alaska.
Slower warming, including the relatively rare cooling lakes,
was concentrated across western Siberia and the adjacent Ural region,
northern Fennoscandia, and western-to-northwestern Canada
(@fig-long-term-warming-distribution).
Notably,
89% of lakes with cooling trends were located in Europe and North America.
These geographical patterns were strongly spatially autocorrelated
and were unlikely to arise from a spatially random distribution
(Moran’s I = 0.43, permutation P = 0.001; @tbl-spatial-autocorrelation).

> 长期增温率具有明确空间组织。高增温集中在北极西伯利亚中东部、加拿大北部和阿拉斯加；慢增温和少数降温湖集中在西西伯利亚—乌拉尔、芬诺斯堪的纳维亚北部与加拿大西至西北部。89% 降温湖位于欧洲和北美。空间分布显著非随机。

::: {#tbl-spatial-autocorrelation .list-table}

- - Statistic
  - Value
- - Spatial unit
  - Occupied 1° cells; lake-equal mean long-term Sen slope
- - Occupied cells
  - `r spatial_autocorrelation$n_cells`
- - Queen-neighbour pairs
  - `r spatial_autocorrelation$n_neighbour_pairs`
- - Moran’s I
  - `r fmt_num(spatial_autocorrelation$moran_i, 2)`
- - Two-sided permutation P
  - `r fmt_num(spatial_autocorrelation$permutation_p, 3)` (999 permutations)
:::

Warming rates also differed markedly among latitude bands.
Lakes at 60--85°N had a median warming rate of 0.258°C decade^-1^,
0.133°C decade^-1^ higher than lakes at 30--60°N
(0.125°C decade^-1^;
bootstrap 95% CI of the median difference: 0.131--0.135°C decade^-1^).
Continental distributions were uneven: Europe had the highest median rate
(0.261°C decade^-1^),
followed by North America (0.166°C decade^-1^),
whereas medians in Africa, Asia, Oceania and South America
ranged from 0.103 to 0.162°C decade^-1^.
Full distributions are shown in
@fig-long-term-warming-latitude-bands
and @fig-long-term-warming-continent-distributions.

```{r}
#| label: fig-long-term-warming-latitude-bands
#| fig-cap: "Long-term lake warming rates by latitude band. Violin width shows the within-band distribution; points show medians. Values are lake-equal and each lake contributes one observation."
#| fig-width: 8
#| fig-height: 4

ggplot(latitude_summary, aes(latitude_band, long_term_warming_decade)) +
  geom_hline(yintercept = 0, colour = col_text_soft, linewidth = wid_axis) +
  geom_violin(
    fill = col_inchart_fill, alpha = .45, colour = col_inchart_line,
    linewidth = wid_inchart_line, scale = "width"
  ) +
  stat_summary(
    fun = median, geom = "point", shape = 21,
    fill = pal_direction[["warming"]], size = 2
  ) +
  coord_cartesian(
    ylim = quantile(latitude_summary$long_term_warming_decade, c(.01, .99))
  ) +
  labs(x = NULL, y = "Long-term Sen slope (°C decade⁻¹)") +
  theme_base(grid = TRUE) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

```{r}
#| label: fig-long-term-warming-continent-distributions
#| fig-cap: "Long-term lake warming rates by continent. Violin width shows the distribution of sampled lakes; points show medians. Continental categories are descriptive geographic summaries, not causal groups."
#| fig-width: 8
#| fig-height: 4

ggplot(continent_summary, aes(Continent, long_term_warming_decade)) +
  geom_hline(yintercept = 0, colour = col_text_soft, linewidth = wid_axis) +
  geom_violin(
    fill = col_inchart_fill, alpha = .45, colour = col_inchart_line,
    linewidth = wid_inchart_line, scale = "width"
  ) +
  stat_summary(
    fun = median, geom = "point", shape = 21,
    fill = pal_direction[["warming"]], size = 2
  ) +
  coord_cartesian(
    ylim = quantile(continent_summary$long_term_warming_decade, c(.01, .99))
  ) +
  labs(x = NULL, y = "Long-term Sen slope (°C decade⁻¹)") +
  theme_base(grid = TRUE) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

Trend significance broadly followed warming magnitude.
Of lakes in the upper quartile of warming rates,
79.5% showed significant warming trends,
compared with 18.6% of remaining lakes.
Thus, stronger net warming was more likely to yield a detectable long-term trend,
although significant warming was not restricted to the fastest-warming regions.

<!-- Neither net warming magnitude nor trend significance identifies
when warming occurred within the 40-year record.
Lakes with similar long-term slopes can still follow
contrasting local-rate histories.
We therefore examined endpoint-aligned trailing-10-year Sen slopes
to resolve within-period changes in warming rate. -->
