---
title: "Climate-variability links"
aliases:
  - "/explorations/warming-acceleration/draft/04-teleconnection-sensitivity.html"
  - "/explorations/warming-temporal-pathways/draft/04-teleconnection-sensitivity.html"
---

```{r}
#| include: false
source("../../../../R/core/style.R")
source("../../../../R/domains/climate-links-seasonal.R")
source("../../../../R/domains/climate-links-selected.R")
tele_spatial <- prepare_selected_tele_spatial_fields()
tele_neighbors <- prepare_selected_tele_neighbor_diagnostics(tele_spatial$primary)
tele_composites <- prepare_selected_tele_trajectory_composites(spatial = tele_spatial)
```

## Exploratory screening identifies a retained summer association family

For each lake, raw liquid-water JJA LSWT is linearly detrended, then residual anomalies are correlated with predeclared circulation-index seasons. Candidate indices, seasons, and lags were screened before this focused comparison. Most combinations were weak, region-limited, or did not survive the geographic robustness sequence. We therefore report the retained prior-summer NAO/AO family and retain the rest as exploratory comparisons.

> 对每湖 raw 液态 JJA LSWT 做线性去趋势，再将残差与预设环流指数季节相关。候选指数、季节和滞后先经筛查；多数关联弱、区域受限或未通过地理稳健性序列。故正文报告保留的前年夏季 NAO/AO 关联族，其余仍为探索性比较。

We ask where this association differs, and whether its spatial field co-locates with PCA scores. This is descriptive association, not teleconnection attribution.

> 对每湖 raw 液态 JJA LSWT 线性去趋势，再将残差与预设环流指数季节相关。问题是关联在哪里不同、是否与 PCA 分数空间共定位。这是描述性关联，不是遥相关归因。

```{r}
#| label: fig-results-tele-primary-fields
#| fig-cap: "Signed Fisher-z association fields for JJA NAO/AO in year t−1 and JJA LSWT in year t. Purple/green encode sign, not warming/cooling: near-white cells are weak association, saturated cells are stronger signed association."
ggplot(tele_spatial$primary) +
  geom_rect(aes(xmin = lon_min, xmax = lon_max, ymin = lat_min, ymax = lat_max, fill = tele_fisher_z), colour = NA) +
  the_geom_coastline() + facet_wrap(~tele_index) +
  scale_fill_gradient2(low = "#542788", mid = "white", high = "#1B7837", name = "Fisher-z") +
  coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 85), expand = FALSE) + labs(x = "Longitude", y = "Latitude") +
  theme_map_grid(base_size = 8) + theme(legend.position = "bottom", legend.key.width = unit(24, "mm"))
```

Prior-summer NAO/AO and current-summer LSWT form strongest retained family. Fields are spatially continuous: neighbour correlations are 0.71 and 0.73 across 908 pairs. Strong opposite-sign neighbours are rare: NAO 5 of 213, AO 3 of 193. PDO, MAM NAO and DJF PDO remain regional or coverage-limited comparisons.

> 上一年夏季 NAO/AO 与当年夏季 LSWT 是最强保留关联族。场在 908 个邻接格网对中的相关为 0.71、0.73；强异号邻接很少，NAO 为 5/213，AO 为 3/193。PDO、MAM NAO 和 DJF PDO 仍只是区域性或覆盖受限比较。

PC1 adds little held-out prediction of this sensitivity field. PC2 and joint PC2--PC3 scores add stable skill across grid choices, continent omissions and decade omissions. Positive and negative association poles differ in PC2 and in late local warming-speed history: by 2020 positive poles are near 0.069 °C yr⁻¹; negative poles near −0.014 (NAO) and −0.018 (AO).

> PC1 对该敏感性场的空间留出预测几乎无增益；PC2 和联合 PC2–PC3 在格网、大陆删除与年代删除中保持稳定增益。关联正负端在 PC2 和后期局部增温速度历史上不同：2020 年正端约 0.069 °C yr⁻¹，负端约 −0.014（NAO）和 −0.018（AO）。

```{r}
#| label: fig-results-tele-speed-poles
#| fig-cap: "Trailing 10-year annual and JJA Sen-speed composites for negative and positive NAO/AO association poles."
#| fig-width: 12
#| fig-height: 7
ggplot(tele_composites$speeds, aes(year, speed_mean, colour = sensitivity_pole, fill = sensitivity_pole)) +
  geom_hline(yintercept = 0, colour = col_grid, linewidth = wid_grid) +
  geom_ribbon(aes(ymin = speed_q25, ymax = speed_q75), alpha = .16, colour = NA) +
  geom_line(linewidth = wid_inchart_line) + facet_grid(representation ~ tele_index, scales = "free_y") +
  labs(x = "Endpoint year", y = "Trailing 10-year Sen speed (°C yr⁻¹)", colour = NULL, fill = NULL) +
  theme_facet() + theme(legend.position = "bottom")
```

This co-location does not show NAO/AO causes PC2, nor identify circulation pathway. Grid, LOCO and LODO tests test stability, not independent discovery.

> 此共定位不表示 NAO/AO 造成 PC2，也不识别环流路径。格网、LOCO 和 LODO 检验稳定性，不是独立发现。
