---
title: ERA5 Association Scope and Missing Driver Data
---

```{r}
#| include: false
source("../../../shared/R/figure-style.R")
source("../../../shared/R/era5-association-scope-helpers.R")
list2env(prepare_era5_association_data(), envir = environment())
```

## Available first-pass inputs

The current ERA5 extraction provides lake-nearest-grid monthly and annual summaries for wind speed, surface pressure, and total precipitation over 1981--2020. These variables can be used for descriptive association checks with PCA scores and cooling cohorts; they are not a complete lake heat-budget dataset and cannot support attribution.

> 当前 ERA5 只有风速、地面气压、降水的湖泊最近网格月/年汇总。可先做 PCA 与降温 cohort 的描述性关联，不能构成完整热收支归因。

The first-pass model should report both a geography-only baseline and the incremental association after adding the available ERA5 variables. It should use spatially blocked validation, not only in-sample $R^2$, and should call all coefficients associations rather than effects.

> 首轮模型要并列地理基线与加入 ERA5 后的增量解释，并做空间分块验证；所有系数均称关联，不称效应。

```{r}
#| label: fig-era5-spatial-block-cv
#| fig-cap: "Spatial-block cross-validated R² for PC scores. Geography-only models are compared with geography plus available ERA5 variables; negative values indicate worse-than-mean held-out prediction."
#| fig-width: 7
#| fig-height: 3.8

ggplot(cv_results, aes(pc, spatial_block_cv_r2, fill = model)) +
  geom_hline(yintercept = 0, colour = col_text_soft, linewidth = wid_inchart_line) +
  geom_col(position = position_dodge(.7), width = .62) +
  scale_fill_manual(values = c("Geography only" = col_grid, "Geography + available ERA5" = pal_direction_fill[["warming"]])) +
  labs(x = NULL, y = "Spatial-block CV R²", fill = NULL) +
  theme_base(grid = TRUE) + theme(legend.position = "bottom")
```

@fig-era5-spatial-block-cv is deliberately a predictive association check. It asks whether wind, pressure, and precipitation summaries improve held-out spatial prediction beyond broad geography; it does not estimate physical forcing effects.

> @fig-era5-spatial-block-cv 是预测式关联检验：检验现有 ERA5 变量是否改善空间留出预测，不估计物理强迫效应。

## Missing variables to acquire before attribution

::: {.list-table}

- - Missing input
  - Why it matters
  - Minimum use
- - 2 m air temperature
  - Atmospheric thermal forcing and lake--air contrast.
  - Baseline climate trajectory.
- - Downward shortwave and longwave radiation
  - Direct surface energy input; cloud changes cannot be represented without them.
  - Heat-budget attribution.
- - Humidity, evaporation / latent heat flux
  - Evaporative cooling is a major candidate mechanism for LSWT response.
  - Energy-balance and mediation tests.
- - Wind components or mixing proxies
  - Wind speed alone cannot identify directional fetch or mixing regime.
  - Lake-specific mixing sensitivity.
- - Glacier inventory, upstream catchments, discharge and inflow temperature
  - Required to test glacier-connected cold inflow.
  - Glacier-meltwater hypothesis.
- - Lake depth, area, clarity / trophic state and catchment land cover
  - Thermal storage, optical absorption and watershed forcing are confounders/moderators.
  - Process-stratified modelling.
:::

> 缺失的关键变量包括气温、短/长波辐射、湿度/蒸发、混合代理、冰川—流域—入流链，以及湖泊与流域属性；补齐前不能做机制归因。

## Interpretation boundary

In an unadjusted first-pass rank correlation, PC1 and PC2 are moderately associated with mean surface pressure ($\rho=-0.33$ and $-0.35$), while PC3 is associated with the precipitation trend ($\rho=-0.33$). All other available annual ERA5 associations are small to modest. These are geographic co-patterns, not driver estimates: surface pressure in particular is strongly structured by elevation and regional circulation.

> 未调整的首轮秩相关中，PC1/PC2 与平均地面气压约为 -0.33/-0.35，PC3 与降水趋势约为 -0.33；其他关联较弱。这些是地理共变，尤其气压受海拔和区域环流强烈影响，不能当作驱动估计。

An association between an ERA5 trend and a PC score can identify a candidate co-varying pattern. It cannot distinguish direct forcing, mediation by radiation or evaporation, common geography, or retrieval artefacts. Glacier meltwater is a separate, untested hypothesis until hydrological connectivity is observed.

> ERA5 与 PC 的关联只能提出共同变化候选机制，不能区分直接强迫、中介、共同地理背景或反演伪影。冰川融水仍需水文连通证据。
