---
title: "Teleconnection-Sensitivity Screen for Secondary PCA Structure"
---

```{r}
#| label: setup-pca-teleconnection-screen
#| include: false
source("../../../shared/R/figure-style.R")
source("../../../shared/R/pca-teleconnection-helpers.R")
list2env(prepare_pca_teleconnection_display(), envir = environment())
```

## Question and design

This is a bounded external-coherence screen, not a teleconnection attribution analysis. For each lake, Step 10 first removes a linear trend from raw annual LSWT values, excludes the 0 °C frozen-state encoding, and correlates residual annual anomalies with a predeclared annual index. Correlations with at least 30 paired years are Fisher-$z$ transformed and aggregated to the same equal-area cells used by PCA. The spatial question is then whether fixed PC1, or the joint rank-exchanging PC2--PC3 score pair, improves held-out prediction of that cell sensitivity after continuous geography and lake morphology are already included.

> 这是受限的外部一致性筛查，不是遥相关归因。Step 10 先对每湖 raw 年均 LSWT 去线性趋势，排除 0 °C 冰冻编码，再与预先指定的年指数相关；至少 30 个配对年后 Fisher-$z$ 转换，并汇总至 PCA 等面积格网。问题是：在连续地理与湖泊形态已纳入后，固定 PC1 或可换序的 PC2–PC3 联合分数，能否提高对格网敏感性的空间留出预测。

The predeclared screen contains Niño 3.4, PDO, NAO, and AO at contemporaneous and one-year-leading positions. AMO is excluded: 40 years cannot resolve a multidecadal index. NAO and AO are treated as one correlated North Atlantic/Arctic circulation family, not independent mechanisms. PCA loading time series are never correlated directly with these indices. Positive index lead means index value in $t-1$ is paired with lake anomaly in $t$.

> 预先指定 Niño 3.4、PDO、NAO、AO 的同期与提前一年位置。AMO 不纳入：40 年不足以解析多年代指数。NAO 与 AO 作为相关的北大西洋/北极环流 family，不视为独立机制；不直接将 PCA loading 时间序列与指数相关。正 lead 表示 $t-1$ 年指数对应 $t$ 年湖泊异常。

## Main screen

```{r}
#| label: fig-pca-teleconnection-screen
#| fig-cap: "Spatial-block held-out R² increment beyond geography and lake morphology for PC1 or the joint PC2--PC3 score pair. Positive values improve prediction of a cell's Fisher-z transformed annual LSWT--index sensitivity. Each column uses contiguous blocks of a different size."
#| fig-width: 11
#| fig-height: 7

ggplot(screen_increment, aes(block_label, heldout_r2_increment, colour = addition, group = addition)) +
  geom_hline(yintercept = 0, colour = col_axis, linewidth = wid_axis) +
  geom_line(linewidth = wid_inchart_line) +
  geom_point(size = 1.8) +
  facet_grid(index ~ lag_years, scales = "free_y", labeller = label_both) +
  scale_colour_manual(values = c(
    "+ PC1" = col_text_soft,
    "+ PC2--PC3 subspace" = pal_direction[["warming"]]
  )) +
  labs(x = "Contiguous spatial block", y = "Held-out R² increment", colour = NULL) +
  theme_base(grid = TRUE) +
  theme_facet(base_size = 8) +
  theme(legend.position = "bottom")
```

At lag 0, adding PC2--PC3 raises held-out $R^2$ at every block scale for NAO, AO, PDO, and Niño 3.4. NAO rises from about $-0.08$--$0.01$ under geography/morphology to $0.25$--$0.30$ with PC2--PC3. PDO rises from $0.28$--$0.32$ to $0.45$--$0.48$. PC1 adds no stable gain. One-year leads do not show the same repeated PC2--PC3 pattern.

> 同期下，PC2–PC3 在 NAO、AO、PDO、Niño 3.4 的全部 block 尺度均提高留出 $R^2$。NAO 由地理/形态下约 $-0.08$–$0.01$ 升至 $0.25$–$0.30$；PDO 由 $0.28$–$0.32$ 升至 $0.45$–$0.48$。PC1 没有稳定增益；提前一年也未重现这种 PC2–PC3 模式。

This result has passed only first gate. A common temperature source can still create a spatial association: PCA uses low-frequency STL trajectories, while sensitivity uses detrended raw annual anomalies, but both originate from GLAST. The screen therefore identifies a reproducible cross-scale co-location, not an independent demonstration of forcing.

> 此结果只通过第一道门槛。PCA 使用低频 STL 轨迹，敏感性使用去趋势 raw 年异常，尺度不同但均来自 GLAST；共同数据源仍可造成空间关联。因此它识别可重复的跨尺度共定位，不是独立 forcing 证据。

## Independent spatial robustness checks

```{r}
#| label: fig-pca-teleconnection-grid-loco
#| fig-cap: "Robustness of held-out R² increments for contemporaneous NAO and PDO sensitivity. Left: three equal-area PCA grids with comparable physical block sizes. Right: target-grid leave-one-continent-out PCA refits; each point is one omitted continent."
#| fig-width: 12
#| fig-height: 5

grid_plot <- ggplot(grid_increment, aes(grid_label, heldout_r2_increment, colour = addition, group = addition)) +
  geom_hline(yintercept = 0, colour = col_axis, linewidth = wid_axis) +
  geom_line(linewidth = wid_inchart_line) +
  geom_point(size = 1.8) +
  facet_wrap(~index, nrow = 1) +
  scale_colour_manual(values = c(
    "+ PC1" = col_text_soft,
    "+ PC2--PC3 subspace" = pal_direction[["warming"]]
  )) +
  labs(x = "Equal-area grid", y = "Held-out R² increment", colour = NULL) +
  theme_base(grid = TRUE) + theme_facet(base_size = 8) + theme(legend.position = "bottom")

loco_plot <- ggplot(loco_increment, aes(omitted_continent, heldout_r2_increment, colour = addition, group = addition)) +
  geom_hline(yintercept = 0, colour = col_axis, linewidth = wid_axis) +
  geom_line(linewidth = wid_inchart_line) +
  geom_point(size = 1.8) +
  facet_wrap(~index, nrow = 1, scales = "free_y") +
  scale_colour_manual(values = c(
    "+ PC1" = col_text_soft,
    "+ PC2--PC3 subspace" = pal_direction[["warming"]]
  )) +
  labs(x = "Omitted continent", y = "Held-out R² increment", colour = NULL) +
  theme_base(grid = TRUE) + theme_facet(base_size = 8) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1), legend.position = "bottom")

grid_plot + loco_plot + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

The NAO/AO family and PDO pass the full gate. Their PC2--PC3 increments remain positive at 36 × 11, 72 × 21, and 144 × 42 equal-area grids. They also remain positive after every continent is omitted and PCA is refitted. Niño 3.4 fails this stricter test because its PC2--PC3 increment vanishes when North America is omitted. It is not promoted.

> NAO/AO family 与 PDO 通过完整门槛。PC2–PC3 增益在 36 × 11、72 × 21、144 × 42 格网均为正；去除任一大洲并重拟 PCA 后也为正。Niño 3.4 在去北美时增益消失，未通过更严格检验，不提升为结果。

## Spatial fields and interpretation limit

```{r}
#| label: fig-pca-teleconnection-sensitivity-maps
#| fig-cap: "Equal-area-cell Fisher-z transformed contemporaneous annual LSWT sensitivity to NAO, AO, and PDO. Values summarize lake-level correlations after each lake's linear annual trend is removed. These maps are sensitivity fields, not maps of causal forcing or statistical significance."
#| fig-width: 12
#| fig-height: 4.2

ggplot() +
  geom_rect(data = map_data, aes(xmin = lon_min, xmax = lon_max, ymin = lat_min, ymax = lat_max, fill = tele_fisher_z), colour = NA) +
  the_geom_coastline() +
  facet_wrap(~index, nrow = 1) +
  scale_fill_distiller(palette = "RdBu", direction = -1, name = "Fisher-z sensitivity") +
  coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 85), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") +
  theme_map_grid(base_size = 8) + theme(legend.position = "bottom")
```

The supported statement is narrow: cells that differ along the recurring PC2--PC3 low-frequency trajectory subspace also differ, beyond this geography/morphology baseline, in how detrended annual lake anomalies co-vary with contemporaneous North Atlantic/Arctic and North Pacific circulation indices. This gives PCA an external climate-consistency interpretation: secondary trajectory heterogeneity is geographically structured in the same places as interannual circulation sensitivity.

> 可支持的表述很窄：沿可重复 PC2–PC3 低频轨迹子空间不同的格网，在控制此处地理/形态基线后，对同期北大西洋/北极及北太平洋环流指数的去趋势年异常协变敏感性也不同。这给 PCA 一个外部气候一致性解释：次级轨迹异质性与年际环流敏感性在空间上同位。

It does not identify a unique index, establish a pathway, or show that NAO, AO, or PDO causes a PCA mode. NAO and AO are correlated. Step 10 uses ordinary Pearson correlations after linear detrending, without time-series autocorrelation correction; lake-level nominal $p$ values and post-hoc strongest-index labels are not used. Spatial block CV tests geographic generalization only. Seasonal index analysis is not opened yet: it requires an independent seasonal hypothesis and temporal-autocorrelation-aware inference.

> 它不识别唯一指数、不建立作用路径、不证明 NAO、AO 或 PDO 造成某个 PCA 模态。NAO/AO 相互相关；Step 10 为线性去趋势后的普通 Pearson 相关，未校正时间自相关；不使用逐湖名义 $p$ 值或事后最强指数标签。空间 block CV 只检验地理泛化。暂不扩展季节指数分析，除非先有独立季节假说与考虑时间自相关的推断方案。
