---
title: "External Interpretation of Spatially Balanced PCA"
---

```{r}
#| label: setup-pca-external-interpretation
#| include: false
source("../../../shared/R/figure-style.R")
source("../../../shared/R/pca-external-interpretation-helpers.R")
list2env(prepare_pca_external_interpretation_sensitivity(), envir = environment())
```

## Fixed question and predictor boundary

PCA identifies covariance structure and does not itself identify a driver. This diagnostic keeps the spatially balanced PCA fixed, aggregates every predictor to the same equal-area cells, and asks whether predeclared external predictor families improve spatially held-out prediction. PC1 is tested as a scalar response. PC2--PC3 is tested jointly by total squared prediction error across both scores, so rank exchange or an orthogonal rotation within their plane does not change the response definition. Raw long-term warming and raw warming-speed change are parallel outcomes.

> PCA 识别协变结构，不直接识别驱动。本诊断固定空间平衡 PCA，并将所有 predictor 汇总到同一等面积格网后做空间留出预测。PC1 为单响应；PC2–PC3 以两个 score 的总平方预测误差联合评估，故排序交换或平面内正交旋转不改变响应定义。raw 长期增温与 raw 增温速度变化为平行结果。

The first-pass panel is deliberately restricted to curated inputs already available: geography/lake morphology, annual ERA5-Land wind and precipitation summaries, and their JJA-minus-DJF contrasts and contrast trends. It excludes air temperature, radiation, humidity, evaporation, runoff, snow, and independent forcing data. ERA5-Land-driven FLAKE information also participates in GLAST reconstruction. Therefore every result below is predictive association, never heat-budget attribution.

> 第一轮只使用已整理输入：地理/湖泊形态、ERA5-Land 年风速与降水汇总，以及其 JJA 减 DJF 对比和对比趋势。气温、辐射、湿度、蒸发、径流、积雪、独立 forcing 数据均缺失；ERA5-Land 驱动 FLAKE 还参与 GLAST 重建。因此下列结果均为预测关联，不能称热收支归因。

## Spatial hold-out result

```{r}
#| label: fig-pca-external-cv
#| fig-cap: "Contiguous spatial-block held-out R² for nested external-association models. PC2--PC3 is evaluated jointly as a two-dimensional subspace. Negative values are worse than the held-out mean."
#| fig-width: 11
#| fig-height: 6

ggplot(cv_results, aes(block_label, spatial_block_cv_r2, colour = model, group = model)) +
  geom_hline(yintercept = 0, colour = col_axis, linewidth = wid_axis) +
  geom_line(linewidth = wid_inchart_line) +
  geom_point(size = 2) +
  facet_wrap(~outcome, ncol = 2, scales = "free_y") +
  scale_colour_manual(values = c(
    "Geography + morphology" = col_text_soft,
    "+ annual wind / precipitation" = pal_direction[["warming"]],
    "+ seasonal wind / precipitation contrast" = pal_direction[["cooling"]]
  )) +
  labs(x = "Contiguous spatial block", y = "Held-out R²", colour = NULL) +
  theme_base(grid = TRUE) +
  theme_facet(base_size = 9) +
  theme(legend.position = "bottom")
```

For PC1, geography/morphology yields held-out $R^2=0.247$--$0.289$. Adding annual wind and precipitation raises it to $0.318$--$0.377$ at all three block scales. Adding JJA-minus-DJF contrast terms adds no further gain. Raw long-term warming has only a small annual-forcing increment ($0.012$--$0.015$), and the seasonal extension removes it. Thus the current panel predicts part of PC1 score geography, but does not provide a matched and robust explanation of raw long-term warming.

> PC1 中，地理/形态模型留出 $R^2=0.247$–$0.289$；加入年风速与降水后，在三种 block 尺度升至 $0.318$–$0.377$。再加入 JJA 减 DJF 对比项没有增益。raw 长期增温的年 forcing 增益仅为 $0.012$–$0.015$，季节扩展后消失。因此当前面板可预测部分 PC1 空间 score，却不能稳健解释 raw 长期增温。

PC2--PC3 joint-subspace held-out $R^2$ is near zero or negative for every model and block scale. Raw warming-speed change is likewise near zero under geography/morphology and becomes more negative after either forcing extension. This fails the predeclared evidence gate: no available predictor family gives positive, cross-scale held-out gain for both the secondary PCA structure and its parallel raw rate outcome.

> PC2–PC3 联合子空间在所有模型、全部 block 尺度的留出 $R^2$ 接近零或为负。raw 增温速度变化在地理/形态下同样接近零，加入任一 forcing 扩展后更负。它未通过预设门槛：现有 predictor family 未能同时为次级 PCA 结构及其平行 raw 速度结果提供跨尺度正增益。

## Consequence

The correct conclusion is limited. Current wind and precipitation summaries are insufficient to interpret the secondary PCA structure. PC1's restricted association is retained as a data-availability result, not a driver claim, because the raw warming parallel check is weak and key heat-budget terms are missing. Further explanatory work requires a new, physically complete and provenance-aware predictor panel before another model family is added.

> 结论应受限：现有风速与降水汇总不足以解释次级 PCA 结构。PC1 的受限关联只保留为数据可用性结果，不称驱动；raw 增温平行检验弱，且关键热收支项缺失。后续解释工作须先建立物理完整、注明数据生成链的 predictor 面板，再新增模型。
