---
title: "Selected seasonal teleconnection candidates"
format: html
execute:
  echo: false
  warning: false
---

```{r}
#| label: setup
source("../../../shared/R/figure-style.R")
source("../../../shared/R/pca-seasonal-teleconnection-helpers.R")
source("../../../shared/R/pca-selected-seasonal-teleconnection-helpers.R")
selected_tele_spatial <- prepare_selected_tele_spatial_fields()
selected_tele_neighbors <- prepare_selected_tele_neighbor_diagnostics(selected_tele_spatial$primary)
selected_tele_composites <- prepare_selected_tele_trajectory_composites(
  spatial = selected_tele_spatial
)
selected_tele_base <- selected_tele_spatial$payload
selected_tele_increment_base <- selected_tele_increment(selected_tele_base$cv_results) |>
  left_join(selected_tele_base$candidates, by = "response_season") |>
  mutate(candidate_label = paste0(
    response, " LSWT | ", predictor, " ", tele_index,
    " | lag ", lag_quarters, if_else(lag_quarters == 1, " quarter", " quarters")
  ))
pc23_label <- "pc23_increment"
```

## Why this validation exists

Step 18 maps a broad lag surface. Its median absolute cell correlation is a display statistic, not a candidate-ranking statistic: a modest local association can still have a spatial field that aligns strongly with PCA scores. Step 19 therefore fixes seven interpretable season/index/lag candidates and re-estimates lake-level detrended correlations before cell Fisher-z aggregation.

> Step 18 的中位绝对相关仅用于展示，不能用于候选排序。Step 19 固定 7 个具有明确季节和滞后含义的候选，在湖泊层面重算去趋势相关，再聚合为格网 Fisher-z 场。

For every candidate, held-out prediction compares geography and lake morphology alone with additions of PC1, PC2, PC3, and their combinations. A positive increment means that a PCA score pattern helps predict where the lake-temperature sensitivity field is stronger or weaker; it is not a causal effect of the index on lake temperature.

> 每个候选均比较仅地理/湖泊背景模型与加入各 PC 后的空间留出预测。正增量表示 PCA 空间格局有助于预测敏感性场的位置差异，不表示遥相关对湖温的因果效应。

## Baseline score additions

```{r}
#| label: fig-selected-tele-baseline
#| fig-cap: "Held-out spatial-prediction increments at the reference equal-area grid."
selected_tele_increment_base |>
  filter(addition %in% c("pc1_increment", "pc2_increment", "pc3_increment", "pc23_increment", "pc123_increment")) |>
  mutate(addition = recode(addition,
    pc1_increment = "+ PC1", pc2_increment = "+ PC2", pc3_increment = "+ PC3",
    pc23_increment = "+ PC2--PC3", pc123_increment = "+ PC1--PC3"
  )) |>
  ggplot(aes(candidate_label, heldout_r2_increment, fill = addition)) +
  geom_hline(yintercept = 0, colour = col_grid, linewidth = wid_grid) +
  geom_col(position = position_dodge(width = .8), width = .72) +
  coord_flip() +
  labs(x = NULL, y = "Held-out R² increment", fill = NULL) +
  theme_base() + theme(legend.position = "bottom")
```

## Stability checks

The full validation retains only the PC2--PC3 joint addition. It separates an association that stays positive after changing grid resolution and omitting individual decades from one that depends mainly on a single continent. This is a stability screen, not an independent replication because candidates were selected after inspecting Step 18. The heavy grid/LOCO/LODO panel is generated outside the render path because the underlying LODO table contains 2.58 million lake-candidate rows.

> 验证重点是 PC2--PC3 联合增量：跨格网、跨十年仍为正的关联，与主要依赖某一大陆的关联必须分开。候选来自 Step 18 之后，因此这不是独立重复验证。LODO 原表含 258 万行，完整稳定性图在页面外生成，避免渲染阶段反复读取大表。

## Spatial heterogeneity of retained NAO/AO fields

The retained NAO/AO family does not have one global sign. Adjacent cells can have opposite Fisher-z associations: a positive value means warmer detrended JJA anomalies tend to accompany a positive index in the prior JJA; a negative value means the reverse. Geography and lake morphology explain part of this layout, but the residual fields still align with PC2 and PC3.

> 保留的 NAO/AO 关联并非全球同号。相邻格网可有相反 Fisher-z：正值表示当年夏季去趋势湖温偏暖，往往对应上一年夏季指数为正；负值则相反。地理与湖泊形态解释部分格局，但残差场仍与 PC2、PC3 对齐。

### Apparent local sign reversals

Most visible sign boundaries are not neighbouring strong but opposite responses. Across 908 east/north adjacent cell pairs, raw-field neighbour correlations are 0.71 for NAO and 0.73 for AO. At the conservative threshold where both cells have `|Fisher-z| >= 0.15`, only 5 of 213 NAO pairs and 3 of 193 AO pairs have opposite signs. The map therefore contains spatially smooth sensitivity fields with weak-field zero crossings, plus a small number of genuine strong boundaries.

> 大部分颜色交界并非两个相邻区域出现强而相反的响应。908 个东西/南北邻接格网对中，原始场邻接相关：NAO 为 0.71，AO 为 0.73。当两格均满足 `|Fisher-z| >= 0.15`，NAO 仅 213 对中的 5 对、AO 仅 193 对中的 3 对异号。地图主要是连续敏感性场跨越弱零线，只有少数真正强边界。

```{r}
#| label: fig-selected-tele-primary-fields
#| fig-cap: "Signed lake-temperature sensitivity fields for the retained prior-summer NAO/AO family."
#| fig-width: 12
#| fig-height: 4.5
ggplot(selected_tele_spatial$primary) +
  geom_rect(aes(xmin = lon_min, xmax = lon_max, ymin = lat_min, ymax = lat_max,
    fill = tele_fisher_z), colour = NA) +
  the_geom_coastline() +
  facet_wrap(~tele_index, nrow = 1) +
  scale_fill_distiller(palette = "RdBu", direction = -1,
    name = "Fisher-z association") +
  coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 85), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") +
  theme_map_grid(base_size = 8) + theme(legend.position = "bottom")
```

The NAO and AO fields agree as a family rather than as identical indices: their cell-wise Spearman correlation is 0.68 for the raw fields and 0.69 after removing the geography/morphology baseline. Their positive and negative regions remain spatially mixed. This supports family-level replication of a heterogeneous sensitivity field, not a globally uniform NAO/AO response.

> NAO 与 AO 场作为同一环流家族具有一致性，但并非相同指数：原始格网场 Spearman 为 0.68；去除地理/形态背景后为 0.69。正、负区域仍相互交错，支持“异质敏感性场的家族复现”，不支持全球同向响应。

```{r}
#| label: fig-selected-tele-primary-residuals
#| fig-cap: "Residual NAO/AO sensitivity after continuous geography and lake-morphology adjustment."
#| fig-width: 12
#| fig-height: 4.5
ggplot(selected_tele_spatial$primary) +
  geom_rect(aes(xmin = lon_min, xmax = lon_max, ymin = lat_min, ymax = lat_max,
    fill = geography_residual), colour = NA) +
  the_geom_coastline() +
  facet_wrap(~tele_index, nrow = 1) +
  scale_fill_distiller(palette = "RdBu", direction = -1,
    name = "Residual Fisher-z") +
  coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 85), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") +
  theme_map_grid(base_size = 8) + theme(legend.position = "bottom")
```

```{r}
#| label: fig-selected-tele-pc2-map
#| fig-cap: "PC2 score map on the same equal-area cells as the retained NAO/AO sensitivity fields."
#| fig-width: 12
#| fig-height: 4.5
ggplot(selected_tele_spatial$pc2) +
  geom_rect(aes(xmin = lon_min, xmax = lon_max, ymin = lat_min, ymax = lat_max,
    fill = pc2), colour = NA) +
  the_geom_coastline() +
  scale_fill_distiller(palette = "RdBu", direction = -1, name = "PC2 score") +
  coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 85), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") +
  theme_map_grid(base_size = 8) + theme(legend.position = "bottom")
```

## Trajectory context of the sensitivity field

The poles describe the ends of a continuous signed association field. They are not response classes and do not denote association magnitude. In both NAO and AO maps, positive-association cells have higher mean PC2 scores, while negative-association cells have lower PC2 scores. Their raw annual and JJA trajectories then separate mainly through changing local warming speed: the positive-association pole retains or increases positive late-period speed, whereas the negative-association pole approaches near-zero or negative late-period speed.

> 两端只是连续有符号关联场的展示，不是响应类别，也不表示关联强弱。NAO 和 AO 中，正关联端平均 PC2 更高，负关联端 PC2 更低。两端 raw 年均与 JJA 轨迹的主要分离来自局部增温速度变化：正关联端后期仍保持或提高正速度，负关联端后期接近零或转负。

```{r}
#| label: fig-selected-tele-trajectory-composites
#| fig-cap: "Raw annual and JJA LSWT anomaly composites at negative and positive association poles of the NAO/AO fields. Lines are equal-cell means; ribbons show cell interquartile ranges."
#| fig-width: 12
#| fig-height: 7
ggplot(selected_tele_composites$trajectories,
  aes(year, anomaly_mean, colour = sensitivity_pole, fill = sensitivity_pole)) +
  geom_hline(yintercept = 0, colour = col_grid, linewidth = wid_grid) +
  geom_ribbon(aes(ymin = anomaly_q25, ymax = anomaly_q75), alpha = .16, colour = NA) +
  geom_line(linewidth = wid_inchart_line) +
  facet_grid(representation ~ tele_index, scales = "free_y") +
  labs(x = NULL, y = "Baseline-centred LSWT anomaly (°C)", colour = NULL, fill = NULL) +
  theme_facet() + theme(legend.position = "bottom")
```

```{r}
#| label: fig-selected-tele-speed-composites
#| fig-cap: "Trailing 10-year Sen-speed composites at negative and positive association poles of the NAO/AO fields."
#| fig-width: 12
#| fig-height: 7
ggplot(selected_tele_composites$speeds,
  aes(year, speed_mean, colour = sensitivity_pole, fill = sensitivity_pole)) +
  geom_hline(yintercept = 0, colour = col_grid, linewidth = wid_grid) +
  geom_ribbon(aes(ymin = speed_q25, ymax = speed_q75), alpha = .16, colour = NA) +
  geom_line(linewidth = wid_inchart_line) +
  facet_grid(representation ~ tele_index, scales = "free_y") +
  labs(x = "Endpoint year", y = "Trailing 10-year Sen speed (°C yr⁻¹)", colour = NULL, fill = NULL) +
  theme_facet() + theme(legend.position = "bottom")
```

## Current interpretation boundary

The prior-summer NAO/AO to current-summer lake-temperature field is the strongest retained family: PC1 adds almost nothing, whereas PC2--PC3 adds positive held-out skill at every tested grid, continent omission, and decade omission. It supports a narrower result: the secondary low-frequency trajectory geometry, especially PC2, co-locates with heterogeneous sensitivity to the NAO/AO family. It does not show that NAO or AO produces PC2 or PC3.

> 上一年夏季 NAO/AO 与当年夏季湖温敏感性场，是目前最稳的关联族：PC1 几乎没有增益，PC2--PC3 在全部格网、剔除大陆、剔除十年测试中保持正增益。它支持“次级低频轨迹几何，特别是 PC2，与 NAO/AO 敏感性空间异质性共定位”；不支持 NAO/AO 产生 PC2 或 PC3 的因果结论。

The spring-to-summer NAO field and annual PDO fields also carry PC2-related signal, but both weaken sharply when North America is omitted. DJF PDO remains a cold-season exploratory observation because its lake coverage is restricted by frozen states. These branches remain descriptive and should not be elevated to the manuscript's central claim.

> 春季到夏季 NAO 场和年均 PDO 场也有 PC2 信号，但剔除北美后显著减弱。DJF PDO 受冰期限制，有效湖泊覆盖较窄；都只能作为描述性结果，不能进入文章核心结论。
