---
title: "Lake-Context Exclusion Check for Seasonal Teleconnection Association"
---

```{r}
#| label: setup-pca-jja-tele-lake-context
#| include: false
source("../../../shared/R/figure-style.R")
source("../../../shared/R/pca-seasonal-teleconnection-helpers.R")
list2env(prepare_pca_jja_lake_context_display(), envir = environment())
```

## Question

The retained JJA NAO/AO lag-1 result already controls area, mean depth, elevation, and distance to coast. This check asks whether its PC2--PC3 association is instead a proxy for further static lake or watershed context already present in project metadata.

> 保留的 JJA NAO/AO 滞后一年结果已控制面积、平均深度、高程和离海距离。本检验问：PC2–PC3 关联是否只是项目元数据中其他静态湖泊/流域背景的代理。

The expanded context contains log residence time, discharge, watershed area, volume, shoreline development, local terrain slope, and reservoir fraction. `Res_time = -1` and `Slope_100 = -1` are metadata missing codes. Every model uses the same 568 equal-area cells with complete context, so differences cannot result from changing lake coverage.

> 扩展背景包括对数停留时间、流量、流域面积、体积、岸线发育度、局地坡度和水库比例。`Res_time = -1`、`Slope_100 = -1` 是元数据缺失编码。所有模型使用相同的 568 个 context-complete 等面积格网，差异不来自样本覆盖变化。

## Target-grid comparison

```{r}
#| label: fig-pca-jja-tele-lake-context
#| fig-cap: "Spatial-block held-out R² after adding existing static lake and watershed context. Expanded context does not remove PC2--PC3 gain for either JJA NAO or AO lag-1 association field. All lines use 568 context-complete cells."
#| fig-width: 12
#| fig-height: 6.5

ggplot(model_comparison,
  aes(model, spatial_block_cv_r2, colour = block_label, group = block_label)) +
  geom_hline(yintercept = 0, colour = col_axis, linewidth = wid_axis) +
  geom_line(linewidth = wid_inchart_line) +
  geom_point(size = 1.7) +
  facet_wrap(~index, nrow = 1) +
  scale_colour_manual(values = c("4 × 3 bins" = col_text_soft,
    "6 × 3 bins" = pal_direction[["warming"]], "8 × 3 bins" = col_inchart_line)) +
  scale_x_discrete(labels = c("Geo +\ncore lake", "+ PC1", "+ PC2\n+ PC3",
    "Geo +\nfull context", "+ PC1\nafter context", "+ PC2 + PC3\nafter context",
    "+ all PCs\nafter context")) +
  labs(x = NULL, y = "Spatial-block held-out R²", colour = "Block layout") +
  theme_base(grid = TRUE) + theme_facet(base_size = 8) +
  theme(axis.text.x = element_text(size = 7), legend.position = "bottom")
```

Expanded context alone does not predict either association field under spatial hold-out. At reference blocks, NAO changes from $-0.087$ with core lake context to $-0.115$ with expanded context; AO changes from $-0.141$ to $-0.146$. PC2--PC3 after expanded context remains positive: $0.290$ for NAO and $0.183$ for AO. PC1 still gives no gain.

> 扩展背景本身在空间留出下不能预测两种关联场。参考 blocks 中，NAO 从核心湖泊背景的 $-0.087$ 变为扩展背景的 $-0.115$；AO 从 $-0.141$ 变为 $-0.146$。扩展背景后 PC2–PC3 仍为正：NAO $0.290$、AO $0.183$；PC1 仍无增益。

## Stability of the exclusion

```{r}
#| label: fig-pca-jja-tele-context-validation
#| fig-cap: "PC2--PC3 held-out R² increment over expanded lake context. Positive increments persist across three equal-area grids, leave-one-continent-out PCA refits, and leave-one-decade-out association calculations."
#| fig-width: 12
#| fig-height: 7

ggplot(validation_increment,
  aes(label, pc23_increment, colour = index, group = index)) +
  geom_hline(yintercept = 0, colour = col_axis, linewidth = wid_axis) +
  geom_line(linewidth = wid_inchart_line) +
  geom_point(size = 1.8) +
  facet_wrap(~validation, ncol = 1, scales = "free_x") +
  scale_colour_manual(values = c("NAO" = pal_direction[["warming"]], "AO" = col_inchart_line)) +
  labs(x = NULL, y = "PC2--PC3 held-out R² increment", colour = NULL) +
  theme_base(grid = TRUE) + theme_facet(base_size = 8) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1), legend.position = "bottom")
```

All expanded-context increments remain positive across grids, continent deletions, and decade omissions. Weakest case is NAO after North America is omitted: absolute held-out $R^2$ is slightly negative ($-0.011$), yet it exceeds expanded-context baseline by $0.123$. This excludes one specific alternative only: currently available static lake and watershed context does not reduce the retained association to a geography-plus-morphology proxy.

> 扩展背景后，所有格网、去大洲、去年代增益均为正。最弱是 NAO 去北美：绝对留出 $R^2$ 略负（$-0.011$），但仍比扩展背景高 $0.123$。它只排除一个特定替代解释：现有静态湖泊/流域背景不能将保留关联化约为地理加形态代理。

This remains an exclusion check, not a mechanism. Lake context has been averaged within cells and cannot represent seasonal inflow, mixing, or heat exchange. It does not establish that circulation is the missing cause; it narrows what the PC2--PC3 signal is not explained by.

> 这仍是排除检验，不是机制。湖泊背景已在格网内平均，不能代表季节入流、混合或热交换；它不证明环流是缺失原因，只缩小 PC2–PC3 信号不能被什么解释。
