---
title: "Seasonal Teleconnection Lag Surface"
---

```{r}
#| label: setup-pca-seasonal-lag-surface
#| include: false
source("../../../shared/R/figure-style.R")
source("../../../shared/R/pca-seasonal-lag-surface-helpers.R")
list2env(prepare_pca_seasonal_lag_surface(), envir = environment())
```

## Descriptive lag landscape

Step 18 expands the fixed same-season screen into a complete temporal landscape. For each equal-area cell, it first averages lake-level linearly detrended residual LSWT, then correlates that cell-mean residual series with the index. Seasonal responses span quarter lags 0--12; annual responses pair annual LSWT with each predictor season at lags 0--3 years. This is not Step 17's lake-level Fisher-z estimand.

> Step 18 将固定同季筛查扩展为完整时间图谱。每个等面积格网先平均逐湖线性去趋势 residual LSWT，再与指数相关。季节响应覆盖 0–12 个季度 lag；年均响应与每个指数季节做 0–3 年 lag。它不是 Step 17 的逐湖 Fisher-z estimand。

```{r}
#| label: fig-seasonal-lag-heatmap
#| fig-cap: "Median absolute equal-area-cell correlation across quarterly lags. Each panel is a response season; each row is an index. Lag 0 is matching season, lag 1 is preceding season, lag 4 is matching season in previous year. Magnitude does not establish direction, pathway, or an optimal physical lag."
#| fig-width: 12
#| fig-height: 8

ggplot(seasonal_heatmap, aes(lag_label, index, fill = median_abs_r)) +
  geom_tile() +
  facet_wrap(~response_series, ncol = 1) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1, name = "Median |r|") +
  labs(x = "Quarter lag", y = NULL) +
  theme_base(grid = FALSE) + theme_facet(base_size = 8)
```

The heatmap is a complete screen, not a ranking device. Broad ridges across adjacent lags indicate persistence or a seasonally extended association; isolated peaks require later lake-level validation. It is therefore used to identify temporal structures worth describing, not to claim that the largest cell correlation is a transport time.

> heatmap 是完整筛查，不是排行榜。相邻 lag 的连续高值可表示持续性或跨季关联；孤立峰值必须做逐湖验证。它用于识别值得描述的时间结构，不把最大相关直接解释为传播时间。

```{r}
#| label: fig-annual-lag-heatmap
#| fig-cap: "Median absolute equal-area-cell correlation of annual LSWT with each seasonal index at 0--3 year lags. Each panel is an index; predictor season remains explicit because annual LSWT has no single seasonal phase."
#| fig-width: 10
#| fig-height: 7

ggplot(annual_heatmap, aes(lag_label, predictor_season, fill = median_abs_r)) +
  geom_tile() +
  facet_wrap(~index, ncol = 2) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1, name = "Median |r|") +
  labs(x = "Year lag", y = "Index season") +
  theme_base(grid = FALSE) + theme_facet(base_size = 8)
```

## Cell-wise descriptive peak lag

For each index and response season, each cell receives the quarter lag at which its absolute cell-mean association is largest. Colour marks this selected lag; opacity marks its magnitude. A low-opacity cell has no clear dominant lag even though a colour is assigned. Maps therefore reveal regional differences in the *descriptive peak* without treating every peak as a robust delayed response.

> 对每个指数、响应季节，每格取绝对格网关联最大的季度 lag。颜色表示该 lag，透明度表示其强度。低透明度格网没有清晰主导 lag，即使仍被赋色。地图展示区域间“描述性峰值”差异，不把每个峰值当作稳健延迟响应。

```{r}
#| label: fig-nao-best-lag-map
#| fig-cap: "NAO descriptive peak quarter lag by response season. Opacity is maximum absolute cell correlation."
#| fig-width: 12
#| fig-height: 6

ggplot(filter(best_lag, index == "NAO")) +
  geom_rect(aes(xmin = lon_min, xmax = lon_max, ymin = lat_min, ymax = lat_max,
    fill = best_lag, alpha = max_abs_r), colour = NA) +
  the_geom_coastline() + facet_wrap(~response_series, nrow = 1) +
  scale_fill_viridis_d(option = "turbo", name = "Peak quarter lag") +
  scale_alpha_continuous(range = c(.12, 1), name = "Maximum |r|") +
  coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 85), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") + theme_map_grid(base_size = 8) +
  theme(legend.position = "bottom")
```

```{r}
#| label: fig-ao-best-lag-map
#| fig-cap: "AO descriptive peak quarter lag by response season. Opacity is maximum absolute cell correlation."
#| fig-width: 12
#| fig-height: 6

ggplot(filter(best_lag, index == "AO")) +
  geom_rect(aes(xmin = lon_min, xmax = lon_max, ymin = lat_min, ymax = lat_max,
    fill = best_lag, alpha = max_abs_r), colour = NA) +
  the_geom_coastline() + facet_wrap(~response_series, nrow = 1) +
  scale_fill_viridis_d(option = "turbo", name = "Peak quarter lag") +
  scale_alpha_continuous(range = c(.12, 1), name = "Maximum |r|") +
  coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 85), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") + theme_map_grid(base_size = 8) +
  theme(legend.position = "bottom")
```

```{r}
#| label: fig-pdo-best-lag-map
#| fig-cap: "PDO descriptive peak quarter lag by response season. Opacity is maximum absolute cell correlation."
#| fig-width: 12
#| fig-height: 6

ggplot(filter(best_lag, index == "PDO")) +
  geom_rect(aes(xmin = lon_min, xmax = lon_max, ymin = lat_min, ymax = lat_max,
    fill = best_lag, alpha = max_abs_r), colour = NA) +
  the_geom_coastline() + facet_wrap(~response_series, nrow = 1) +
  scale_fill_viridis_d(option = "turbo", name = "Peak quarter lag") +
  scale_alpha_continuous(range = c(.12, 1), name = "Maximum |r|") +
  coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 85), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") + theme_map_grid(base_size = 8) +
  theme(legend.position = "bottom")
```

```{r}
#| label: fig-nino-best-lag-map
#| fig-cap: "Niño 3.4 descriptive peak quarter lag by response season. Opacity is maximum absolute cell correlation."
#| fig-width: 12
#| fig-height: 6

ggplot(filter(best_lag, index == "Nino34")) +
  geom_rect(aes(xmin = lon_min, xmax = lon_max, ymin = lat_min, ymax = lat_max,
    fill = best_lag, alpha = max_abs_r), colour = NA) +
  the_geom_coastline() + facet_wrap(~response_series, nrow = 1) +
  scale_fill_viridis_d(option = "turbo", name = "Peak quarter lag") +
  scale_alpha_continuous(range = c(.12, 1), name = "Maximum |r|") +
  coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 85), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") + theme_map_grid(base_size = 8) +
  theme(legend.position = "bottom")
```
