---
title: Seasonal, Extreme-Temperature, and Ice Dynamics
---

## Question and boundary

This module asks how each lake's changing local annual warming rate co-varies through time with seasonal rates, warm/cold 30-day temperature rates, and ice-duration state. It does not begin by classifying lakes as full-period, late-period, or decelerating cooling types.

> 本模块考察每个湖的局部年增温率如何随时间与季节速率、最暖/最冷 30 天温度速率及冰期状态共同变化。不以全期、末段或减速降温标签作为入口。

The module is descriptive. A synchronised seasonal rate is not a formal contribution to annual warming, because Theil--Sen slopes are not additive across annual and seasonal series. A temperature--ice correlation is not a causal response estimate, and neither pattern identifies glacier meltwater.

> 本模块是描述性的。季节 Sen slope 不能相加为年均 slope，故同步变化不是严格贡献；温度—冰期相关不是因果响应，也不能识别冰川融水。

## Common endpoint axis

All temperature series use a trailing 10-year Theil--Sen slope, indexed to its endpoint year. The 1981--1990 slope is stored at 1990, producing a common 1990--2020 local-rate axis for annual, DJF, MAM, JJA, SON, annual maximum 30-day temperature, and annual minimum 30-day temperature.

> 所有温度序列均用终点对齐的 10 年 Theil--Sen slope；1981–1990 窗口记为 1990。annual、四季、最暖 30 天和最冷 30 天因此共享 1990–2020 轴。

Annual ice days use a trailing 10-year arithmetic mean on that same endpoint axis. Ice duration is treated as a state background, not converted into a warming-speed analogue.

> 年冰日使用同一终点轴的 10 年滑动均值。冰期是状态背景，不被强行转为“增温速度”。

In the current producer, an all-ice / no-valid-nonfreezing aggregation period is recorded as finite `0.0 °C`; it is not an `NA`. Consequently, an undefined Spearman coefficient can arise from a constant rolling-rate sequence, especially for persistently frozen seasons. Step 14 records this separately from the number of endpoint pairs.

> 当前 producer 将全冰或无有效非冻结日时段记为有限的 `0.0 °C`，不记为 `NA`。因此 Spearman 不可定义可能来自恒定的滚动 rate，尤其是持续冻结季节；Step 14 将其与 endpoint 对数分开记录。

## Diagnostic layers

### Seasonal dynamics

For each lake and each season, retain the full local-rate sequence and calculate its cross-endpoint Spearman alignment with annual local rate:

$$
\rho_s=\operatorname{Spearman}(r_{\mathrm{annual}}(t),r_s(t)).
$$

Report all four values rather than assigning a single dominant season. Maps describe continuous spatial variation in seasonal alignment and in endpoint-rate change. A summer-associated pattern requires JJA to be distinct relative to DJF, MAM, and SON, not merely negative.

> 每湖、每季保留完整局部 rate 序列，并计算其与 annual rate 的跨终点 Spearman 对齐。四季均报告，不强制指定单一主导季节。夏季关联必须相对其他季节具有特异性，不能只看 JJA 为负。

### Extreme-temperature dynamics

The same analysis is applied to annual maximum and minimum 30-day temperatures. Their alignments with annual rate distinguish whether changing peak warmth or changing cold-period warmth more closely co-varies with a lake's local annual-rate history.

> 对年最暖/最冷 30 天温度做同样分析。它们与 annual rate 的对齐可描述局部年速率历史更接近 peak warmth 还是 cold-period warmth 的变化。

The occurrence dates of those 30-day periods are retained for later circular-date analysis but are not part of the first-pass inference.

> 两个 30 天窗口的出现日期保留给后续环形日期分析，暂不进入首轮推断。

### Ice-duration dynamics

For each lake, calculate:

$$
\rho_{\mathrm{ice}}=
\operatorname{Spearman}(r_{\mathrm{annual}}(t),\bar I_{10}(t)).
$$

Also retain baseline ice duration and 1990--2020 ice-state change. Interpret positive, negative, and near-zero values as distinct response patterns, not as evidence for a universal ice mechanism.

> 每湖计算 annual rate 与 10 年冰日均值的 Spearman 对齐，并保留基线冰期和 1990–2020 冰态变化。正、负、近零都是不同响应模式，不代表统一冰机制。

## Statistical boundary

All 31 endpoints share overlapping observations. Cross-endpoint correlations are therefore descriptive alignment measures: no ordinary correlation p-values, causal language, or strict seasonal contribution fractions will be reported. Spatial aggregation requires a minimum lake count per cell and reports distributional summaries rather than area-weighted global fields.

> 31 个 endpoint 高度重叠。相关性仅描述对齐结构：不报告普通相关 p 值、不作因果表述、不计算严格季节贡献比例。空间图需设置每格最小湖泊数，报告分布汇总，不作面积加权全球场。

## Warming- and cooling-state branches

Overall alignment can conceal different behaviour when annual local rate is positive versus negative. Step 14 therefore repeats each seasonal and extreme-temperature comparison using only endpoints with positive annual rate, and only endpoints with negative annual rate. Each branch requires at least eight finite endpoint pairs and non-constant ranked values.

> overall 对齐可能掩盖 annual rate 为正与为负时的不同表现。Step 14 因此分别只用 annual rate 为正、只用 annual rate 为负的 endpoint 重算季节与极值对齐。每个分支至少需 8 个有限 endpoint 对，且秩序列不能恒定。

These sign-conditioned coefficients quantify relative co-variation within warming or cooling states. They do not identify which season contributes an additive share of annual warming, because rolling Theil--Sen rates across seasonal series are not decomposable in that way.

> 这些条件相关量化增温或降温状态内的相对协变，不能识别某季节对 annual warming 的可加贡献；不同季节的滚动 Theil--Sen rate 不可这样分解。

## Seasonal ice branch

Monthly ice-day counts are summed into DJF, MAM, JJA, and SON series using the same DJF year convention as temperature. For each season, retain a trailing-10-year ice-duration mean and an ice-loss rate, defined as the negative trailing-10-year Sen slope of ice days. Positive ice-loss rate means fewer seasonal ice days per year.

> 月冰日按与温度一致的 DJF 年份规则汇总为四季冰日。每季保留 10 年冰期均值与冰损失 rate；后者为冰日 10 年 Sen slope 取负，正值表示该季每年冰日减少。

Annual-rate alignment with seasonal ice loss is a targeted descriptive test of synchronous thermal-rate and ice-loss change. It is more appropriate than treating a persistent winter/spring `0.0 °C` frozen state as a continuously varying thermal signal. It remains neither a glacier-meltwater test nor causal evidence.

> annual rate 与季节冰损失的对齐用于描述同步热速率与失冰变化；相比把持续冬春 `0.0 °C` 冻结状态当连续温度信号更合适。但它仍不是冰川融水检验或因果证据。

## Outputs and decision gate

Step 14 produces lake-by-endpoint rate/state files and a per-lake alignment summary. The first outputs are:

1. multi-series trajectories for representative lakes and spatially aggregated dynamics;
2. maps of seasonal, maximum/minimum-temperature, and ice-state alignment;
3. continuous distributions of endpoint-rate and ice-state change.

> Step 14 输出逐湖逐终点 rate/state 文件和对齐摘要。首轮产物是代表湖多序列轨迹、空间汇总动态图、季节/极值/冰态对齐地图，以及 endpoint 变化的连续分布。

## First-pass spatial diagnostics

```{r}
#| label: setup-seasonal-ice-diagnostics
#| include: false
#| cache: false

source("../../../shared/R/figure-style.R")
source("../../../shared/R/seasonal-ice-diagnostics-helpers.R")
list2env(prepare_seasonal_ice_data(), envir = environment())

alignment_availability_grid <- prepare_alignment_availability_grids(min_lakes = 3)
alignment_grid <- prepare_alignment_grids(min_lakes = 3)
alignment_summary <- summarise_alignment()
conditional_summary <- summarise_conditional_alignment()
conditional_temperature_grid <- prepare_conditional_alignment_grids(
  "Temperature", c("JJA", "SON", "Max 30-day"), min_lakes = 3
)
conditional_ice_grid <- prepare_conditional_alignment_grids(
  "Seasonal ice loss", c("DJF", "MAM", "JJA", "SON"), min_lakes = 3
)
thermal_asymmetry_grid <- prepare_thermal_state_grid(
  c("JJA − DJF contrast", "Seasonal range", "Seasonal SD"), min_lakes = 3
)
thermal_asymmetry_summary <- summarise_thermal_state()
ice_alignment_pair_summary <- summarise_ice_alignment_pairs()
ice_alignment_pair_grid <- prepare_ice_alignment_pair_grid(min_lakes = 3)
direct_ice_phase_summary <- summarise_direct_ice_phase()
fmt_pct <- function(x) scales::percent(x, accuracy = .1)
fmt_rho <- function(x) formatC(x, format = "f", digits = 2)
```

Before interpreting a correlation map, its availability must be shown. Figure @fig-alignment-availability gives the fraction of lakes in each occupied 1° cell for which the annual-rate comparison is defined. A low fraction normally indicates a constant seasonal, extreme-temperature, or ice-state sequence; it does not indicate missing temperature observations.

> 解释相关地图前，需先显示其可定义性。@fig-alignment-availability 给出每个有湖的 1° 格网中 annual-rate 比较可定义的湖泊比例。低比例通常表示季节、极值或冰态序列恒定，并非温度观测缺失。

```{r}
#| label: fig-alignment-availability
#| fig-cap: "Fraction of lakes per 1° cell with a defined annual-rate alignment. Cells contain at least three lakes."
#| fig-width: 12
#| fig-height: 6
#| column: page
#| cache: false

ggplot(alignment_availability_grid) +
  geom_tile(aes(lon_cell + .5, lat_cell + .5, fill = defined_fraction), width = 1, height = 1) +
  the_geom_coastline() +
  facet_wrap(~series, ncol = 4) +
  scale_fill_paletteer_c(
    palette = "grDevices::YlGnBu",
    limits = c(0, 1),
    name = "Defined fraction"
  ) +
  scale_x_continuous(breaks = seq(-180, 180, 90), expand = c(0, 0)) +
  scale_y_continuous(breaks = seq(-60, 90, 30), expand = c(0, 0)) +
  coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 90), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") +
  theme_map_grid() +
  theme_facet() +
  theme(legend.position = "bottom")
```

Where at least three correlations are defined, Figure @fig-alignment-strength maps the within-cell median alignment. This is a descriptive spatial aggregation, not an area-weighted global field and not a causal seasonal contribution map.

> 对至少有 3 个可定义相关的格网，@fig-alignment-strength 显示其中位对齐。它是描述性空间汇总，不是面积加权全球场，也不是季节因果贡献图。

```{r}
#| label: fig-alignment-strength
#| fig-cap: "Median within-lake Spearman alignment between annual local warming rate and each rolling seasonal, extreme-temperature, or ice-state sequence, summarized within 1° cells."
#| fig-width: 12
#| fig-height: 6
#| column: page
#| cache: false

ggplot(alignment_grid) +
  geom_tile(aes(lon_cell + .5, lat_cell + .5, fill = median_alignment), width = 1, height = 1) +
  the_geom_coastline() +
  facet_wrap(~series, ncol = 4) +
  scale_fill_paletteer_c(
    palette = "grDevices::TealRose",
    limits = c(-1, 1),
    name = "Median alignment"
  ) +
  scale_x_continuous(breaks = seq(-180, 180, 90), expand = c(0, 0)) +
  scale_y_continuous(breaks = seq(-60, 90, 30), expand = c(0, 0)) +
  coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 90), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") +
  theme_map_grid() +
  theme_facet() +
  theme(legend.position = "bottom")
```

Figure @fig-alignment-distribution retains the lake-level distribution behind these medians. It distinguishes a globally weak comparison from a geographically mixed comparison with positive and negative modes that cancel in the median.

> @fig-alignment-distribution 保留这些中位数背后的湖泊尺度分布。它区分“全球普遍弱”与“正负模式并存、在中位数中抵消”的情况。

```{r}
#| label: fig-alignment-distribution
#| fig-cap: "Distribution of within-lake Spearman alignment between annual local rate and seasonal, extreme-temperature, ice-state, or thermal-asymmetry rate. Only defined correlations enter each panel."
#| fig-width: 12
#| fig-height: 7
#| column: page
#| cache: false

ggplot(filter(alignment_long, is.finite(spearman_alignment))) +
  geom_histogram(aes(spearman_alignment), binwidth = .05, boundary = 0, fill = col_inchart_fill, colour = col_inchart_geom_border, linewidth = wid_inchart_geom_border) +
  geom_vline(xintercept = 0, colour = col_axis, linewidth = wid_axis) +
  facet_wrap(~series, ncol = 5) +
  scale_x_continuous(limits = c(-1, 1), breaks = seq(-1, 1, .5)) +
  labs(x = "Within-lake Spearman alignment", y = "Lake count") +
  theme_base(grid = TRUE) +
  theme_facet()
```

At the lake level, JJA, SON, and maximum-30-day alignments are defined for `r fmt_pct(alignment_summary |> filter(series == "JJA") |> pull(defined_fraction))`, `r fmt_pct(alignment_summary |> filter(series == "SON") |> pull(defined_fraction))`, and `r fmt_pct(alignment_summary |> filter(series == "Max 30-day") |> pull(defined_fraction))` of lakes, respectively. Their median alignments are `r fmt_rho(alignment_summary |> filter(series == "JJA") |> pull(median_alignment))`, `r fmt_rho(alignment_summary |> filter(series == "SON") |> pull(median_alignment))`, and `r fmt_rho(alignment_summary |> filter(series == "Max 30-day") |> pull(median_alignment))`. Thus, across the broad set of lakes where a comparison is possible, local annual-rate dynamics most often co-vary with warm-season and warm-extreme dynamics.

> 湖泊尺度上，JJA、SON 与最暖 30 天对齐分别在 `r fmt_pct(alignment_summary |> filter(series == "JJA") |> pull(defined_fraction))`、`r fmt_pct(alignment_summary |> filter(series == "SON") |> pull(defined_fraction))`、`r fmt_pct(alignment_summary |> filter(series == "Max 30-day") |> pull(defined_fraction))` 湖泊中可定义；中位相关分别为 `r fmt_rho(alignment_summary |> filter(series == "JJA") |> pull(median_alignment))`、`r fmt_rho(alignment_summary |> filter(series == "SON") |> pull(median_alignment))`、`r fmt_rho(alignment_summary |> filter(series == "Max 30-day") |> pull(median_alignment))`。故在可比较湖泊中，annual rate 更常与暖季及暖极值动态协变。

DJF, MAM, and especially minimum-30-day comparisons are conditional branches, not globally comparable layers: their alignment is defined for only `r fmt_pct(alignment_summary |> filter(series == "DJF") |> pull(defined_fraction))`, `r fmt_pct(alignment_summary |> filter(series == "MAM") |> pull(defined_fraction))`, and `r fmt_pct(alignment_summary |> filter(series == "Min 30-day") |> pull(defined_fraction))` of lakes. Persistent `0.0 °C` frozen states make their local-rate sequences constant. These layers remain useful for the subset with temporal variation, but cannot be used to generalise about global cold-season control.

> DJF、MAM，尤其最冷 30 天是条件性分支，不是可全球比较的层：可定义比例仅为 `r fmt_pct(alignment_summary |> filter(series == "DJF") |> pull(defined_fraction))`、`r fmt_pct(alignment_summary |> filter(series == "MAM") |> pull(defined_fraction))`、`r fmt_pct(alignment_summary |> filter(series == "Min 30-day") |> pull(defined_fraction))`。持续 `0.0 °C` 冻结状态使局部 rate 恒定。它们仍可用于有时间变异的子集，但不能外推为全球冷季主导结论。

The ice-state alignment is defined for `r fmt_pct(alignment_summary |> filter(series == "Ice duration") |> pull(defined_fraction))` of lakes, but its median is `r fmt_rho(alignment_summary |> filter(series == "Ice duration") |> pull(median_alignment))`. This does not support a single global ice-duration response. The spatial field should instead be used to locate contrasting response settings for later, explicitly conditional investigation.

> 冰态对齐在 `r fmt_pct(alignment_summary |> filter(series == "Ice duration") |> pull(defined_fraction))` 湖泊中可定义，但中位数仅为 `r fmt_rho(alignment_summary |> filter(series == "Ice duration") |> pull(median_alignment))`。这不支持单一的全球冰期响应；空间图应只用于定位对比性响应环境，供后续有条件地深入调查。

## Sign-conditioned temperature alignment

Among endpoints with positive annual local rate, median alignment is `r fmt_rho(conditional_summary |> filter(annual_state == "Positive annual rate", family == "Temperature", series == "JJA") |> pull(median_alignment))` for JJA, `r fmt_rho(conditional_summary |> filter(annual_state == "Positive annual rate", family == "Temperature", series == "SON") |> pull(median_alignment))` for SON, and `r fmt_rho(conditional_summary |> filter(annual_state == "Positive annual rate", family == "Temperature", series == "Max 30-day") |> pull(median_alignment))` for maximum 30-day temperature. For negative annual-rate endpoints, corresponding values are `r fmt_rho(conditional_summary |> filter(annual_state == "Negative annual rate", family == "Temperature", series == "JJA") |> pull(median_alignment))`, `r fmt_rho(conditional_summary |> filter(annual_state == "Negative annual rate", family == "Temperature", series == "SON") |> pull(median_alignment))`, and `r fmt_rho(conditional_summary |> filter(annual_state == "Negative annual rate", family == "Temperature", series == "Max 30-day") |> pull(median_alignment))`.

> 在 annual local rate 为正的 endpoint 中，JJA、SON、最暖 30 天的中位对齐分别为 `r fmt_rho(conditional_summary |> filter(annual_state == "Positive annual rate", family == "Temperature", series == "JJA") |> pull(median_alignment))`、`r fmt_rho(conditional_summary |> filter(annual_state == "Positive annual rate", family == "Temperature", series == "SON") |> pull(median_alignment))`、`r fmt_rho(conditional_summary |> filter(annual_state == "Positive annual rate", family == "Temperature", series == "Max 30-day") |> pull(median_alignment))`；annual rate 为负时分别为 `r fmt_rho(conditional_summary |> filter(annual_state == "Negative annual rate", family == "Temperature", series == "JJA") |> pull(median_alignment))`、`r fmt_rho(conditional_summary |> filter(annual_state == "Negative annual rate", family == "Temperature", series == "SON") |> pull(median_alignment))`、`r fmt_rho(conditional_summary |> filter(annual_state == "Negative annual rate", family == "Temperature", series == "Max 30-day") |> pull(median_alignment))`。

Figure @fig-sign-conditioned-temperature maps these two conditional profiles. Tile opacity is the fraction of lakes in the cell with a defined conditional correlation, preventing sparse branches from looking equally certain.

> @fig-sign-conditioned-temperature 显示两种条件 profile 的空间分布。格网透明度表示其中条件相关可定义的湖泊比例，避免稀疏分支看似同样可靠。

```{r}
#| label: fig-sign-conditioned-temperature
#| fig-cap: "Median conditional Spearman alignment between annual local rate and warm-season or warm-extreme rate. Rows condition on the sign of annual local rate; opacity shows fraction of lakes with a defined conditional alignment."
#| fig-width: 12
#| fig-height: 5
#| column: page
#| cache: false

ggplot(conditional_temperature_grid) +
  geom_tile(
    aes(lon_cell + .5, lat_cell + .5, fill = median_alignment, alpha = defined_fraction),
    width = 1, height = 1
  ) +
  the_geom_coastline() +
  facet_grid(annual_state ~ series) +
  scale_fill_paletteer_c(
    palette = "grDevices::TealRose", limits = c(-1, 1), name = "Median alignment"
  ) +
  scale_alpha_continuous(range = c(.25, 1), limits = c(0, 1), name = "Defined fraction") +
  scale_x_continuous(breaks = seq(-180, 180, 90), expand = c(0, 0)) +
  scale_y_continuous(breaks = seq(-60, 90, 30), expand = c(0, 0)) +
  coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 90), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") +
  theme_map_grid() +
  theme_facet() +
  theme(legend.position = "bottom")
```

## Seasonal ice-loss alignment

The seasonal ice-loss branch replaces frozen-season temperature rate with a directly interpretable ice-change sequence. For positive annual-rate endpoints, JJA ice-loss alignment has median `r fmt_rho(conditional_summary |> filter(annual_state == "Positive annual rate", family == "Seasonal ice loss", series == "JJA") |> pull(median_alignment))`; for negative annual-rate endpoints it is `r fmt_rho(conditional_summary |> filter(annual_state == "Negative annual rate", family == "Seasonal ice loss", series == "JJA") |> pull(median_alignment))`. Figure @fig-sign-conditioned-ice shows whether this and the winter/spring branches are geographically coherent.

> 季节冰损失分支以可解释的冰变化序列替代冻结季节 temperature rate。annual rate 为正时，JJA 冰损失对齐中位数为 `r fmt_rho(conditional_summary |> filter(annual_state == "Positive annual rate", family == "Seasonal ice loss", series == "JJA") |> pull(median_alignment))`；annual rate 为负时为 `r fmt_rho(conditional_summary |> filter(annual_state == "Negative annual rate", family == "Seasonal ice loss", series == "JJA") |> pull(median_alignment))`。@fig-sign-conditioned-ice 用于查看这一模式及冬春分支是否具有空间连贯性。

```{r}
#| label: fig-sign-conditioned-ice
#| fig-cap: "Median conditional Spearman alignment between annual local rate and seasonal ice-loss rate. Positive ice loss means fewer ice days yr⁻¹; opacity shows fraction of lakes with a defined conditional alignment."
#| fig-width: 12
#| fig-height: 5
#| column: page
#| cache: false

ggplot(conditional_ice_grid) +
  geom_tile(
    aes(lon_cell + .5, lat_cell + .5, fill = median_alignment, alpha = defined_fraction),
    width = 1, height = 1
  ) +
  the_geom_coastline() +
  facet_grid(annual_state ~ series) +
  scale_fill_paletteer_c(
    palette = "grDevices::TealRose", limits = c(-1, 1), name = "Median alignment"
  ) +
  scale_alpha_continuous(range = c(.25, 1), limits = c(0, 1), name = "Defined fraction") +
  scale_x_continuous(breaks = seq(-180, 180, 90), expand = c(0, 0)) +
  scale_y_continuous(breaks = seq(-60, 90, 30), expand = c(0, 0)) +
  coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 90), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") +
  theme_map_grid() +
  theme_facet() +
  theme(legend.position = "bottom")
```

## Seasonal thermal asymmetry

Correlation maps ask whether two local-rate sequences co-vary. They do not show whether a particular 10-year annual change is thermally concentrated in the warm or cold part of the year. Step 14 therefore adds three annual thermal-state diagnostics before calculating their endpoint-aligned 10-year Sen rates: fixed warm--cold contrast (`JJA − DJF`), four-season range, and four-season standard deviation. A positive contrast-rate means JJA warms relative to DJF; a negative value means the observable JJA--DJF thermal contrast contracts. Range and standard-deviation rates test whether the conclusion depends on fixing JJA and DJF as the warm/cold pair.

> 相关图回答两个 rate 是否协变，不能显示某个 10 年年均变化在暖季还是冷季更集中。故 Step 14 新增 JJA−DJF 热对比、四季极差和四季标准差，并对其计算终点对齐的 10 年 Sen rate。对比 rate 为正表示 JJA 相对 DJF 增温更快；为负表示可观测的 JJA−DJF 热差收敛。极差与标准差用于检验结论是否依赖固定 JJA/DJF。

These measures are seasonal asymmetry diagnostics, not additive seasonal contributions. For ice-covered lakes, winter temperatures can remain fixed at `0 °C`; a contracting contrast then describes the observed LSWT seasonal cycle, while seasonal ice-day changes retain the separate frozen-state information.

> 这些量是季节不均衡诊断，不是可加的季节贡献。冰湖冬季温度可固定在 `0 °C`；此时热差收敛描述的是可观测 LSWT 年内循环，冻结状态仍需由季节冰日单独表达。

For endpoints with positive annual rate, a positive thermal-asymmetry rate identifies warmer-season-relative amplification; a negative rate identifies cold-season-relative amplification. For endpoints with negative annual rate, signs reverse their interpretation: positive asymmetry means cold-season-relative cooling, while negative asymmetry means warm-season-relative cooling. The terms *relative* and *observed* are essential: both seasons can warm or cool in any quadrant.

> 对 annual rate 为正的 endpoint，热不均衡 rate 为正表示暖季相对强化，为负表示冷季相对强化；对 annual rate 为负的 endpoint，正值表示冷季相对降温更快，负值表示暖季相对降温更快。必须保留“相对”和“可观测”：每个象限中各季均可能同向变化。

Figure @fig-thermal-asymmetry-space maps the within-lake median thermal-asymmetry rate among positive and negative annual-rate endpoints. Tile fill is the median across lakes in each 1° cell, and opacity records the within-lake median fraction of endpoints with positive asymmetry. This separates spatial heterogeneity from a global median that can mix opposite response modes.

> @fig-thermal-asymmetry-space 显示正／负 annual-rate endpoint 内，每湖热不均衡 rate 的中位数，再按 1° 格网汇总。填色为格网内中位 rate，透明度为各湖“热不均衡为正” endpoint 比例的中位数。这样可避免全球中位数混合相反响应模式。

```{r}
#| label: fig-thermal-asymmetry-space
#| fig-cap: "Seasonal thermal-asymmetry rates conditional on annual local-rate sign. Fill: within-cell median of each lake's median endpoint rate; opacity: within-cell median fraction of positive diagnostic rates. Positive JJA − DJF change means warm-season-relative amplification."
#| fig-width: 12
#| fig-height: 6
#| column: page
#| cache: false

ggplot(thermal_asymmetry_grid) +
  geom_tile(
    aes(lon_cell + .5, lat_cell + .5, fill = median_rate, alpha = positive_rate_fraction),
    width = 1, height = 1
  ) +
  the_geom_coastline() +
  facet_grid(annual_state ~ diagnostic) +
  scale_fill_paletteer_c(
    palette = "grDevices::TealRose", name = "Median diagnostic rate (°C yr⁻¹)"
  ) +
  scale_alpha_continuous(range = c(.25, 1), limits = c(0, 1), name = "Positive-rate fraction") +
  scale_x_continuous(breaks = seq(-180, 180, 90), expand = c(0, 0)) +
  scale_y_continuous(breaks = seq(-60, 90, 30), expand = c(0, 0)) +
  coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 90), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") +
  theme_map_grid() +
  theme_facet() +
  theme(legend.position = "bottom")
```

At lake level, the positive-annual-rate branch has median JJA--DJF contrast rate `r fmt_rho(thermal_asymmetry_summary |> filter(annual_state == "Positive annual rate", diagnostic == "JJA − DJF contrast") |> pull(median_endpoint_rate))` °C yr⁻¹; the negative-annual-rate branch has `r fmt_rho(thermal_asymmetry_summary |> filter(annual_state == "Negative annual rate", diagnostic == "JJA − DJF contrast") |> pull(median_endpoint_rate))` °C yr⁻¹. These global summaries are only reference values; interpretation should follow the spatially organised sign mixtures in @fig-thermal-asymmetry-space.

> 湖泊尺度上，正 annual-rate 分支的 JJA−DJF 热对比 rate 中位数为 `r fmt_rho(thermal_asymmetry_summary |> filter(annual_state == "Positive annual rate", diagnostic == "JJA − DJF contrast") |> pull(median_endpoint_rate))` °C yr⁻¹；负分支为 `r fmt_rho(thermal_asymmetry_summary |> filter(annual_state == "Negative annual rate", diagnostic == "JJA − DJF contrast") |> pull(median_endpoint_rate))` °C yr⁻¹。这些全球汇总只作参照；解释应回到 @fig-thermal-asymmetry-space 中空间组织化的正负混合。

## Paired seasonal ice alignments

Different seasonal ice-loss medians can arise either because the same lakes have opposite seasonal alignments or because each seasonal comparison is defined for a different lake population. These alternatives are tested directly in the overlap set: pair JJA with SON, and DJF with MAM, only where both within-lake annual-rate alignments are defined.

> 不同季节冰损失的中位对齐差异，可能来自同一湖在不同季节相反，也可能只是不同季节的可定义湖群不同。故直接在重叠样本检验 JJA–SON 与 DJF–MAM：仅保留两季 annual-rate 对齐都可定义的湖泊。

The JJA--SON pair contains `r scales::comma(ice_alignment_pair_summary |> filter(pair == "JJA vs SON") |> pull(n_lakes))` lakes; its between-season Spearman association is `r fmt_rho(ice_alignment_pair_summary |> filter(pair == "JJA vs SON") |> pull(spearman_between_seasons))`, and `r fmt_pct(ice_alignment_pair_summary |> filter(pair == "JJA vs SON") |> pull(opposite_sign_fraction))` have opposite signs. The DJF--MAM overlap contains `r scales::comma(ice_alignment_pair_summary |> filter(pair == "DJF vs MAM") |> pull(n_lakes))` lakes, with corresponding values `r fmt_rho(ice_alignment_pair_summary |> filter(pair == "DJF vs MAM") |> pull(spearman_between_seasons))` and `r fmt_pct(ice_alignment_pair_summary |> filter(pair == "DJF vs MAM") |> pull(opposite_sign_fraction))`.

> JJA–SON 重叠样本有 `r scales::comma(ice_alignment_pair_summary |> filter(pair == "JJA vs SON") |> pull(n_lakes))` 湖；两季间 Spearman 为 `r fmt_rho(ice_alignment_pair_summary |> filter(pair == "JJA vs SON") |> pull(spearman_between_seasons))`，符号相反比例为 `r fmt_pct(ice_alignment_pair_summary |> filter(pair == "JJA vs SON") |> pull(opposite_sign_fraction))`。DJF–MAM 重叠样本有 `r scales::comma(ice_alignment_pair_summary |> filter(pair == "DJF vs MAM") |> pull(n_lakes))` 湖，对应为 `r fmt_rho(ice_alignment_pair_summary |> filter(pair == "DJF vs MAM") |> pull(spearman_between_seasons))` 和 `r fmt_pct(ice_alignment_pair_summary |> filter(pair == "DJF vs MAM") |> pull(opposite_sign_fraction))`。

Figure @fig-paired-ice-alignment-space maps the modal sign combination within each occupied 1° cell. Opacity is the share of lakes belonging to that modal combination. It distinguishes genuinely mixed local seasonal responses from broad cells dominated by one paired response pattern.

> @fig-paired-ice-alignment-space 显示每个 1° 格网中最常见的季节冰对齐符号组合；透明度为该组合在格网中的湖泊比例。它区分同地混合的季节响应与被单一组合主导的格网。

```{r}
#| label: fig-paired-ice-alignment-space
#| fig-cap: "Modal paired sign combination of annual-rate alignment with seasonal ice-loss rate within 1° cells. Only lakes with both seasonal alignments defined enter each pair; opacity is modal-class fraction."
#| fig-width: 12
#| fig-height: 5
#| column: page
#| cache: false

ggplot(ice_alignment_pair_grid) +
  geom_tile(
    aes(lon_cell + .5, lat_cell + .5, fill = sign_class, alpha = class_fraction),
    width = 1, height = 1
  ) +
  the_geom_coastline() +
  facet_wrap(~pair) +
  scale_fill_manual(
    values = ice_pair_palette,
    name = "Modal sign combination"
  ) +
  scale_alpha_continuous(range = c(.25, 1), limits = c(0, 1), name = "Modal-class fraction") +
  scale_x_continuous(breaks = seq(-180, 180, 90), expand = c(0, 0)) +
  scale_y_continuous(breaks = seq(-60, 90, 30), expand = c(0, 0)) +
  coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 90), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") +
  theme_map_grid() +
  theme_facet() +
  theme(legend.position = "bottom")
```

For JJA--SON, opposite signs are not a minor tail: the modal cross-season pattern is positive JJA alignment and negative SON alignment. Thus the positive global JJA median and slightly negative SON median do not describe one uniform annual-ice response; they combine spatially organised seasonal phase differences. DJF--MAM is more often jointly negative, but its non-zero opposite-sign branch likewise prevents a universal winter/spring ice interpretation.

> 对 JJA–SON，异号不是小尾部：最常见模式是 JJA 对齐为正、SON 对齐为负。因此 JJA 的全球正中位数与 SON 的略负中位数并非单一统一冰响应，而是空间组织化的季节相位差。DJF–MAM 更常同时为负，但异号分支也排除了统一的冬春冰期解释。

## Direct JJA--SON ice-loss configurations

The preceding paired analysis compares *correlations with annual temperature rate*; it cannot determine whether JJA and SON ice days themselves change in opposite directions. This direct diagnostic instead classifies every lake-endpoint pair by the signs of its JJA and SON ice-loss rates, then reports the direct annual ice-loss rate in the same endpoint. Positive ice loss means fewer ice days per year; negative ice loss means more ice days per year. JJA and SON are calendar seasons: their warm/cold interpretation is Northern-Hemisphere-specific and must be stratified by hemisphere before phenological inference.

> 前述配对分析比较的是“与 annual temperature rate 的相关”，不能判断 JJA 与 SON 冰日本身是否反向变化。本节直接按每个湖—endpoint 的 JJA、SON 冰损失 rate 符号分类，并报告同 endpoint 的全年冰损失 rate。正值表示每年冰日减少；负值表示每年冰日增加。JJA/SON 是日历季节；在南半球其暖冷含义相反，任何物候解释必须先分半球。

The `JJA and SON loss` configuration has median annual ice loss `r fmt_rho(direct_ice_phase_summary |> filter(direct_phase == "JJA and SON loss") |> pull(median_annual_ice_loss_rate))` days yr⁻¹ (IQR `r fmt_rho(direct_ice_phase_summary |> filter(direct_phase == "JJA and SON loss") |> pull(q25_annual_ice_loss_rate))` to `r fmt_rho(direct_ice_phase_summary |> filter(direct_phase == "JJA and SON loss") |> pull(q75_annual_ice_loss_rate))`). The opposite direct configurations are near annual balance: `JJA loss / SON gain` has median `r fmt_rho(direct_ice_phase_summary |> filter(direct_phase == "JJA loss / SON gain") |> pull(median_annual_ice_loss_rate))`, and `JJA gain / SON loss` has median `r fmt_rho(direct_ice_phase_summary |> filter(direct_phase == "JJA gain / SON loss") |> pull(median_annual_ice_loss_rate))` days yr⁻¹.

> `JJA and SON loss` 的全年冰损失中位数为 `r fmt_rho(direct_ice_phase_summary |> filter(direct_phase == "JJA and SON loss") |> pull(median_annual_ice_loss_rate))` 天 yr⁻¹（IQR：`r fmt_rho(direct_ice_phase_summary |> filter(direct_phase == "JJA and SON loss") |> pull(q25_annual_ice_loss_rate))` 至 `r fmt_rho(direct_ice_phase_summary |> filter(direct_phase == "JJA and SON loss") |> pull(q75_annual_ice_loss_rate))`）。两种直接反向组合接近全年平衡：`JJA loss / SON gain` 中位数为 `r fmt_rho(direct_ice_phase_summary |> filter(direct_phase == "JJA loss / SON gain") |> pull(median_annual_ice_loss_rate))`，`JJA gain / SON loss` 为 `r fmt_rho(direct_ice_phase_summary |> filter(direct_phase == "JJA gain / SON loss") |> pull(median_annual_ice_loss_rate))` 天 yr⁻¹。

Figure @fig-direct-ice-phase tests the proposed distinction without imposing an arbitrary annual-loss threshold. Joint JJA--SON loss is consistent with overall shortening; opposite direct signs are consistent with calendar-season redistribution. They do not by themselves prove a delayed freeze-up or earlier break-up, which requires monthly or daily transition-date analysis within hemisphere.

> @fig-direct-ice-phase 不设任意全年失冰阈值，直接检验上述区分。JJA–SON 同时失冰与全年缩短一致；两季直接异号与日历季节再分配一致。但它们本身不能证明冻结后移或融化提前，后者需要分半球的逐月或逐日转折日期分析。

```{r}
#| label: fig-direct-ice-phase
#| fig-cap: "Direct annual ice-loss rate conditional on the signs of JJA and SON ice-loss rates. Points are medians; intervals are interquartile ranges across lake-endpoint observations. Positive means fewer annual ice days yr⁻¹."
#| fig-width: 8
#| fig-height: 4

phase_order <- c(
  "JJA and SON gain", "JJA gain / SON loss", "At least one stable season",
  "JJA loss / SON gain", "JJA and SON loss"
)
ggplot(
  mutate(direct_ice_phase_summary, direct_phase = factor(direct_phase, levels = phase_order)),
  aes(direct_phase, median_annual_ice_loss_rate)
) +
  geom_hline(yintercept = 0, colour = col_axis, linewidth = wid_axis) +
  geom_errorbar(aes(ymin = q25_annual_ice_loss_rate, ymax = q75_annual_ice_loss_rate), width = .15) +
  geom_point(size = 2) +
  coord_flip() +
  labs(x = NULL, y = "Annual ice-loss rate (days yr⁻¹)") +
  theme_base(grid = TRUE)
```
