---
title: "Methods"
aliases:
  - "/explorations/warming-acceleration/manuscript/02-methods.html"
---

```{r}
#| include: false
source("../../../shared/R/figure-style.R")
source("../../../shared/R/descriptive-helpers.R")
source("../../../shared/R/methods-data-helpers.R")
list2env(prepare_methods_data(), envir = environment())
```

## Data and scope

We analyse the 1981--2020 historical GLAST reconstruction for 92,245 lakes.
The product combines satellite information with ERA5-Land-forced FLAKE model
output and calibration; all temperature inferences concern reconstructed lake
surface conditions [@tong2023]. Frozen `0 °C` states are retained as physical
ice states rather than treated as ordinary liquid-water temperatures.

> 分析 92,245 个湖泊 1981--2020 年 GLAST 历史重建。产品整合卫星信息、ERA5-Land 强迫的 FLAKE 输出和校正；所有温度推断都针对重建湖表状态。冻结 `0 °C` 作为物理冰状态保留，不视作普通液态水温。

Records cover all inhabited continents but are spatially uneven, with dense
coverage in northern North America and Europe. Lake-equal summaries therefore
describe the reconstructed lake population; equal-area summaries give each
occupied spatial cell one vote where sampling density would otherwise alter an
aggregate.

> 记录覆盖各有人居大陆，但空间密度不均，北美北部和欧洲最密集。湖泊等权汇总描述重建湖泊总体；当采样密度会改变聚合结果时，等面积汇总让每个被占据空间格网一票。

```{r}
#| label: fig-data-coverage
#| fig-cap: "Spatial coverage and lake-count density of the 92,245 reconstructed lake records. Insets show marginal lake counts by longitude and latitude, and counts by continent."
#| fig-width: 8
#| fig-height: 4

lake_density_figure <- ggplot() +
  geom_tile(
    data = lake_density_data$bins, aes(lon, lat, fill = n),
    width = 1, height = 1
  ) +
  the_geom_coastline() +
  geom_xsideline(
    data = lake_density_data$lon_counts, aes(lon_band, n),
    colour = col_inchart_line, linewidth = wid_inchart_line
  ) +
  geom_ysideline(
    data = lake_density_data$lat_counts, aes(n, lat_band),
    colour = col_inchart_line, linewidth = wid_inchart_line
  ) +
  scale_fill_paletteer_c(
    palette = "grDevices::TealRose", trans = "log10", name = "Lakes"
  ) +
  scale_x_continuous(breaks = seq(-180, 180, 60), expand = c(0, 0)) +
  scale_y_continuous(breaks = seq(-60, 90, 30), expand = c(0, 0)) +
  coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 90), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") +
  theme_map_grid() + theme_facet()

draw_fixed_aspect(
  ggdraw(lake_density_figure) +
    draw_plot(continent_bar, x = .08, y = .12, width = .16, height = .25),
  aspect = 6 / 3
)
```

## Primary trajectory metrics

Long-term warming is the Theil--Sen slope of raw annual-mean LSWT over the full
record, expressed as °C decade⁻¹. Local warming
rate is the trailing, endpoint-aligned 10-year Theil--Sen slope of raw annual
LSWT. The Sen trend of the valid local-rate sequence is termed warming-speed
change. It is an operational description of trajectory evolution, not resolved
instantaneous physical acceleration.

> 长期增温为 raw 年均 LSWT 全期 Theil--Sen slope，并报告为 °C decade⁻¹。局地增温速度为端点对齐的 10 年滑动 Theil--Sen slope。有效局地速度序列的 Sen trend 称为增温速度变化；它描述轨迹演变，不是瞬时物理加速度。

## Seasonal and ice-state diagnostics

Endpoint-aligned 10-year Sen rates are also calculated for seasonal,
extreme-temperature, and ice-day series. These diagnostics compare directional
co-variation with annual local rates. They are not additive decompositions of
annual warming and do not test glacier-meltwater mechanisms.

> 对季节、极端温度和冰日序列也计算端点对齐的 10 年 Sen rate。它们比较与年局地速度的方向共变，不可相加为年增温贡献，也不检验冰川融水机制。

## Spatially balanced trajectory decomposition

Monthly STL with `nt=99` is used only to produce the low-frequency
representation for PCA. Annual STL trajectories are centred on their
1981--1990 mean, averaged in occupied equal-area cells, and decomposed by PCA.
Each lake then receives projected scores on the fixed cell-PCA axes. PCA is not
refitted at lake level.

> `nt=99` 月尺度 STL 只用于得到 PCA 的低频表征。年 STL 轨迹按 1981--1990 均值中心化，在被占据等面积格网中汇总后 PCA；每湖再投影至固定格网 PCA 轴，不在湖泊层级重拟 PCA。

## Stability and external association

We assess PCA structure across grid resolutions and leave-one-continent-out
(LOCO) refits. Spatial association fields are evaluated with spatial hold-outs,
LOCO, and leave-one-decade-out checks. For the retained circulation result, raw
JJA LSWT is linearly detrended for each lake after frozen states are excluded,
then correlated with the predeclared prior-summer index. These checks assess
reproducibility of association, not causal attribution.

> PCA 在不同格网和 LOCO 重拟合下检验。空间关联场以空间留出、LOCO 与 leave-one-decade-out 检查。保留的环流结果中，每湖 raw JJA LSWT 在排除冻结状态后线性去趋势，再与预先定义的上一年夏季指数相关。检验关联可重复性，不做因果归因。
