---
title: Descriptive
---

```{r}
#| label: setup-figures
#| include: false
#| cache: false

source("../../../shared/R/figure-style.R")
source("../../../shared/R/descriptive-helpers.R")
```

## Sample coverage

```{r}
#| cache: true

# Setup for fig-lake-density-map
raw_monthly <- read_csv(
  file.path(data, "01-monthly-temperature", "output", "monthly_temperature.csv")
)
lake_meta_data <- read_csv(
  file.path(data, "00-lake-metadata", "output", "lake_metadata.csv")
)
lake_warming_metrics <- read_csv(
  file.path(
    data,
    "06-lake-warming-metrics",
    "output",
    "period12_robustfalse_ni5_no0_nt199",
    "lake_warming_metrics.csv"
  ),
  col_select = c(
    lake_id,
    raw_annual_mean_temp_mean,
    raw_annual_mean_temp_sen_slope_40yr,
    stl_annual_trend_sen_slope_40yr,
    stl_annual_trend_diff_sen_slope_1e3
  )
)

lakes_num <- nrow(raw_monthly)
start_year <- as.integer(sub(".*?(\\d{4}).*", "\\1", names(raw_monthly)[4]))
end_year <- as.integer(sub(".*?(\\d{4}).*", "\\1", tail(names(raw_monthly), 1)))
```

GLAST: `r lakes_num` lakes, `r start_year`--`r end_year`.

## Lake density map

```{r}
#| label: fig-lake-density-map
#| fig-cap: "Lake density map with 1° × 1° grid cells."
#| fig-width: 8
#| fig-height: 4
#| column: body-outset
#| cache: true

lake_density_data <- bin_lake_locations(raw_monthly |> select(lon, lat))
continent_bar <- make_continent_inchart(make_continent_summary(lake_meta_data))

# Main map with ggside (no legend)
lake_density_figure <- ggplot() +
  geom_tile(
    data = lake_density_data$bins,
    aes(x = lon, y = lat, fill = n),
    width = 1, height = 1, color = NA
  ) +
  the_geom_coastline() +
  geom_xsideline(
    data = lake_density_data$lon_counts,
    aes(x = lon_band, y = n),
    inherit.aes = FALSE,
    color = col_inchart_line,
    linewidth = wid_inchart_line
  ) +
  geom_ysideline(
    data = lake_density_data$lat_counts,
    aes(x = n, y = lat_band),
    inherit.aes = FALSE,
    color = col_inchart_line,
    linewidth = wid_inchart_line
  ) +
  scale_fill_paletteer_c(
    palette = "grDevices::TealRose",
    direction = 1,
    trans = "log10",
    name = "Lakes count",
    guide = legendry::guide_colbar(
      key = "log",
      second_guide = "none",
    )
  ) +
  scale_x_continuous(breaks = seq(-180, 180, by = 60), expand = c(0, 0)) +
  scale_y_continuous(breaks = seq(-60, 90, by = 30), expand = c(0, 0)) +
  scale_xsidey_continuous(
    breaks = c(600, 1200),
    limits = c(0, 1200),
    expand = c(0, 0)
  ) +
  scale_ysidex_continuous(
    breaks = c(4000, 8000),
    limits = c(0, 8000),
    labels = scales::label_number(scale_cut = scales::cut_short_scale()),
    expand = c(0, 0)
  ) +
  coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 90), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") +
  theme_map_grid() +
  theme_facet() + 
  theme(
    legend.position = c(0.54, 0),
    ggside.axis.ticks = element_line(),
    ggside.panel.scale.x = 1.5 / 5,
    ggside.panel.scale.y = 1.5 / 12, 
  )

lake_density_composite <- ggdraw(lake_density_figure) +
  draw_plot(continent_bar, x = 0.10, y = 0.16, width = 0.16, height = 0.25) 

draw_fixed_aspect(lake_density_composite, aspect = 6 / 3)
```

## Long-term warming && acceleration

```{r}
#| include: false
#| cache: true

# Setup for long-term warming summaries
fmt_count <- function(x) scales::comma(x, accuracy = 1)
fmt_num <- function(x, digits = 2) formatC(x, format = "f", digits = digits)
fmt_pct <- function(x, digits = 1) fmt_num(100 * x, digits)

summarise_warming <- function(x) {
  x <- x[is.finite(x)]
  n_finite <- length(x)
  n_warming <- sum(x > 0)
  n_cooling <- sum(x <= 0)

  tibble(
    n_finite = n_finite,
    n_warming = n_warming,
    n_cooling = n_cooling,
    prop_warming = n_warming / n_finite,
    mean = mean(x),
    sd = sd(x),
    median = median(x),
    q25 = quantile(x, 0.25),
    q75 = quantile(x, 0.75)
  ) 
}

raw_warming_summary <- summarise_warming(
  lake_warming_metrics$raw_annual_mean_temp_sen_slope_40yr
)
stl_warming_summary <- summarise_warming(
  lake_warming_metrics$stl_annual_trend_sen_slope_40yr
)
```

```{r}
#| include: false
#| cache: true

# Setup for long-term acceleration summaries
summarise_acceleration <- function(x) {
  x <- x[is.finite(x)]
  n_finite <- length(x)
  n_positive <- sum(x > 0)
  n_negative <- sum(x <= 0)

  tibble(
    n_finite = n_finite,
    n_positive = n_positive,
    n_negative = n_negative,
    prop_positive = n_positive / n_finite,
    mean = mean(x),
    sd = sd(x),
    median = median(x),
    q25 = quantile(x, 0.25),
    q75 = quantile(x, 0.75)
  )
}

accel_summary <- summarise_acceleration(
  lake_warming_metrics$stl_annual_trend_diff_sen_slope_1e3
)

joint_state_summary <- lake_warming_metrics |>
  transmute(
    warming_speed = raw_annual_mean_temp_sen_slope_40yr,
    acceleration = stl_annual_trend_diff_sen_slope_1e3,
    state = case_when(
      warming_speed > 0 & acceleration > 0 ~ "warming + accelerating",
      warming_speed > 0 & acceleration <= 0 ~ "warming + decelerating",
      warming_speed <= 0 & acceleration > 0 ~ "cooling + accelerating",
      warming_speed <= 0 & acceleration <= 0 ~ "cooling + decelerating",
      TRUE ~ NA_character_
    )
  ) |>
  filter(is.finite(warming_speed), is.finite(acceleration), !is.na(state)) |>
  count(state, name = "n") |>
  mutate(
    prop = n / sum(n),
    state = factor(
      state,
      levels = c(
        "warming + accelerating",
        "warming + decelerating",
        "cooling + accelerating",
        "cooling + decelerating"
      )
    )
  ) |>
  arrange(state)

state_count <- function(label) {
  joint_state_summary |>
    filter(state == label) |>
    pull(n)
}

state_prop <- function(label) {
  joint_state_summary |>
    filter(state == label) |>
    pull(prop)
}
```

@tbl-warming-summary-raw and @tbl-warming-summary-stl-trend
shows the statistics of warming.
Compared with the raw annual mean temperature series,
the STL trend component is smoother and more stable,
so it identifies warming in more lakes
(91,802 / 92,245, 99.5%, versus 85,569 / 92,245, 92.8% for the raw series).
It also has a smaller standard deviation and a more concentrated quartile range,
although its mean is lower (0.573 versus 0.881 for the raw series).

::: {.column-margin}
Use Sen's slope to determine whether a lake is warming
and how is the trend like.
:::

Though almost all the lakes are warming,
the warming is not uniform.
Acceleration of `r fmt_count(accel_summary$n_positive)`
(`r fmt_pct(accel_summary$prop_positive)`%) lakes is positive.
Standard deviation and quartile range are also large, indicating that the acceleration varies widely across lakes.

::: {.column-margin} 
Here **acceleration** is measured as Sen's slope of the annual first-differenced STL trend. Positive values mean faster warming, slower cooling, or a transition toward warming.
:::

::: {.column-page-inset}
::: {#tbl-warming-summary layout-ncol=3}
::: {.list-table header-rows=0}

- - Warming count
  - `r fmt_count(raw_warming_summary$n_warming)`
    (`r fmt_pct(raw_warming_summary$prop_warming)`%)

- - Mean (°C/40yr)
  - `r fmt_num(raw_warming_summary$mean, 3)` ±
    `r fmt_num(raw_warming_summary$sd, 3)`

- - Quartile (°C/40yr)
  - `r fmt_num(raw_warming_summary$q25, 2)`,
    `r fmt_num(raw_warming_summary$median, 2)`,
    `r fmt_num(raw_warming_summary$q75, 2)`

Raw warming {#tbl-warming-summary-raw}
:::

::: {.list-table header-rows=0}

- - Warming count
  - `r fmt_count(stl_warming_summary$n_warming)`
    (`r fmt_pct(stl_warming_summary$prop_warming)`%)

- - Mean (°C/40yr)
  - `r fmt_num(stl_warming_summary$mean, 3)` ±
    `r fmt_num(stl_warming_summary$sd, 3)`

- - Quartile (°C/40yr)
  - `r fmt_num(stl_warming_summary$q25, 2)`,
    `r fmt_num(stl_warming_summary$median, 2)`,
    `r fmt_num(stl_warming_summary$q75, 2)`

STL trend warming {#tbl-warming-summary-stl-trend}
:::

::: {.list-table header-rows=0}

- - Accelerating count
  - `r fmt_count(accel_summary$n_positive)`
    (`r fmt_pct(accel_summary$prop_positive)`%)

- - Mean (°C/40yr)
  - `r fmt_num(accel_summary$mean, 3)` ±
    `r fmt_num(accel_summary$sd, 3)`

- - Quartile (°C/40yr)
  - `r fmt_num(accel_summary$q25, 2)`,
    `r fmt_num(accel_summary$median, 2)`,
    `r fmt_num(accel_summary$q75, 2)`

STL trend acceleration {#tbl-accel-summary}
:::

Summary of warming status.
:::
:::

The joint warming--acceleration states divide the lakes into four groups:

::: {#tbl-joint-warming-acceleration-states .list-table}

- - State
  - n
  - %

- - warming + accelerating
  - `r fmt_count(state_count("warming + accelerating"))`
  - `r fmt_pct(state_prop("warming + accelerating"), 2)`

- - warming + decelerating
  - `r fmt_count(state_count("warming + decelerating"))`
  - `r fmt_pct(state_prop("warming + decelerating"), 2)`

- - cooling + accelerating
  - `r fmt_count(state_count("cooling + accelerating"))`
  - `r fmt_pct(state_prop("cooling + accelerating"), 2)`

- - cooling + decelerating
  - `r fmt_count(state_count("cooling + decelerating"))`
  - `r fmt_pct(state_prop("cooling + decelerating"), 2)`

Joint warming and acceleration states.
:::

These four states summarize signs rather than effect sizes. In particular, a
positive acceleration means that a trend is becoming more positive, not that a
lake is necessarily warming rapidly. The pairwise distributions in
@fig-warming-acceleration-scatterplot-matrix therefore show both the central
structure and the sparse combinations that are obscured by the state counts in
@tbl-joint-warming-acceleration-states.

```{r}
#| label: fig-warming-acceleration-scatterplot-matrix
#| fig-cap: "Pairwise relationships among mean temperature, warming speed, and acceleration. Diagonal panels show probability densities; upper panels show Pearson r (with significance stars), Spearman ρ, and LOESS pseudo-R. Lower panels show KDE contours for observations above the 5th percentile of local density; lower-density observations are shown as points with opacity proportional to local density."
#| fig-width: 6
#| fig-height: 6
#| cache: true

library(GGally)

scatter_matrix_data <- lake_warming_metrics |>
  transmute(
    `Mean temperature` = raw_annual_mean_temp_mean,
    `Warming speed` = raw_annual_mean_temp_sen_slope_40yr,
    Acceleration = stl_annual_trend_diff_sen_slope_1e3
  ) |>
  filter(
    is.finite(`Mean temperature`),
    is.finite(`Warming speed`),
    is.finite(Acceleration)
  )


scatter_matrix_core_quantile <- 0.05

scatter_matrix_lower <- function(data, mapping, ...) {
  panel_data <- tibble(
    x = rlang::eval_tidy(mapping$x, data),
    y = rlang::eval_tidy(mapping$y, data)
  ) |>
    filter(is.finite(x), is.finite(y))
  kde <- split_kde_core(
    panel_data,
    core_quantile = scatter_matrix_core_quantile,
    contour_n = 6
  )

  set.seed(20260710)
  smooth_data <- kde$points |>
    slice_sample(prop = 0.1) 

  ggplot(kde$points, aes(x = x, y = y)) + 
    geom_point(
      data = kde$outliers,
      aes(alpha = density),
      colour = col_text_soft,
      size = 0.28,
      stroke = 0
    ) +
    scale_alpha_continuous(
      range = c(0.06, 0.55),
      guide = "none"
    ) +
    geom_path(
      data = kde$contours,
      aes(x = x, y = y, group = group),
      inherit.aes = FALSE,
      colour = col_inchart_line,
      linewidth = wid_inchart_line * 0.6,
      alpha = 0.5
    ) +
    geom_smooth(
      data = smooth_data,
      method = "loess",
      formula = y ~ x,
      se = TRUE,
      span = 0.65,
      color = col_inchart_line,
      alpha = 0.8,
      linewidth = wid_inchart_line,
      na.rm = TRUE
    ) +
    theme_scatter_matrix(grid = TRUE)
}

scatter_matrix_density <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_density(
      aes(y = after_stat(density)),
      fill = scales::alpha(col_inchart_fill, 0.28),
      color = col_inchart_line,
      linewidth = wid_inchart_line,
      na.rm = TRUE
    ) +
    theme_scatter_matrix(grid = FALSE)
}

scatter_matrix_cor <- function(data, mapping, ...) {
  x <- rlang::eval_tidy(mapping$x, data)
  y <- rlang::eval_tidy(mapping$y, data)

  ok <- is.finite(x) & is.finite(y)
  x <- x[ok]
  y <- y[ok]

  # Pearson correlation + p-value
  pearson_test <- cor.test(x, y, method = "pearson")
  r <- pearson_test$estimate
  p <- pearson_test$p.value

  # Spearman rank correlation
  rho <- cor(x, y, method = "spearman")

  # LOESS pseudo-R (same span as lower-panel smooth)
  loess_fit <- loess(y ~ x, span = 0.65)
  y_pred <- predict(loess_fit)
  ss_res <- sum((y - y_pred)^2)
  ss_tot <- sum((y - mean(y))^2)
  r2_loess <- 1 - ss_res / ss_tot
  r_loess <- sign(r) * sqrt(max(0, r2_loess))

  # Significance stars
  stars <- dplyr::case_when(
    p < 0.001 ~ "***",
    p < 0.01  ~ "**",
    p < 0.05  ~ "*",
    .default  = ""
  )

  label <- paste0(
    "r = ", fmt_num(r, 2), stars, "\n",
    "\u03C1 = ", fmt_num(rho, 2), "\n",
    "R[L] = ", fmt_num(r_loess, 2)
  )

  ggplot() +
    annotate(
      "text",
      x = 0.5,
      y = 0.5,
      label = label,
      size = 2.85,
      lineheight = 1.32,
      color = col_text
    ) +
    scale_x_continuous(limits = c(0, 1), expand = c(0, 0)) +
    scale_y_continuous(limits = c(0, 1), expand = c(0, 0)) +
    theme_scatter_matrix(grid = FALSE) +
    theme( 
      axis.text = element_blank(),
      axis.ticks = element_blank()
    )
}

scatter_matrix <- GGally::ggpairs(
  scatter_matrix_data,
  columns = 1:3,
  columnLabels = c(
    "Mean temperature\n(°C)",
    "Warming speed\n(°C / 40 yr)",
    "Acceleration\n(10⁻³ °C / yr²)"
  ),
  upper = list(continuous = scatter_matrix_cor),
  diag = list(continuous = scatter_matrix_density),
  lower = list(continuous = scatter_matrix_lower),
  axisLabels = "show",
  progress = FALSE
) +
  theme(
    panel.spacing = unit(0.12, "lines")
  )

scatter_matrix
```

Mean temperature is negatively associated with cumulative warming ($r = -0.47$),
while the warming--acceleration association is much weaker ($r = 0.19$). Thus,
the widespread warming reported in @tbl-warming-summary-stl-trend does not
imply a common temporal pathway: lakes with similar 40-year warming can be
accelerating, decelerating, or changing little in their recent trend.

## Spatial pattern

The hexagons in @fig-spatial-warming-acceleration-hex summarize lake-level
estimates within 5° cells containing at least five lakes. Comparing the upper
and lower panels separates where warming is strongest from where the underlying
trend is becoming more or less positive; the two patterns need not coincide.

```{r}
#| label: spatial-warming-acceleration-hex-setup
#| include: false
#| cache: true

spatial_hex <- prepare_spatial_hex(
  metrics = lake_warming_metrics,
  metadata = lake_meta_data,
  hex_height = 5,
  min_lakes = 5,
  lon_limits = c(-180, 180),
  lat_limits = c(-60, 85)
)

spatial_metrics <- spatial_hex$metrics
spatial_hex_summary <- spatial_hex$summary |>
  mutate(id = row_number())
spatial_hex_poly <- spatial_hex$polygons
lon_min <- spatial_hex$limits$lon[[1]]
lon_max <- spatial_hex$limits$lon[[2]]
lat_min <- spatial_hex$limits$lat[[1]]
lat_max <- spatial_hex$limits$lat[[2]]

# Fixed rounded mapping limits keep the two aligned legend axes readable.
warming_limit <- 2
acceleration_limit <- 3

spatial_hex_summary <- spatial_hex_summary |>
  mutate(
    warming_scaled = warming_speed / warming_limit,
    acceleration_scaled = acceleration / acceleration_limit
  )
spatial_hex_poly <- spatial_hex_poly |>
  left_join(
    spatial_hex_summary |>
      select(id, warming_scaled, acceleration_scaled),
    by = "id"
  )
```

```{r}
#| label: fig-spatial-warming-acceleration-hex
#| fig-cap: "Spatial pattern of lake warming and acceleration.
#|   Hexagons aggregate lake-level metrics in 5° geographic bins containing at least five lakes. Colors use a shared divergent scale normalized to fixed rounded mapping limits; the central colorbar shows warming values on the upper axis and acceleration values on the lower axis."
#| fig-width: 8
#| fig-height: 8
#| cache: true

spatial_map_base <- list(
  the_geom_coastline(),
  scale_x_continuous(
    breaks = seq(-180, 180, by = 60),
    expand = c(0, 0)
  ),
  scale_y_continuous(
    breaks = seq(-60, 90, by = 30),
    expand = c(0, 0)
  ),
  coord_cartesian(
    xlim = c(lon_min, lon_max),
    ylim = c(lat_min, lat_max),
    expand = FALSE
  ),
  labs(x = "Longitude", y = "Latitude"),
  theme_map_grid(base_size = 9),
  theme(
    legend.position = "bottom",
    legend.key.width = unit(20, "mm")
  )
)

spatial_hex_layer <- function(value_col) {
  geom_polygon(
    data = spatial_hex_poly,
    aes(x = lon, y = lat, group = id, fill = .data[[value_col]]),
    inherit.aes = FALSE,
    color = "white",
    linewidth = 0.08,
    alpha = 0.96
  )
}

spatial_legend_breaks <- c(-1, -0.5, 0, 0.5, 1)

spatial_legend_guide <- legendry::guide_colbar(
  first_guide = legendry::guide_axis_base(
    key = legendry::key_manual(
      aesthetic = spatial_legend_breaks,
      label = format_axis_last_unit(
        spatial_legend_breaks * warming_limit,
        "°C / 40 yr",
        1
      )
    ),
    title = "Warming"
  ),
  second_guide = legendry::guide_axis_base(
    key = legendry::key_manual(
      aesthetic = spatial_legend_breaks,
      label = format_axis_last_unit(
        spatial_legend_breaks * acceleration_limit,
        "10⁻³ °C / yr²",
        1
      )
    ),
    title = "Acceleration"
  ),
  suppress_labels = "none",
  show = FALSE,
  nbin = 60
)

spatial_shared_fill_scale <- function() {
  scale_fill_paletteer_c(
    palette = "grDevices::Blue-Red 3",
    direction = 1,
    limits = c(-1, 1),
    oob = squish,
    name = NULL,
    guide = spatial_legend_guide
  )
}

spatial_warming_map <- ggplot() +
  spatial_hex_layer("warming_scaled") +
  spatial_map_base +
  spatial_shared_fill_scale()

spatial_acceleration_map <- ggplot() +
  spatial_hex_layer("acceleration_scaled") +
  spatial_map_base +
  spatial_shared_fill_scale()

(spatial_warming_map / guide_area() / spatial_acceleration_map) +
  plot_layout(
    heights = c(1, 0.13, 1),
    guides = "collect"
  )
```

```{r}
#| label: tbl-spatial-continent-summary-setup
#| include: false
#| cache: true

spatial_continent_summary <- spatial_metrics |>
  left_join(
    lake_meta_data |>
      select(lake_id, Continent),
    by = "lake_id"
  ) |>
  mutate(
    Continent = if_else(
      is.na(Continent) | trimws(Continent) == "",
      "Unknown",
      Continent
    ),
    state = case_when(
      warming_speed > 0 & acceleration > 0 ~ "warming + accelerating",
      warming_speed > 0 & acceleration <= 0 ~ "warming + decelerating",
      warming_speed <= 0 & acceleration > 0 ~ "cooling + accelerating",
      warming_speed <= 0 & acceleration <= 0 ~ "cooling + decelerating",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(state)) |>
  group_by(Continent) |>
  summarise(
    n = n(),
    warming_pct = mean(warming_speed > 0),
    accelerating_pct = mean(acceleration > 0),
    warming_accelerating_pct = mean(state == "warming + accelerating"),
    mean_warming = mean(warming_speed, na.rm = TRUE),
    mean_acceleration = mean(acceleration, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    continent_abbr = case_when(
      Continent == "Africa" ~ "AF",
      Continent == "Asia" ~ "AS",
      Continent == "Europe" ~ "EU",
      Continent == "North America" ~ "NA",
      Continent == "South America" ~ "SA",
      Continent == "Oceania" ~ "OC",
      Continent == "Antarctica" ~ "AN",
      Continent == "Unknown" ~ "UN",
      TRUE ~ toupper(substr(Continent, 1, 2))
    )
  ) |>
  arrange(desc(n))

continent_stat <- function(abbr, column) {
  value <- spatial_continent_summary |>
    filter(continent_abbr == abbr) |>
    pull({{ column }})

  if (length(value) == 0 || !is.finite(value[[1]])) {
    return(NA_real_)
  }

  value[[1]]
}
```

::: {.column-page-inset}
::: {#tbl-spatial-continent-summary .list-table}

- - Continent
  - Warming
  - Accelerating
  - Warming + accelerating
  - Mean warming
  - Mean acceleration
- - **NA**
  - `r fmt_pct(continent_stat("NA", warming_pct))`%
  - **`r fmt_pct(continent_stat("NA", accelerating_pct))`%**
  - **`r fmt_pct(continent_stat("NA", warming_accelerating_pct))`%**
  - `r fmt_num(continent_stat("NA", mean_warming), 2)` °C / 40 yr
  - **`r fmt_num(continent_stat("NA", mean_acceleration), 2)` ×10⁻³ °C / yr²**
- - EU
  - `r fmt_pct(continent_stat("EU", warming_pct))`%
  - `r fmt_pct(continent_stat("EU", accelerating_pct))`%
  - `r fmt_pct(continent_stat("EU", warming_accelerating_pct))`%
  - `r fmt_num(continent_stat("EU", mean_warming), 2)` °C / 40 yr
  - `r fmt_num(continent_stat("EU", mean_acceleration), 2)` ×10⁻³ °C / yr²
- - AS
  - `r fmt_pct(continent_stat("AS", warming_pct))`%
  - `r fmt_pct(continent_stat("AS", accelerating_pct))`%
  - `r fmt_pct(continent_stat("AS", warming_accelerating_pct))`%
  - `r fmt_num(continent_stat("AS", mean_warming), 2)` °C / 40 yr
  - `r fmt_num(continent_stat("AS", mean_acceleration), 2)` ×10⁻³ °C / yr²
- - SA
  - `r fmt_pct(continent_stat("SA", warming_pct))`%
  - `r fmt_pct(continent_stat("SA", accelerating_pct))`%
  - `r fmt_pct(continent_stat("SA", warming_accelerating_pct))`%
  - `r fmt_num(continent_stat("SA", mean_warming), 2)` °C / 40 yr
  - `r fmt_num(continent_stat("SA", mean_acceleration), 2)` ×10⁻³ °C / yr²
- - AF
  - `r fmt_pct(continent_stat("AF", warming_pct))`%
  - `r fmt_pct(continent_stat("AF", accelerating_pct))`%
  - `r fmt_pct(continent_stat("AF", warming_accelerating_pct))`%
  - `r fmt_num(continent_stat("AF", mean_warming), 2)` °C / 40 yr
  - `r fmt_num(continent_stat("AF", mean_acceleration), 2)` ×10⁻³ °C / yr²
- - OC
  - `r fmt_pct(continent_stat("OC", warming_pct))`%
  - `r fmt_pct(continent_stat("OC", accelerating_pct))`%
  - `r fmt_pct(continent_stat("OC", warming_accelerating_pct))`%
  - `r fmt_num(continent_stat("OC", mean_warming), 2)` °C / 40 yr
  - `r fmt_num(continent_stat("OC", mean_acceleration), 2)` ×10⁻³ °C / yr²

Continent-level spatial summary
:::
:::

@tbl-spatial-continent-summary reinforces this distinction. Europe combines
high mean warming with predominantly positive acceleration, whereas North
America has high warming prevalence but a negative mean acceleration.
Continental values are descriptive aggregates: uneven lake coverage and
within-continent spatial structure mean they should not be interpreted as
area-weighted climate means.



