---
title: Descriptive
---

```{r}
#| label: setup-figures
#| include: false

source("../../../shared/R/figure-style.R")
```

## Sample coverage

```{r}
#| cache: true

# Setup for fig-lake-density-map
raw_monthly <- read_csv(
  file.path(data, "01-monthly-temperature", "output", "monthly_temperature.csv")
)
lake_meta_data <- read_csv(
  file.path(data, "00-lake-metadata", "output", "lake_metadata.csv")
)
lake_warming_metrics <- read_csv(
  file.path(
    data,
    "06-lake-warming-metrics",
    "output",
    "period12_robustfalse_ni5_no0_nt199",
    "lake_warming_metrics.csv"
  ),
  col_select = c(
    lake_id,
    raw_annual_mean_temp_mean,
    raw_annual_mean_temp_sen_slope_40yr,
    stl_annual_trend_sen_slope_40yr,
    stl_annual_trend_diff_sen_slope_1e3
  )
)

lakes_num <- nrow(raw_monthly)
start_year <- as.integer(sub(".*?(\\d{4}).*", "\\1", names(raw_monthly)[4]))
end_year <- as.integer(sub(".*?(\\d{4}).*", "\\1", tail(names(raw_monthly), 1)))
```

GLAST: `r lakes_num` lakes, `r start_year`--`r end_year`.

## Lake density map

```{r}
#| label: fig-lake-density-map
#| fig-cap: "Lake density map with 1° × 1° grid cells."
#| fig-width: 8
#| fig-height: 4
#| column: body-outset-right
#| cache: true

lake_xy <- raw_monthly |>
  select(lon, lat) |>
  filter(is.finite(lon), is.finite(lat))

lake_bins <- lake_xy |>
  mutate(
    lon_bin = pmin(floor(lon + 180), 359),
    lat_bin = pmin(floor(lat + 90), 179),
    lon = -179.5 + lon_bin,
    lat = -89.5 + lat_bin
  ) |>
  count(lon, lat, name = "n")

# Side panel counts by 1° geographic bands
lon_band_counts <- lake_xy |>
  mutate(lon_band = pmin(floor(lon + 180), 359) - 179.5) |>
  count(lon_band, name = "n")

lat_band_counts <- lake_xy |>
  mutate(lat_band = pmin(floor(lat + 90), 179) - 89.5) |>
  count(lat_band, name = "n")

# Coastline from rnaturalearth
coast <- ne_coastline(scale = 110, returnclass = "sf")
coast_line <- st_cast(coast, "LINESTRING")
coast_df <- st_coordinates(coast_line) |> as.data.frame()

# Filter data
lake_bins_plot <- lake_bins |> filter(lat >= -60)
lat_band_counts_plot <- lat_band_counts |> filter(lat_band >= -60)
coast_df_plot <- coast_df |> filter(Y >= -60)

# Lake-count composition by continent. This remains a figure-level summary:
# the data pipeline only provides lake-level metadata; the plot derives this
# lightweight aggregation locally.
continent_table <- lake_meta_data |>
  mutate(
    Continent = if_else(
      is.na(Continent) | trimws(Continent) == "",
      "Unknown",
      Continent
    )
  ) |>
  count(Continent, name = "n") |>
  arrange(desc(n)) |>
  mutate(
    row = row_number(),
    y = -row,
    n_label = scales::comma(n)
  )

table_row_lines <- tibble(
  y = -seq(0.5, nrow(continent_table) + 0.5, by = 1)
)

continent_bar <- ggplot() +
  geom_segment(
    data = table_row_lines,
    aes(x = 0.03, xend = 0.97, y = y, yend = y),
    color = scales::alpha("grey35", 0.25),
    linewidth = 0.2
  ) +
  geom_text(
    aes(x = 0.06, y = 0, label = "Continent"),
    hjust = 0,
    fontface = "bold",
    size = 2.2,
    color = "grey15"
  ) +
  geom_text(
    aes(x = 0.96, y = 0, label = "Count"),
    hjust = 1,
    fontface = "bold",
    size = 2.2,
    color = "grey15"
  ) +
  geom_text(
    data = continent_table,
    aes(x = 0.06, y = y, label = Continent),
    hjust = 0,
    size = 2.05,
    color = "grey15"
  ) +
  geom_text(
    data = continent_table,
    aes(x = 0.96, y = y, label = n_label),
    hjust = 1,
    size = 2.05,
    color = "grey15"
  ) +
  coord_cartesian(
    xlim = c(0, 1),
    ylim = c(-nrow(continent_table) - 0.65, 0.65),
    expand = FALSE,
    clip = "off"
  ) +
  theme_void() +
  theme(
    plot.background = element_rect(
      fill = "white",
      color = "grey25",
      linewidth = 0.3
    ),
    plot.margin = margin(4, 5, 4, 5)
  )

# Main map with ggside (no legend)
lake_density_figure <- ggplot() +
  geom_tile(
    data = lake_bins_plot,
    aes(x = lon, y = lat, fill = n),
    width = 1, height = 1, color = NA
  ) +
  geom_path(
    data = coast_df_plot,
    aes(x = X, y = Y, group = L1),
    color = "grey10", linewidth = 0.3
  ) +
  geom_xsideline(
    data = lon_band_counts,
    aes(x = lon_band, y = n),
    inherit.aes = FALSE,
    color = col_line,
    linewidth = 0.3
  ) +
  geom_ysideline(
    data = lat_band_counts_plot,
    aes(x = n, y = lat_band),
    inherit.aes = FALSE,
    color = col_line,
    linewidth = 0.3
  ) +
  scale_fill_paletteer_c(
    palette = "grDevices::TealRose",
    direction = 1,
    trans = "log10",
    name = "Lakes count",
    guide = legendry::guide_colbar(
      key = "log",
      second_guide = "none",
    )
  ) +
  scale_x_continuous(breaks = seq(-180, 180, by = 60), expand = c(0, 0)) +
  scale_y_continuous(breaks = seq(-60, 90, by = 60), expand = c(0, 0)) +
  scale_xsidey_continuous(
    breaks = c(600, 1200),
    limits = c(0, 1200),
    expand = c(0, 0)
  ) +
  scale_ysidex_continuous(
    breaks = c(4000, 8000),
    limits = c(0, 8000),
    labels = scales::label_number(scale_cut = scales::cut_short_scale()),
    expand = c(0, 0)
  ) +
  coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 90), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") +
  theme_atomic_facet() +
  theme(
    panel.grid = element_line(
      color = "grey85",
      linewidth = 0.3,
      linetype = "dashed"
    ),
    plot.margin = margin(10, 10, 10, 10),
    legend.position = c(0.54, 0),
    legend.justification = c(0, 0),
    legend.direction = "horizontal",
    legend.title.position = "top",
    legend.frame = element_rect(linewidth = 0.3, color = "grey10"),
    legend.ticks = element_line(linewidth = 0.3),
    legend.key.height = unit(3, "mm"),
    legend.text = element_text(size = 6),
    legend.title = element_text(size = 7),
    legend.background = element_blank(),
    ggside.axis.ticks = element_line(),
    ggside.panel.scale.x = 1.5 / 5,
    ggside.panel.scale.y = 1.5 / 12,
  )

lake_density_composite <- ggdraw(lake_density_figure) +
  draw_plot(continent_bar, x = 0.10, y = 0.16, width = 0.16, height = 0.25)

draw_fixed_aspect(lake_density_composite, aspect = 6 / 3)
```

## Long-term warming && acceleration

```{r}
#| include: false
#| cache: true

# Setup for long-term warming summaries
fmt_count <- function(x) scales::comma(x, accuracy = 1)
fmt_num <- function(x, digits = 2) formatC(x, format = "f", digits = digits)
fmt_pct <- function(x, digits = 1) fmt_num(100 * x, digits)

summarise_warming <- function(x) {
  x <- x[is.finite(x)]
  n_finite <- length(x)
  n_warming <- sum(x > 0)
  n_cooling <- sum(x <= 0)

  tibble(
    n_finite = n_finite,
    n_warming = n_warming,
    n_cooling = n_cooling,
    prop_warming = n_warming / n_finite,
    mean = mean(x),
    sd = sd(x),
    median = median(x),
    q25 = quantile(x, 0.25),
    q75 = quantile(x, 0.75)
  )
}

raw_warming_summary <- summarise_warming(
  lake_warming_metrics$raw_annual_mean_temp_sen_slope_40yr
)
stl_warming_summary <- summarise_warming(
  lake_warming_metrics$stl_annual_trend_sen_slope_40yr
)
```

```{r}
#| include: false
#| cache: true

# Setup for long-term acceleration summaries
summarise_acceleration <- function(x) {
  x <- x[is.finite(x)]
  n_finite <- length(x)
  n_positive <- sum(x > 0)
  n_negative <- sum(x <= 0)

  tibble(
    n_finite = n_finite,
    n_positive = n_positive,
    n_negative = n_negative,
    prop_positive = n_positive / n_finite,
    mean = mean(x),
    sd = sd(x),
    median = median(x),
    q25 = quantile(x, 0.25),
    q75 = quantile(x, 0.75)
  )
}

accel_summary <- summarise_acceleration(
  lake_warming_metrics$stl_annual_trend_diff_sen_slope_1e3
)

joint_state_summary <- lake_warming_metrics |>
  transmute(
    warming_speed = raw_annual_mean_temp_sen_slope_40yr,
    acceleration = stl_annual_trend_diff_sen_slope_1e3,
    state = case_when(
      warming_speed > 0 & acceleration > 0 ~ "warming + accelerating",
      warming_speed > 0 & acceleration <= 0 ~ "warming + decelerating",
      warming_speed <= 0 & acceleration > 0 ~ "cooling + accelerating",
      warming_speed <= 0 & acceleration <= 0 ~ "cooling + decelerating",
      TRUE ~ NA_character_
    )
  ) |>
  filter(is.finite(warming_speed), is.finite(acceleration), !is.na(state)) |>
  count(state, name = "n") |>
  mutate(
    prop = n / sum(n),
    state = factor(
      state,
      levels = c(
        "warming + accelerating",
        "warming + decelerating",
        "cooling + accelerating",
        "cooling + decelerating"
      )
    )
  ) |>
  arrange(state)

state_count <- function(label) {
  joint_state_summary |>
    filter(state == label) |>
    pull(n)
}

state_prop <- function(label) {
  joint_state_summary |>
    filter(state == label) |>
    pull(prop)
}
```

@tbl-warming-summary-raw and @tbl-warming-summary-stl-trend
shows the statistics of warming.
Compared with the raw annual mean temperature series,
the STL trend component is smoother and more stable,
so it identifies warming in more lakes
(91,802 / 92,245, 99.5%, versus 85,569 / 92,245, 92.8% for the raw series).
It also has a smaller standard deviation and a more concentrated quartile range,
although its mean is lower (0.573 versus 0.881 for the raw series).

::: {.column-margin}
Use Sen's slope to determine whether a lake is warming
and how is the trend like.
:::

Though almost all the lakes are warming,
the warming is not uniform.
Acceleration of `r fmt_count(accel_summary$n_positive)`
(`r fmt_pct(accel_summary$prop_positive)`%) lakes is positive.
Standard deviation and quartile range are also large, indicating that the acceleration varies widely across lakes.

::: {.column-margin} 
Here **acceleration** is measured as Sen's slope of the annual first-differenced STL trend. Positive values mean faster warming, slower cooling, or a transition toward warming.
:::

::: {.column-page-inset}
::: {#tbl-warming-summary layout-ncol=3}
::: {.list-table header-rows=0}

- - Warming count
  - `r fmt_count(raw_warming_summary$n_warming)`
    (`r fmt_pct(raw_warming_summary$prop_warming)`%)

- - Mean (°C/40yr)
  - `r fmt_num(raw_warming_summary$mean, 3)` ±
    `r fmt_num(raw_warming_summary$sd, 3)`

- - Quartile (°C/40yr)
  - `r fmt_num(raw_warming_summary$q25, 2)`,
    `r fmt_num(raw_warming_summary$median, 2)`,
    `r fmt_num(raw_warming_summary$q75, 2)`

Raw warming {#tbl-warming-summary-raw}
:::

::: {.list-table header-rows=0}

- - Warming count
  - `r fmt_count(stl_warming_summary$n_warming)`
    (`r fmt_pct(stl_warming_summary$prop_warming)`%)

- - Mean (°C/40yr)
  - `r fmt_num(stl_warming_summary$mean, 3)` ±
    `r fmt_num(stl_warming_summary$sd, 3)`

- - Quartile (°C/40yr)
  - `r fmt_num(stl_warming_summary$q25, 2)`,
    `r fmt_num(stl_warming_summary$median, 2)`,
    `r fmt_num(stl_warming_summary$q75, 2)`

STL trend warming {#tbl-warming-summary-stl-trend}
:::

::: {.list-table header-rows=0}

- - Accelerating count
  - `r fmt_count(accel_summary$n_positive)`
    (`r fmt_pct(accel_summary$prop_positive)`%)

- - Mean (°C/40yr)
  - `r fmt_num(accel_summary$mean, 3)` ±
    `r fmt_num(accel_summary$sd, 3)`

- - Quartile (°C/40yr)
  - `r fmt_num(accel_summary$q25, 2)`,
    `r fmt_num(accel_summary$median, 2)`,
    `r fmt_num(accel_summary$q75, 2)`

STL trend acceleration {#tbl-accel-summary}
:::

Summary of warming status.
:::
:::

The joint warming--acceleration states divide the lakes into four groups:

::: {#tbl-joint-warming-acceleration-states .list-table}

- - State
  - n
  - %

- - warming + accelerating
  - `r fmt_count(state_count("warming + accelerating"))`
  - `r fmt_pct(state_prop("warming + accelerating"), 2)`

- - warming + decelerating
  - `r fmt_count(state_count("warming + decelerating"))`
  - `r fmt_pct(state_prop("warming + decelerating"), 2)`

- - cooling + accelerating
  - `r fmt_count(state_count("cooling + accelerating"))`
  - `r fmt_pct(state_prop("cooling + accelerating"), 2)`

- - cooling + decelerating
  - `r fmt_count(state_count("cooling + decelerating"))`
  - `r fmt_pct(state_prop("cooling + decelerating"), 2)`

Joint warming and acceleration states.
:::

```{r}
#| label: fig-warming-acceleration-scatterplot-matrix
#| fig-cap: "Pairwise relationships among mean temperature, warming speed, and acceleration. Diagonal panels show probability densities; upper panels show Pearson r (with significance stars), Spearman ρ, and LOESS pseudo-R; lower panels show scatter plots with LOESS smooths."
#| fig-width: 4
#| fig-height: 4
#| cache: true

library(GGally)

scatter_matrix_data <- lake_warming_metrics |>
  transmute(
    `Mean temperature` = raw_annual_mean_temp_mean,
    `Warming speed` = raw_annual_mean_temp_sen_slope_40yr,
    Acceleration = stl_annual_trend_diff_sen_slope_1e3
  ) |>
  filter(
    is.finite(`Mean temperature`),
    is.finite(`Warming speed`),
    is.finite(Acceleration)
  )

scatter_matrix_grid_theme <- theme_atomic_base(base_size = 7.5, grid = TRUE) +
  theme(
    panel.grid.major = element_line(
      color = scales::alpha("grey70", 0.32),
      linewidth = 0.22
    ),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "grey76", linewidth = 0.28),
    panel.background = element_rect(fill = "white", color = NA),
    axis.text = element_text(size = 6.2, color = col_text_soft),
    axis.title = element_blank(),
    plot.margin = margin(1.5, 1.5, 1.5, 1.5)
  )

scatter_matrix_plain_theme <- theme_atomic_base(base_size = 7.5, grid = FALSE) +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "grey76", linewidth = 0.28),
    panel.background = element_rect(fill = "white", color = NA),
    axis.text = element_text(size = 6.2, color = col_text_soft),
    axis.title = element_blank(),
    plot.margin = margin(1.5, 1.5, 1.5, 1.5)
  )

scatter_matrix_bin_density <- function(x, y, bins = 120) {
  x_breaks <- seq(min(x), max(x), length.out = bins + 1)
  y_breaks <- seq(min(y), max(y), length.out = bins + 1)
  x_bin <- pmin(findInterval(x, x_breaks, all.inside = TRUE), bins)
  y_bin <- pmin(findInterval(y, y_breaks, all.inside = TRUE), bins)
  bin_id <- paste(x_bin, y_bin, sep = "_")
  as.numeric(table(bin_id)[bin_id])
}

scatter_matrix_lower <- function(data, mapping, ...) {
  x <- rlang::eval_tidy(mapping$x, data)
  y <- rlang::eval_tidy(mapping$y, data)
  panel_data <- tibble(
    x = x,
    y = y,
    density = scatter_matrix_bin_density(x, y)
  ) |>
    filter(is.finite(x), is.finite(y), is.finite(density))

  set.seed(20260710)
  smooth_data <- panel_data |>
    slice_sample(prop = 0.1)

  ggplot(panel_data, aes(x = x, y = y)) +
    geom_point(
      aes(color = density),
      size = 0.42,
      alpha = 0.32,
      stroke = 0,
      na.rm = TRUE
    ) +
    geom_smooth(
      data = smooth_data,
      method = "loess",
      formula = y ~ x,
      se = TRUE,
      span = 0.65,
      color = "grey10",
      alpha = 0.8,
      linewidth = 0.4,
      na.rm = TRUE
    ) +
    scale_color_paletteer_c(
      palette = "grDevices::Spectral",
      direction = -1,
      guide = "none"
    ) +
    scatter_matrix_grid_theme
}

scatter_matrix_density <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_density(
      aes(y = after_stat(density)),
      fill = scales::alpha("grey55", 0.28),
      color = col_line,
      linewidth = 0.4,
      na.rm = TRUE
    ) +
    scatter_matrix_plain_theme
}

scatter_matrix_cor <- function(data, mapping, ...) {
  x <- rlang::eval_tidy(mapping$x, data)
  y <- rlang::eval_tidy(mapping$y, data)

  ok <- is.finite(x) & is.finite(y)
  x <- x[ok]
  y <- y[ok]

  # Pearson correlation + p-value
  pearson_test <- cor.test(x, y, method = "pearson")
  r <- pearson_test$estimate
  p <- pearson_test$p.value

  # Spearman rank correlation
  rho <- cor(x, y, method = "spearman")

  # LOESS pseudo-R (same span as lower-panel smooth)
  loess_fit <- loess(y ~ x, span = 0.65)
  y_pred <- predict(loess_fit)
  ss_res <- sum((y - y_pred)^2)
  ss_tot <- sum((y - mean(y))^2)
  r2_loess <- 1 - ss_res / ss_tot
  r_loess <- sign(r) * sqrt(max(0, r2_loess))

  # Significance stars
  stars <- dplyr::case_when(
    p < 0.001 ~ "***",
    p < 0.01  ~ "**",
    p < 0.05  ~ "*",
    .default  = ""
  )

  label <- paste0(
    "r = ", fmt_num(r, 2), stars, "\n",
    "\u03C1 = ", fmt_num(rho, 2), "\n",
    "R[L] = ", fmt_num(r_loess, 2)
  )

  ggplot() +
    annotate(
      "text",
      x = 0.5,
      y = 0.5,
      label = label,
      size = 2.85,
      lineheight = 1.32,
      color = col_text
    ) +
    scale_x_continuous(limits = c(0, 1), expand = c(0, 0)) +
    scale_y_continuous(limits = c(0, 1), expand = c(0, 0)) +
    scatter_matrix_plain_theme +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank()
    )
}

scatter_matrix <- GGally::ggpairs(
  scatter_matrix_data,
  columns = 1:3,
  columnLabels = c(
    "Mean temperature\n(°C)",
    "Warming speed\n(°C / 40 yr)",
    "Acceleration\n(10⁻³ °C / yr²)"
  ),
  upper = list(continuous = scatter_matrix_cor),
  diag = list(continuous = scatter_matrix_density),
  lower = list(continuous = scatter_matrix_lower),
  axisLabels = "show",
  progress = FALSE
) +
  theme_atomic_base(base_size = 8, grid = TRUE) +
  theme(
    strip.background = element_rect(fill = "grey92", color = "grey75", linewidth = 0.25),
    strip.text = element_text(size = 6.8, face = "bold", color = col_text),
    panel.grid = element_blank(),
    panel.border = element_rect(color = "grey78", linewidth = 0.25),
    axis.text = element_text(size = 6.2, color = col_text_soft),
    axis.title = element_text(size = 7, color = col_text),
    panel.spacing = unit(0.12, "lines"),
    plot.margin = margin(3, 3, 3, 3)
  )

scatter_matrix
```



## Spatial pattern

```{r}
#| label: fig-spatial-warming-acceleration-hex
#| fig-cap: "Spatial pattern of lake warming and acceleration. Hexagons aggregate lake-level metrics in 5° geographic bins containing at least five lakes. Colors use a shared divergent scale normalized to fixed rounded mapping limits; the central colorbar shows warming values on the upper axis and acceleration values on the lower axis."
#| fig-width: 8
#| fig-height: 8
#| cache: true

spatial_metrics <- lake_warming_metrics |>
  transmute(
    lake_id,
    warming_speed = raw_annual_mean_temp_sen_slope_40yr,
    acceleration = stl_annual_trend_diff_sen_slope_1e3
  ) |>
  left_join(
    lake_meta_data |>
      select(lake_id, lon, lat),
    by = "lake_id"
  ) |>
  filter(
    is.finite(lon),
    is.finite(lat),
    is.finite(warming_speed),
    is.finite(acceleration)
  )

coast_world <- ne_coastline(scale = 110, returnclass = "sf") |>
  st_cast("LINESTRING") |>
  st_coordinates() |>
  as_tibble() |>
  rename(lon = X, lat = Y, segment = L1)

lon_min <- -180
lon_max <- 180
lat_min <- -60
lat_max <- 85

hex_height_deg <- 5
hex_side_deg <- hex_height_deg / sqrt(3)

hex_round <- function(q, r) {
  x <- q
  z <- r
  y <- -x - z

  rx <- round(x)
  ry <- round(y)
  rz <- round(z)

  x_diff <- abs(rx - x)
  y_diff <- abs(ry - y)
  z_diff <- abs(rz - z)

  choose_x <- x_diff > y_diff & x_diff > z_diff
  choose_y <- !choose_x & y_diff > z_diff

  rx <- ifelse(choose_x, -ry - rz, rx)
  ry <- ifelse(choose_y, -rx - rz, ry)
  rz <- ifelse(!choose_x & !choose_y, -rx - ry, rz)

  list(q = rx, r = rz)
}

spatial_hex_points <- spatial_metrics |>
  filter(
    between(lon, lon_min, lon_max),
    between(lat, lat_min, lat_max)
  ) |>
  mutate(
    x = lon - lon_min,
    y = lat - lat_min,
    q_float = (2 / 3 * x) / hex_side_deg,
    r_float = (-1 / 3 * x + sqrt(3) / 3 * y) / hex_side_deg
  )

spatial_hex_rounded <- hex_round(
  spatial_hex_points$q_float,
  spatial_hex_points$r_float
)

spatial_hex_points <- spatial_hex_points |>
  mutate(
    q_hex = spatial_hex_rounded$q,
    r_hex = spatial_hex_rounded$r
  )

spatial_hex_summary <- spatial_hex_points |>
  group_by(q_hex, r_hex) |>
  summarise(
    n = n(),
    warming_speed = mean(warming_speed, na.rm = TRUE),
    acceleration = mean(acceleration, na.rm = TRUE),
    lon_c = lon_min + hex_side_deg * 3 / 2 * first(q_hex),
    lat_c = lat_min + hex_side_deg * sqrt(3) *
      (first(r_hex) + first(q_hex) / 2),
    .groups = "drop"
  ) |>
  filter(
    n >= 5,
    between(lon_c, lon_min - hex_height_deg, lon_max + hex_height_deg),
    between(lat_c, lat_min - hex_height_deg, lat_max + hex_height_deg)
  )

# Fixed rounded mapping limits keep the two aligned legend axes readable.
# Values beyond these limits are squished to the colourbar endpoints.
warming_limit <- 2
acceleration_limit <- 3

spatial_hex_summary <- spatial_hex_summary |>
  mutate(
    warming_scaled = warming_speed / warming_limit,
    acceleration_scaled = acceleration / acceleration_limit
  )

make_hexagon <- function(cx, cy, side, id) {
  theta <- pi / 180 * seq(0, 300, by = 60)
  tibble(
    id = id,
    lon = cx + side * cos(theta),
    lat = cy + side * sin(theta)
  )
}

spatial_hex_poly <- bind_rows(lapply(seq_len(nrow(spatial_hex_summary)), function(i) {
  make_hexagon(
    spatial_hex_summary$lon_c[[i]],
    spatial_hex_summary$lat_c[[i]],
    hex_side_deg,
    i
  )
})) |>
  left_join(
    spatial_hex_summary |>
      mutate(id = row_number()),
    by = "id"
  )

spatial_map_base <- list(
  geom_path(
    data = coast_world |> filter(lat >= lat_min, lat <= lat_max),
    aes(x = lon, y = lat, group = segment),
    inherit.aes = FALSE,
    color = "grey10",
    linewidth = 0.3
  ),
  scale_x_continuous(
    breaks = seq(-180, 180, by = 60),
    expand = c(0, 0)
  ),
  scale_y_continuous(
    breaks = seq(-60, 90, by = 60),
    expand = c(0, 0)
  ),
  coord_cartesian(
    xlim = c(lon_min, lon_max),
    ylim = c(lat_min, lat_max),
    expand = FALSE
  ),
  labs(x = "Longitude", y = "Latitude"),
  theme_atomic_map(base_size = 9),
  theme(
    panel.grid = element_line(
      color = "grey85",
      linewidth = 0.3,
      linetype = "dashed"
    ),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.text.position = "top",
    legend.title = element_text(size = 6.4, color = col_text),
    legend.frame = element_rect(linewidth = 0.3, color = "grey10"),
    legend.ticks = element_line(linewidth = 0.3),
    legend.key.width = unit(30, "mm"),
    legend.key.height = unit(3.2, "mm"),
    legend.text = element_text(size = 6),
    legend.background = element_blank(),
    plot.title = element_text(size = 10, face = "bold", color = col_text),
    plot.margin = margin(4, 5, 4, 5)
  )
)

spatial_hex_layer <- function(value_col) {
  geom_polygon(
    data = spatial_hex_poly,
    aes(x = lon, y = lat, group = id, fill = .data[[value_col]]),
    inherit.aes = FALSE,
    color = "white",
    linewidth = 0.08,
    alpha = 0.96
  )
}

spatial_fmt_axis <- function(x, digits = 2) {
  label <- formatC(x, format = "f", digits = digits)
  label <- sub("0+$", "", label)
  label <- sub("\\.$", "", label)
  label[label == "-0"] <- "0"
  label
}

spatial_fmt_axis_unit <- function(x, unit, digits = 2) {
  label <- spatial_fmt_axis(x, digits)
  label[length(label)] <- paste(label[length(label)], unit)
  label
}

spatial_legend_breaks <- c(-1, -0.5, 0, 0.5, 1)

spatial_legend_guide <- legendry::guide_colbar(
  first_guide = legendry::guide_axis_base(
    key = legendry::key_manual(
      aesthetic = spatial_legend_breaks,
      label = spatial_fmt_axis_unit(
        spatial_legend_breaks * warming_limit,
        "°C / 40 yr",
        1
      )
    ),
    title = "Warming"
  ),
  second_guide = legendry::guide_axis_base(
    key = legendry::key_manual(
      aesthetic = spatial_legend_breaks,
      label = spatial_fmt_axis_unit(
        spatial_legend_breaks * acceleration_limit,
        "10⁻³ °C / yr²",
        1
      )
    ),
    title = "Acceleration"
  ),
  suppress_labels = "none",
  show = FALSE,
  nbin = 60
)

spatial_shared_fill_scale <- function() {
  scale_fill_paletteer_c(
    palette = "grDevices::Blue-Red 3",
    direction = 1,
    limits = c(-1, 1),
    oob = squish,
    name = NULL,
    guide = spatial_legend_guide
  )
}

spatial_warming_map <- ggplot() +
  spatial_hex_layer("warming_scaled") +
  spatial_map_base +
  spatial_shared_fill_scale()
  # labs(title = "A  Warming speed")

spatial_acceleration_map <- ggplot() +
  spatial_hex_layer("acceleration_scaled") +
  spatial_map_base +
  spatial_shared_fill_scale()
  # labs(title = "B  Acceleration")

(spatial_warming_map / guide_area() / spatial_acceleration_map) +
  plot_layout(
    heights = c(1, 0.13, 1),
    guides = "collect"
  )
```

```{r}
#| label: tbl-spatial-continent-summary-setup
#| include: false
#| cache: true

spatial_continent_summary <- spatial_metrics |>
  left_join(
    lake_meta_data |>
      select(lake_id, Continent),
    by = "lake_id"
  ) |>
  mutate(
    Continent = if_else(
      is.na(Continent) | trimws(Continent) == "",
      "Unknown",
      Continent
    ),
    state = case_when(
      warming_speed > 0 & acceleration > 0 ~ "warming + accelerating",
      warming_speed > 0 & acceleration <= 0 ~ "warming + decelerating",
      warming_speed <= 0 & acceleration > 0 ~ "cooling + accelerating",
      warming_speed <= 0 & acceleration <= 0 ~ "cooling + decelerating",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(state)) |>
  group_by(Continent) |>
  summarise(
    n = n(),
    warming_pct = mean(warming_speed > 0),
    accelerating_pct = mean(acceleration > 0),
    warming_accelerating_pct = mean(state == "warming + accelerating"),
    mean_warming = mean(warming_speed, na.rm = TRUE),
    mean_acceleration = mean(acceleration, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    continent_abbr = case_when(
      Continent == "Africa" ~ "AF",
      Continent == "Asia" ~ "AS",
      Continent == "Europe" ~ "EU",
      Continent == "North America" ~ "NA",
      Continent == "South America" ~ "SA",
      Continent == "Oceania" ~ "OC",
      Continent == "Antarctica" ~ "AN",
      Continent == "Unknown" ~ "UN",
      TRUE ~ toupper(substr(Continent, 1, 2))
    )
  ) |>
  arrange(desc(n))

continent_stat <- function(abbr, column) {
  value <- spatial_continent_summary |>
    filter(continent_abbr == abbr) |>
    pull({{ column }})

  if (length(value) == 0 || !is.finite(value[[1]])) {
    return(NA_real_)
  }

  value[[1]]
}
```

::: {.column-page-inset}
::: {#tbl-spatial-continent-summary .list-table}

- - Continent
  - Warming
  - Accelerating
  - Warming + accelerating
  - Mean warming
  - Mean acceleration
- - **NA**
  - `r fmt_pct(continent_stat("NA", warming_pct))`%
  - **`r fmt_pct(continent_stat("NA", accelerating_pct))`%**
  - **`r fmt_pct(continent_stat("NA", warming_accelerating_pct))`%**
  - `r fmt_num(continent_stat("NA", mean_warming), 2)` °C / 40 yr
  - **`r fmt_num(continent_stat("NA", mean_acceleration), 2)` ×10⁻³ °C / yr²**
- - EU
  - `r fmt_pct(continent_stat("EU", warming_pct))`%
  - `r fmt_pct(continent_stat("EU", accelerating_pct))`%
  - `r fmt_pct(continent_stat("EU", warming_accelerating_pct))`%
  - `r fmt_num(continent_stat("EU", mean_warming), 2)` °C / 40 yr
  - `r fmt_num(continent_stat("EU", mean_acceleration), 2)` ×10⁻³ °C / yr²
- - AS
  - `r fmt_pct(continent_stat("AS", warming_pct))`%
  - `r fmt_pct(continent_stat("AS", accelerating_pct))`%
  - `r fmt_pct(continent_stat("AS", warming_accelerating_pct))`%
  - `r fmt_num(continent_stat("AS", mean_warming), 2)` °C / 40 yr
  - `r fmt_num(continent_stat("AS", mean_acceleration), 2)` ×10⁻³ °C / yr²
- - SA
  - `r fmt_pct(continent_stat("SA", warming_pct))`%
  - `r fmt_pct(continent_stat("SA", accelerating_pct))`%
  - `r fmt_pct(continent_stat("SA", warming_accelerating_pct))`%
  - `r fmt_num(continent_stat("SA", mean_warming), 2)` °C / 40 yr
  - `r fmt_num(continent_stat("SA", mean_acceleration), 2)` ×10⁻³ °C / yr²
- - AF
  - `r fmt_pct(continent_stat("AF", warming_pct))`%
  - `r fmt_pct(continent_stat("AF", accelerating_pct))`%
  - `r fmt_pct(continent_stat("AF", warming_accelerating_pct))`%
  - `r fmt_num(continent_stat("AF", mean_warming), 2)` °C / 40 yr
  - `r fmt_num(continent_stat("AF", mean_acceleration), 2)` ×10⁻³ °C / yr²
- - OC
  - `r fmt_pct(continent_stat("OC", warming_pct))`%
  - `r fmt_pct(continent_stat("OC", accelerating_pct))`%
  - `r fmt_pct(continent_stat("OC", warming_accelerating_pct))`%
  - `r fmt_num(continent_stat("OC", mean_warming), 2)` °C / 40 yr
  - `r fmt_num(continent_stat("OC", mean_acceleration), 2)` ×10⁻³ °C / yr²

Continent-level spatial summary
:::
:::

```{r}
#| label: fig-spatial-warming-acceleration-states
#| fig-cap: "Geographic distribution of four warming--acceleration states. Points show lakes colored by state; marginal ridgelines show longitude and latitude counts in 2° bands."
#| fig-width: 10.5
#| fig-height: 6.8
#| cache: true
#| eval: false
#| include: false

spatial_state_levels <- c(
  "warming + accelerating",
  "warming + decelerating",
  "cooling + accelerating",
  "cooling + decelerating"
)

spatial_state_labels <- c(
  "warming + accelerating" = "Warming + accelerating",
  "warming + decelerating" = "Warming + decelerating",
  "cooling + accelerating" = "Cooling + accelerating",
  "cooling + decelerating" = "Cooling + decelerating"
)

spatial_area_limits <- lake_meta_data |>
  filter(is.finite(Lake_area), Lake_area > 0) |>
  pull(Lake_area) |>
  quantile(c(0.05, 0.95), na.rm = TRUE)

spatial_state_data <- spatial_metrics |>
  left_join(
    lake_meta_data |>
      select(lake_id, Lake_area),
    by = "lake_id"
  ) |>
  filter(is.finite(Lake_area), Lake_area > 0) |>
  mutate(
    area_clipped = pmin(pmax(Lake_area, spatial_area_limits[[1]]), spatial_area_limits[[2]]),
    point_size = rescale(
      sqrt(log10(area_clipped)),
      to = c(0.18, 1.2),
      from = range(sqrt(log10(area_clipped)), na.rm = TRUE)
    ),
    state = case_when(
      warming_speed > 0 & acceleration > 0 ~ spatial_state_levels[[1]],
      warming_speed > 0 & acceleration <= 0 ~ spatial_state_levels[[2]],
      warming_speed <= 0 & acceleration > 0 ~ spatial_state_levels[[3]],
      warming_speed <= 0 & acceleration <= 0 ~ spatial_state_levels[[4]],
      TRUE ~ NA_character_
    ),
    state = factor(state, levels = spatial_state_levels)
  ) |>
  filter(
    !is.na(state),
    is.finite(point_size),
    between(lon, lon_min, lon_max),
    between(lat, lat_min, lat_max)
  )

spatial_state_summary <- spatial_state_data |>
  count(state, name = "n") |>
  mutate(
    prop = n / sum(n),
    label = paste0(spatial_state_labels[as.character(state)], "\n", percent(prop, accuracy = 0.1))
  )

spatial_bin_width <- 2
spatial_lon_centers <- seq(lon_min + spatial_bin_width / 2, lon_max - spatial_bin_width / 2,
  by = spatial_bin_width
)
spatial_lat_centers <- seq(lat_min + spatial_bin_width / 2, lat_max - spatial_bin_width / 2,
  by = spatial_bin_width
)

smooth_spatial_count <- function(x) {
  smoothed <- as.numeric(stats::filter(x, rep(1 / 3, 3), sides = 2))
  ifelse(is.na(smoothed), x, smoothed)
}

ridge_gap <- 0.55
ridge_height <- 0.92
ridge_baseline <- rev(seq_along(spatial_state_levels)) * ridge_gap
names(ridge_baseline) <- spatial_state_levels

spatial_lon_ridges <- spatial_state_data |>
  mutate(
    lon_bin = floor((lon - lon_min) / spatial_bin_width),
    lon_center = lon_min + lon_bin * spatial_bin_width + spatial_bin_width / 2
  ) |>
  count(state, lon_center, name = "n") |>
  complete(
    state = factor(spatial_state_levels, levels = spatial_state_levels),
    lon_center = spatial_lon_centers,
    fill = list(n = 0)
  ) |>
  group_by(state) |>
  mutate(
    n_smooth = smooth_spatial_count(sqrt(n)),
    baseline = ridge_baseline[as.character(state)]
  ) |>
  ungroup() |>
  mutate(
    n_scaled = n_smooth / max(n_smooth, na.rm = TRUE),
    y = baseline + n_scaled * ridge_height
  ) |>
  arrange(desc(baseline), lon_center)

spatial_lat_ridges <- spatial_state_data |>
  mutate(
    lat_bin = floor((lat - lat_min) / spatial_bin_width),
    lat_center = lat_min + lat_bin * spatial_bin_width + spatial_bin_width / 2
  ) |>
  count(state, lat_center, name = "n") |>
  complete(
    state = factor(spatial_state_levels, levels = spatial_state_levels),
    lat_center = spatial_lat_centers,
    fill = list(n = 0)
  ) |>
  group_by(state) |>
  mutate(
    n_smooth = smooth_spatial_count(sqrt(n)),
    baseline = ridge_baseline[as.character(state)]
  ) |>
  ungroup() |>
  mutate(
    n_scaled = n_smooth / max(n_smooth, na.rm = TRUE),
    x = baseline + n_scaled * ridge_height
  ) |>
  arrange(desc(baseline), lat_center)

state_map <- ggplot() +
  geom_path(
    data = coast_world |> filter(lat >= lat_min, lat <= lat_max),
    aes(x = lon, y = lat, group = segment),
    color = "grey35",
    linewidth = 0.16
  ) +
  geom_point(
    data = spatial_state_data,
    aes(x = lon, y = lat, color = state, size = point_size),
    alpha = 0.32,
    stroke = 0
  ) +
  scale_color_manual(
    values = pal_state,
    labels = spatial_state_labels,
    name = NULL
  ) +
  scale_size_identity(guide = "none") +
  coord_fixed(
    ratio = 1,
    xlim = c(lon_min, lon_max),
    ylim = c(lat_min, lat_max),
    expand = FALSE
  ) +
  labs(x = NULL, y = NULL) +
  theme_atomic_map(base_size = 9) +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 7),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = margin(0, 0, 0, 0)
  )

ridge_theme <- theme_void(base_size = 8) +
  theme(
    legend.position = "none",
    plot.margin = margin(0, 0, 0, 0)
  )

state_top_ridges <- ggplot(
  spatial_lon_ridges,
  aes(x = lon_center, group = state, color = state, fill = state)
) +
  geom_ribbon(aes(ymin = baseline, ymax = y), alpha = 0.23, color = NA) +
  geom_line(aes(y = y), linewidth = 0.45) +
  geom_hline(
    data = tibble(
      state = factor(spatial_state_levels, levels = spatial_state_levels),
      baseline = ridge_baseline
    ),
    aes(yintercept = baseline, color = state),
    linewidth = 0.18,
    alpha = 0.45
  ) +
  scale_color_manual(values = pal_state, guide = "none") +
  scale_fill_manual(values = pal_state, guide = "none") +
  scale_x_continuous(limits = c(lon_min, lon_max), expand = c(0, 0)) +
  scale_y_continuous(
    limits = c(0.45, max(ridge_baseline) + ridge_height + 0.28),
    expand = c(0, 0)
  ) +
  ridge_theme

state_right_ridges <- ggplot(
  spatial_lat_ridges,
  aes(y = lat_center, group = state, color = state, fill = state)
) +
  geom_ribbon(aes(xmin = baseline, xmax = x), orientation = "y", alpha = 0.23, color = NA) +
  geom_line(aes(x = x), linewidth = 0.45, orientation = "y") +
  geom_vline(
    data = tibble(
      state = factor(spatial_state_levels, levels = spatial_state_levels),
      baseline = ridge_baseline
    ),
    aes(xintercept = baseline, color = state),
    linewidth = 0.18,
    alpha = 0.45
  ) +
  scale_color_manual(values = pal_state, guide = "none") +
  scale_fill_manual(values = pal_state, guide = "none") +
  scale_y_continuous(limits = c(lat_min, lat_max), expand = c(0, 0)) +
  scale_x_continuous(
    limits = c(0.45, max(ridge_baseline) + ridge_height + 0.28),
    expand = c(0, 0)
  ) +
  ridge_theme

state_share_plot <- ggplot(spatial_state_summary, aes(x = prop, y = state, fill = state)) +
  geom_col(width = 0.62, color = "white", linewidth = 0.25) +
  geom_text(
    aes(label = percent(prop, accuracy = 0.1)),
    hjust = -0.08,
    size = 2.5,
    color = col_text
  ) +
  scale_fill_manual(values = pal_state, guide = "none") +
  scale_x_continuous(labels = percent, expand = expansion(mult = c(0, 0.22))) +
  scale_y_discrete(labels = spatial_state_labels) +
  labs(x = NULL, y = NULL) +
  theme_atomic_base(base_size = 8, grid = FALSE) +
  theme(
    axis.text.y = element_text(size = 6.6, color = col_text),
    axis.text.x = element_text(size = 6.5, color = col_text_soft),
    panel.grid.major.x = element_line(color = "grey85", linewidth = 0.25),
    panel.grid.major.y = element_blank(),
    legend.position = "none",
    plot.margin = margin(2, 2, 2, 2)
  )

state_legend <- get_legend(state_map + theme(legend.position = "right"))
state_map <- state_map + theme(legend.position = "none")

state_figure <- ggdraw() +
  draw_label(
    "Geographic distribution of warming--acceleration states",
    x = 0.07,
    y = 0.98,
    hjust = 0,
    vjust = 1,
    fontface = "bold",
    size = 13,
    color = col_text
  ) +
  draw_plot(state_top_ridges, x = 0.07, y = 0.72, width = 0.63, height = 0.18) +
  draw_plot(state_map, x = 0.07, y = 0.14, width = 0.63, height = 0.56) +
  draw_plot(state_right_ridges, x = 0.72, y = 0.14, width = 0.14, height = 0.56) +
  draw_plot(state_share_plot, x = 0.72, y = 0.73, width = 0.23, height = 0.17) +
  draw_plot(state_legend, x = 0.74, y = 0.42, width = 0.24, height = 0.22)

state_figure
```

