---
title: Regional
---

```{r}
#| label: setup-figures-regional
#| include: false

source("../../../shared/R/figure-style.R")
source("../../../shared/R/regional-helpers.R")
```

## Temporal response types

We classify lakes by the shape of their 1981--2020 annual STL-trend response.
The current canonical classification uses per-lake baseline anomalies relative
to 1981--1990, PCA retaining 95% of the trajectory variance, and $k$-means
clustering. The selected solution has five clusters ($K = 5$).

These clusters describe low-frequency lake-surface-temperature response types:
lakes with similar STL-trend trajectories are grouped together regardless of
continental labels. They complement the lake-level warming and acceleration
metrics in Chapter 1: those metrics summarize overall change and its recent
evolution, whereas the clusters retain the sequence of change through time.

@fig-stl-cluster-map-violins asks whether these trajectory types have geographic
structure. The map and violin panels place each type in the broader
distributions of mean temperature, cumulative warming, and acceleration; cluster
labels should therefore be read as response descriptions rather than regional
categories.

```{r}
#| label: regional-cluster-data-setup
#| include: false
#| cache: true

regional_cluster_data <- prepare_regional_cluster_data()
plot_df <- regional_cluster_data$plot_data
plot_df_map <- regional_cluster_data$map_data
cluster_summary <- regional_cluster_data$summary
cluster_cols <- regional_cluster_data$colours
cluster_labels <- stats::setNames(
  cluster_summary$label,
  as.character(cluster_summary$cluster_label)
)
cluster_metric_long <- prepare_cluster_metric_long(plot_df)
cluster_density_data <- prepare_cluster_density_data(plot_df)
cluster_table <- make_cluster_table_accessors(cluster_summary)
lon_min <- regional_cluster_data$map_limits$lon[[1]]
lon_max <- regional_cluster_data$map_limits$lon[[2]]
lat_min <- regional_cluster_data$map_limits$lat[[1]]
lat_max <- regional_cluster_data$map_limits$lat[[2]]
```

```{r}
#| label: fig-stl-cluster-map-violins
#| fig-cap: "STL-trend response clusters and their thermal metric distributions. (A) Global spatial distribution of the five clusters. (B-D) Violin plots of mean temperature, Sen slope, and STL-trend difference slope by cluster."
#| fig-width: 8
#| fig-height: 6
#| cache: true

# --- Map (no side panels) ---
map_plot <- ggplot() +
  the_geom_coastline() +
  geom_point(
    data = plot_df_map,
    aes(x = lon, y = lat, colour = cluster_label, size = point_size),
    alpha = 0.43,
    stroke = 0
  ) +
  scale_color_manual(
    values = cluster_cols,
    labels = cluster_labels,
    name = NULL
  ) +
  scale_size_identity(guide = "none") +
  scale_x_continuous(breaks = seq(-180, 180, by = 60), expand = c(0, 0)) +
  scale_y_continuous(breaks = seq(-60, 90, by = 30), expand = c(0, 0)) +
  coord_cartesian(xlim = c(lon_min, lon_max), ylim = c(lat_min, lat_max), expand = FALSE) +
  labs(
    title = "A  Spatial distribution of STL-trend response clusters (K=5)",
    x = "Longitude",
    y = "Latitude"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2.6, alpha = 0.9), nrow = 1)) +
  theme_map_grid()

# --- Violin plots ---
violin_theme <- theme_base(base_size = 8, grid = TRUE)

violin_base <- ggplot(
  cluster_metric_long,
  aes(x = cluster_label, y = value, fill = cluster_label, color = cluster_label)
) +
  geom_violin(
    scale = "width",
    trim = TRUE,
    alpha = 0.20,
    linewidth = 0.35,
    na.rm = TRUE
  ) +
  geom_boxplot(
    width = 0.13,
    outlier.shape = NA,
    alpha = 0.55,
    linewidth = 0.25,
    color = col_inchart_line,
    na.rm = TRUE
  ) +
  facet_wrap(~metric, scales = "free_y", nrow = 1) +
  scale_fill_manual(values = cluster_cols, guide = "none") +
  scale_color_manual(values = cluster_cols, guide = "none") +
  labs(x = "Cluster", y = NULL) +
  violin_theme

# --- Combined patchwork layout ---
p_combined <- map_plot / violin_base +
  plot_layout(heights = c(2.2, 1)) +
  plot_annotation(
    title = "STL-trend response clusters of global lakes (K=5)",
    theme = theme(
      plot.title = element_text(size = 13, face = "bold", color = col_text)
    )
  )

save_figure("fig-stl-cluster-map-ridges.png", p_combined, width = 12, height = 9)
p_combined
```

::: {#tbl-stl-cluster-summary .list-table .column-page-inset}

- - Cluster
  - Response type
  - n lakes
  - 2020 anomaly
  - Short interpretation
- - C1
  - `r cluster_table$value("C1", "response_type")`
  - `r cluster_table$count("C1")`
  - `r cluster_table$anomaly("C1")`
  - `r cluster_table$value("C1", "short_interpretation")`
- - C2
  - `r cluster_table$value("C2", "response_type")`
  - `r cluster_table$count("C2")`
  - `r cluster_table$anomaly("C2")`
  - `r cluster_table$value("C2", "short_interpretation")`
- - C3
  - `r cluster_table$value("C3", "response_type")`
  - `r cluster_table$count("C3")`
  - `r cluster_table$anomaly("C3")`
  - `r cluster_table$value("C3", "short_interpretation")`
- - C4
  - `r cluster_table$value("C4", "response_type")`
  - `r cluster_table$count("C4")`
  - `r cluster_table$anomaly("C4")`
  - `r cluster_table$value("C4", "short_interpretation")`
- - C5
  - `r cluster_table$value("C5", "response_type")`
  - `r cluster_table$count("C5")`
  - `r cluster_table$anomaly("C5")`
  - `r cluster_table$value("C5", "short_interpretation")`

Summary of STL-trend response clusters
:::

As shown in @tbl-stl-cluster-summary, C4 is the largest class
(`r cluster_table$count("C4")` lakes) and remains closest to its baseline by
2020. At the other end, C3 has the largest 2020 anomaly
(`r cluster_table$anomaly("C3")`) and represents sustained warming. C1 and C2
concentrate the late-accelerating trajectories, while C5 captures early warming
followed by a plateau. Their spatial pattern in @fig-stl-cluster-map-violins is
structured rather than uniformly mixed, but the classification itself was not
fitted to geography.

@fig-cluster-density-contours compares the joint distributions of warming speed
and acceleration. Contours are normalized separately within each cluster, so
they show where a cluster's typical observations lie; they do not compare
cluster sample sizes.

```{r}
#| label: fig-cluster-density-contours
#| fig-cap: "Two-dimensional density of warming speed and acceleration by STL-trend response cluster. Contour lines show per-cluster density; marginal panels show univariate density by cluster."
#| fig-width: 6
#| fig-height: 6
#| cache: true

# --- Build main plot ---
density_main <- ggplot(
  cluster_density_data,
  aes(x = warming_speed, y = acceleration)
) +
  stat_density_2d(
    geom = "polygon",
    contour_var = "ndensity",
    breaks = c(0.5, 0.65, 0.80, 0.95),
    n = 250,
    aes(fill = cluster_label, alpha = after_stat(level)),
    na.rm = TRUE
  ) +
  geom_density_2d(
    aes(group = cluster_label),
    contour_var = "ndensity",
    breaks = c(0.5, 0.65, 0.8, 0.95),
    n = 250,
    linewidth = 0.4,
    colour = alpha(col_inchart_geom_border, 0.8),
    na.rm = TRUE
  ) +
  scale_fill_manual(
    values = cluster_cols,
    labels = paste0("C", 1:5),
    name = NULL,
    guide = guide_legend(
      nrow = 2,
      byrow = TRUE,
      order = 1,
      override.aes = list(
        alpha = 0.30,
        colour = alpha(col_inchart_geom_border, 0.8),
        linewidth = 0.4
      )
    )
  ) +
  scale_alpha_continuous(range = c(0.05, 0.30), guide = "none") +
  coord_cartesian(xlim = c(-1, 2), ylim = c(-2.5, 3.5), expand = FALSE, clip = "off") +
  labs(
    x = "Sen slope (°C / 40 yr)",
    y = "Acceleration (10⁻³ °C / yr²)"
  ) +
  theme_base(grid = TRUE) +
  theme_facet(base_size = 9) +
  theme(
    panel.grid = element_line(
      color = col_grid_minor,
      linewidth = wid_grid,
      linetype = "dashed"
    ),
    plot.margin = margin(8, 8, 2, 8),
    legend.position = c(0.02, 0),
    legend.justification = c(0, 0),
    legend.direction = "horizontal"
  )

# --- Assemble final figure ---
p_density <- density_main

save_figure("fig-cluster-density-contours.png", p_density, width = 8, height = 8)
p_density
```

The density fields in @fig-cluster-density-contours overlap, indicating that no
single pair of summary metrics fully reproduces the trajectory classification.
Nevertheless, C1 and C2 occupy more positive-acceleration regions, whereas C4
and C5 extend toward negative acceleration. This separation is consistent with
their contrasting late-period responses and shows why the full STL-trend
trajectory adds information beyond a single Sen slope.
