---
title: Regional
draft: true
---

```{r}
#| label: setup-figures-regional
#| include: false

source("../../../shared/R/figure-style.R")
```

## Temporal response types

We classify lakes by the shape of their 1981--2020 annual STL-trend response.
The current canonical classification uses per-lake baseline anomalies relative
to 1981--1990, PCA retaining 95% of the trajectory variance, and $k$-means
clustering. The selected solution has five clusters ($K = 5$).

These clusters describe low-frequency lake-surface-temperature response types:
lakes with similar STL-trend trajectories are grouped together regardless of
continental labels.

::: {.callout-note}
## Finding 10
Five STL-trend response types are identified. The classes separate weak or
plateauing responses from late-accelerating and sustained-warming trajectories;
their spatial distributions provide an external check on whether the data-driven
classification corresponds to coherent regional climate-response patterns.
:::

```{r}
#| label: fig-stl-cluster-map-violins
#| fig-cap: "STL-trend response clusters and their thermal metric distributions. (A) Global spatial distribution of the five clusters. (B-D) Violin plots of mean temperature, Sen slope, and STL-trend difference slope by cluster."
#| fig-width: 12
#| fig-height: 9
#| cache: true

cluster_dir <- file.path(
  data,
  "07-warming-response-clustering",
  "output",
  "stl_trend_period12_robustfalse_ni5_no0_nt199_baseline1981_1990_pca095_k4-8"
)

clusters <- read_csv(
  file.path(cluster_dir, "lake_clusters_recommended.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    cluster_label = paste0("C", cluster),
    cluster_label = factor(cluster_label, levels = paste0("C", 1:5))
  )

profiles <- read_csv(
  file.path(cluster_dir, "cluster_profiles_K5.csv"),
  show_col_types = FALSE
) %>%
  mutate(cluster_label = factor(paste0("C", cluster), levels = paste0("C", 1:5)))

lake_meta <- read_csv(
  file.path(data, "00-lake-metadata", "output", "lake_metadata.csv"),
  show_col_types = FALSE,
  col_select = c(lake_id, Lake_area, log_Lake_area, Continent)
)

regional_metric_data <- read_csv(
  file.path(
    data,
    "06-lake-warming-metrics",
    "output",
    "period12_robustfalse_ni5_no0_nt199",
    "lake_warming_metrics.csv"
  ),
  show_col_types = FALSE,
  col_select = c(
    lake_id,
    mean_temp = raw_annual_mean_temp_mean,
    warming_speed = raw_annual_mean_temp_sen_slope_40yr,
    acceleration = stl_annual_trend_diff_sen_slope_1e3
  )
)

cluster_cols <- c(
  C1 = "#9e2a2f", C2 = "#d4770b", C3 = "#2d8659", C4 = "#00a5cf", C5 = "#7b4d9e"
)

cluster_names <- c(
  C1 = "Late-accelerating moderate warming",
  C2 = "Early decline / late rebound",
  C3 = "Strong sustained warming",
  C4 = "Weak / near-stable warming",
  C5 = "Early warming then plateau"
)

cluster_interpretations <- c(
  C1 = "Slow early change followed by stronger post-2000 warming",
  C2 = "Flat or slightly negative early trajectory, then late warming rebound",
  C3 = "Persistent warming throughout the full 40-year record",
  C4 = "Largest low-response class, close to baseline through time",
  C5 = "Early warming followed by flattening or slight decline"
)

plot_df <- clusters %>%
  left_join(lake_meta, by = "lake_id") %>%
  left_join(regional_metric_data, by = "lake_id") %>%
  mutate(
    point_size = scales::rescale(
      pmin(pmax(log_Lake_area, 0), 2),
      to = c(0.35, 1.35),
      from = c(0, 2)
    ),
    point_size = if_else(is.finite(point_size), point_size, 0.45)
  )

cluster_summary <- plot_df %>%
  group_by(cluster_label) %>%
  summarise(
    n = n(),
    mean_temp = mean(mean_temp, na.rm = TRUE),
    warming_speed = mean(warming_speed, na.rm = TRUE),
    acceleration = mean(acceleration, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(
    profiles %>%
      group_by(cluster_label) %>%
      summarise(
        anomaly_2020 = anomaly_mean[year == 2020][1],
        .groups = "drop"
      ),
    by = "cluster_label"
  ) %>%
  mutate(
    response_type = unname(cluster_names[as.character(cluster_label)]),
    short_interpretation = unname(cluster_interpretations[as.character(cluster_label)]),
    label = paste0(
      as.character(cluster_label), ": ",
      response_type, "\n",
      comma(n), " lakes; 2020 anomaly ", sprintf("%+.2f", anomaly_2020), " °C"
    )
  ) %>%
  arrange(cluster_label)

cluster_table_value <- function(cluster, column) {
  row <- match(cluster, as.character(cluster_summary$cluster_label))
  if (is.na(row)) {
    return(NA)
  }
  cluster_summary[[column]][[row]]
}

cluster_table_count <- function(cluster) {
  comma(cluster_table_value(cluster, "n"))
}

cluster_table_anomaly <- function(cluster) {
  paste0(sprintf("%+.2f", cluster_table_value(cluster, "anomaly_2020")), " °C")
}

cluster_labels <- cluster_summary$label
names(cluster_labels) <- as.character(cluster_summary$cluster_label)

coast <- ne_coastline(scale = 110, returnclass = "sf") %>%
  st_cast("LINESTRING") %>%
  st_coordinates() %>%
  as_tibble() %>%
  rename(lon = X, lat = Y, segment = L1)

lon_min <- -180; lon_max <- 180; lat_min <- -60; lat_max <- 90

plot_df_map <- plot_df %>%
  filter(lat >= lat_min, lat <= lat_max) %>%
  mutate(
    cluster_label = factor(cluster_label, levels = paste0("C", 1:5))
  )
coast_plot <- coast %>% filter(lat >= lat_min, lat <= lat_max)

# --- Map (no side panels) ---
map_plot <- ggplot() +
  geom_path(
    data = coast_plot,
    aes(x = lon, y = lat, group = segment),
    color = "grey10",
    linewidth = 0.3
  ) +
  geom_point(
    data = plot_df_map,
    aes(x = lon, y = lat, colour = cluster_label, size = point_size),
    alpha = 0.43,
    stroke = 0
  ) +
  scale_color_manual(
    values = cluster_cols,
    labels = cluster_labels,
    name = NULL
  ) +
  scale_size_identity(guide = "none") +
  scale_x_continuous(breaks = seq(-180, 180, by = 60), expand = c(0, 0)) +
  scale_y_continuous(breaks = seq(-60, 90, by = 30), expand = c(0, 0)) +
  coord_cartesian(xlim = c(lon_min, lon_max), ylim = c(lat_min, lat_max), expand = FALSE) +
  labs(
    title = "A  Spatial distribution of STL-trend response clusters (K=5)",
    x = "Longitude",
    y = "Latitude"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2.6, alpha = 0.9), nrow = 1)) +
  theme_atomic_base(base_size = 9, grid = FALSE) +
  theme(
    panel.grid = element_line(color = "grey85", linewidth = 0.3, linetype = "dashed"),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.key.height = unit(3, "mm"),
    legend.key.width = unit(7, "mm"),
    legend.text = element_text(size = 7.2),
    legend.background = element_blank()
  )

# --- Violin plots ---
cluster_metric_long <- plot_df %>%
  transmute(
    cluster_label,
    `Mean temperature\n(°C)` = mean_temp,
    `Sen slope\n(°C / 40 yr)` = warming_speed,
    `STL trend diff slope\n(10-3 °C / yr2)` = acceleration
  ) %>%
  pivot_longer(
    cols = -cluster_label,
    names_to = "metric",
    values_to = "value"
  ) %>%
  filter(is.finite(value)) %>%
  mutate(
    metric = factor(
      metric,
      levels = c(
        "Mean temperature\n(°C)",
        "Sen slope\n(°C / 40 yr)",
        "STL trend diff slope\n(10-3 °C / yr2)"
      )
    )
  )

violin_theme <- theme_atomic_base(base_size = 8, grid = TRUE) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "grey92", color = "grey75", linewidth = 0.25),
    strip.text = element_text(size = 7.2, face = "bold", color = col_text),
    axis.text.x = element_text(size = 7, color = col_text),
    axis.text.y = element_text(size = 6.8, color = col_text_soft),
    axis.title.x = element_text(size = 8, color = col_text),
    plot.margin = margin(4, 4, 4, 4)
  )

violin_base <- ggplot(
  cluster_metric_long,
  aes(x = cluster_label, y = value, fill = cluster_label, color = cluster_label)
) +
  geom_violin(
    scale = "width",
    trim = TRUE,
    alpha = 0.20,
    linewidth = 0.35,
    na.rm = TRUE
  ) +
  geom_boxplot(
    width = 0.13,
    outlier.shape = NA,
    alpha = 0.55,
    linewidth = 0.25,
    color = col_line,
    na.rm = TRUE
  ) +
  facet_wrap(~metric, scales = "free_y", nrow = 1) +
  scale_fill_manual(values = cluster_cols, guide = "none") +
  scale_color_manual(values = cluster_cols, guide = "none") +
  labs(x = "Cluster", y = NULL) +
  violin_theme

# --- Combined patchwork layout ---
p_combined <- map_plot / violin_base +
  plot_layout(heights = c(2.2, 1)) +
  plot_annotation(
    title = "STL-trend response clusters of global lakes (K=5)",
    theme = theme(
      plot.title = element_text(size = 13, face = "bold", color = col_text)
    )
  )

save_figure("fig-stl-cluster-map-ridges.png", p_combined, width = 12, height = 9)
p_combined
```

::: {#tbl-stl-cluster-summary .list-table}

- - Cluster
  - Response type
  - n lakes
  - 2020 anomaly
  - Short interpretation
- - C1
  - `r cluster_table_value("C1", "response_type")`
  - `r cluster_table_count("C1")`
  - `r cluster_table_anomaly("C1")`
  - `r cluster_table_value("C1", "short_interpretation")`
- - C2
  - `r cluster_table_value("C2", "response_type")`
  - `r cluster_table_count("C2")`
  - `r cluster_table_anomaly("C2")`
  - `r cluster_table_value("C2", "short_interpretation")`
- - C3
  - `r cluster_table_value("C3", "response_type")`
  - `r cluster_table_count("C3")`
  - `r cluster_table_anomaly("C3")`
  - `r cluster_table_value("C3", "short_interpretation")`
- - C4
  - `r cluster_table_value("C4", "response_type")`
  - `r cluster_table_count("C4")`
  - `r cluster_table_anomaly("C4")`
  - `r cluster_table_value("C4", "short_interpretation")`
- - C5
  - `r cluster_table_value("C5", "response_type")`
  - `r cluster_table_count("C5")`
  - `r cluster_table_anomaly("C5")`
  - `r cluster_table_value("C5", "short_interpretation")`

:::

```{r}
#| label: fig-cluster-density-contours
#| fig-cap: "Two-dimensional density of warming speed and acceleration by STL-trend response cluster. Contour lines show per-cluster density; marginal panels show univariate density by cluster."
#| fig-width: 8
#| fig-height: 7
#| cache: true

density_df <- plot_df %>%
  filter(
    is.finite(warming_speed),
    is.finite(acceleration),
    between(warming_speed, -1, 4),
    between(acceleration, -3, 4)
  )

ggplot(
  density_df,
  aes(x = warming_speed, y = acceleration, colour = cluster_label)
) +
  geom_density_2d(
    breaks = seq(0.1, 0.45, length.out = 5),
    linewidth = 0.6,
    alpha = 0.88,
    na.rm = TRUE
  ) +
  geom_xsidedensity(
    aes(x = warming_speed, fill = cluster_label, group = cluster_label),
    alpha = 0.28,
    linewidth = 0.55
  ) +
  geom_ysidedensity(
    aes(y = acceleration, fill = cluster_label, group = cluster_label),
    alpha = 0.28,
    linewidth = 0.55,
    orientation = "y"
  ) +
  scale_color_manual(
    values = cluster_cols,
    labels = cluster_labels,
    name = NULL
  ) +
  scale_xfill_manual(values = cluster_cols, guide = "none") +
  scale_yfill_manual(values = cluster_cols, guide = "none") +
  scale_xsidey_continuous(expand = c(0, 0)) +
  scale_ysidex_continuous(expand = c(0, 0)) +
  labs(
    x = "Sen slope (°C / 40 yr)",
    y = "Acceleration (10⁻³ °C / yr²)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2.6, alpha = 0.9))) +
  theme_atomic_facet(base_size = 9) +
  theme(
    panel.grid = element_line(color = "grey85", linewidth = 0.3, linetype = "dashed"),
    plot.margin = margin(8, 12, 8, 8),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.key.height = unit(3, "mm"),
    legend.key.width = unit(7, "mm"),
    legend.text = element_text(size = 7.2),
    legend.background = element_blank(),
    ggside.axis.text = element_blank(),
    ggside.axis.ticks = element_blank(),
    ggside.axis.title = element_blank(),
    ggside.panel.grid = element_blank(),
    ggside.panel.scale.x = 1.3 / 5,
    ggside.panel.scale.y = 1.3 / 5
  )
```

<!-- ```{r}
#| label: fig-ts-cluster-map-ridges
#| fig-cap: "Spatial distribution of the eight time-series response clusters. Marginal ridgelines show longitude and latitude density by cluster."
#| fig-width: 12
#| fig-height: 7.4

# --- Shared cluster data ---
cluster_data <- prepare_cluster_data(include_area = TRUE)
trend_sub <- cluster_data$trend_sub
trend_mat <- cluster_data$trend_mat
year_cols <- cluster_data$year_cols
cluster_meta <- trend_sub %>%
  group_by(cluster_label) %>%
  summarise(
    n = n(),
    mean_temp = mean(mean_temp),
    mean_warming = mean(raw_sen_warming_40yr),
    .groups = "drop"
  )
cluster_order <- cluster_data$cluster_order$cluster
new_labels <- paste0("C", 1:8)
names(new_labels) <- as.character(cluster_order)
cluster_cols <- pal_cluster
cluster_labels <- paste0(
  new_labels[as.character(cluster_order)], ": ",
  comma(cluster_meta$n[order(cluster_meta$mean_temp)]), " lakes, ",
  sprintf("%.1f", cluster_meta$mean_temp[order(cluster_meta$mean_temp)]), "°C"
)
names(cluster_labels) <- new_labels[as.character(cluster_order)]
coast <- read_csv(file.path(rdata, "coastline.csv"), show_col_types = FALSE)


# --- Ridge parameters ---
ridge_levels <- new_labels[as.character(cluster_order)]
ridge_height <- 0.92
ridge_gap <- 0.54
smooth_count <- function(x) {
  smoothed <- as.numeric(stats::filter(x, rep(1 / 3, 3), sides = 2))
  ifelse(is.na(smoothed), x, smoothed)
}
ridge_base <- rev(seq_along(ridge_levels)) * ridge_gap
names(ridge_base) <- ridge_levels

lon_min <- -180; lon_max <- 180; lat_min <- -90; lat_max <- 90
bin_width <- 2
lon_centers <- seq(lon_min + bin_width / 2, lon_max - bin_width / 2, by = bin_width)
lat_centers <- seq(lat_min + bin_width / 2, lat_max - bin_width / 2, by = bin_width)

# --- Ridge data ---
lon_counts <- trend_sub %>%
  mutate(lon_center = lon_min + floor((lon - lon_min) / bin_width) * bin_width + bin_width / 2) %>%
  filter(lon_center >= min(lon_centers), lon_center <= max(lon_centers)) %>%
  count(cluster_label, lon_center, name = "n") %>%
  tidyr::complete(cluster_label = ridge_levels, lon_center = lon_centers, fill = list(n = 0)) %>%
  group_by(cluster_label) %>%
  mutate(n_smooth = smooth_count(sqrt(n)), baseline = ridge_base[cluster_label]) %>%
  ungroup()

lon_global_max <- max(lon_counts$n_smooth, na.rm = TRUE)
lon_counts <- lon_counts %>%
  mutate(n_scaled = n_smooth / lon_global_max, y = baseline + n_scaled * ridge_height) %>%
  arrange(desc(baseline), lon_center)

lat_counts <- trend_sub %>%
  mutate(lat_center = lat_min + floor((lat - lat_min) / bin_width) * bin_width + bin_width / 2) %>%
  filter(lat_center >= min(lat_centers), lat_center <= max(lat_centers)) %>%
  count(cluster_label, lat_center, name = "n") %>%
  tidyr::complete(cluster_label = ridge_levels, lat_center = lat_centers, fill = list(n = 0)) %>%
  group_by(cluster_label) %>%
  mutate(n_smooth = smooth_count(sqrt(n)), baseline = ridge_base[cluster_label]) %>%
  ungroup()

lat_global_max <- max(lat_counts$n_smooth, na.rm = TRUE)
lat_counts <- lat_counts %>%
  mutate(n_scaled = n_smooth / lat_global_max, x = baseline + n_scaled * ridge_height) %>%
  arrange(desc(baseline), lat_center)

ridge_base_df <- tibble(cluster_label = ridge_levels, baseline = ridge_base)

# --- Build plots ---
theme_map <- theme_atomic_map()

main_plot <- ggplot() +
  geom_path(data = coast, aes(x = lon, y = lat, group = segment), color = "grey35", linewidth = 0.18) +
  geom_point(data = trend_sub, aes(x = lon, y = lat, colour = cluster_label, size = point_size), alpha = 0.48, stroke = 0.1) +
  scale_color_manual(values = cluster_cols, labels = cluster_labels, name = NULL) +
  scale_size_identity(guide = "none") +
  coord_fixed(ratio = 1, xlim = c(lon_min, lon_max), ylim = c(lat_min, lat_max), expand = FALSE) +
  labs(x = NULL, y = NULL) +
  theme_map +
  theme(
    legend.position = "right",
    axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(),
    plot.margin = margin(0, 0, 0, 0)
  )

ridge_theme <- theme_void(base_size = 8) + theme(plot.margin = margin(0, 0, 0, 0), legend.position = "none")

top_plot <- ggplot(lon_counts, aes(x = lon_center, group = cluster_label, color = cluster_label, fill = cluster_label)) +
  geom_ribbon(aes(ymin = baseline, ymax = y), alpha = 0.24, color = NA) +
  geom_line(aes(y = y), linewidth = 0.52, alpha = 0.96) +
  geom_hline(data = ridge_base_df, aes(yintercept = baseline, color = cluster_label), linewidth = 0.18, alpha = 0.40) +
  scale_color_manual(values = cluster_cols, guide = "none") +
  scale_fill_manual(values = cluster_cols, guide = "none") +
  scale_x_continuous(limits = c(lon_min, lon_max), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0.45, max(ridge_base) + ridge_height + 0.28), expand = c(0, 0)) +
  ridge_theme

right_plot <- ggplot(lat_counts, aes(y = lat_center, group = cluster_label, color = cluster_label, fill = cluster_label)) +
  geom_ribbon(aes(xmin = baseline, xmax = x), orientation = "y", alpha = 0.24, color = NA) +
  geom_line(aes(x = x), linewidth = 0.52, alpha = 0.96, orientation = "y") +
  geom_vline(data = ridge_base_df, aes(xintercept = baseline, color = cluster_label), linewidth = 0.18, alpha = 0.40) +
  scale_color_manual(values = cluster_cols, guide = "none") +
  scale_fill_manual(values = cluster_cols, guide = "none") +
  scale_y_continuous(limits = c(lat_min, lat_max), expand = c(0, 0)) +
  scale_x_continuous(limits = c(0.45, max(ridge_base) + ridge_height + 0.28), expand = c(0, 0)) +
  ridge_theme

# --- Extract legend ---
legend_grob <- get_legend(
  main_plot +
    guides(color = guide_legend(override.aes = list(size = 2.6, alpha = 0.9), ncol = 1)) +
    theme(legend.position = "right")
)
main_plot <- main_plot + theme(legend.position = "none")

# --- Layout ---
figure_width <- 12
figure_height <- 7.4
main_x <- 0.07; main_y <- 0.12
main_w <- 0.68; main_h <- main_w * figure_width / 2 / figure_height
top_h <- 0.20
right_w <- top_h * figure_height / figure_width * 1.15
gap <- 0.025
top_y <- main_y + main_h + gap
right_x <- main_x + main_w + gap
legend_x <- right_x - 0.002
legend_y <- top_y + 0.035

p <- ggdraw() +
  draw_label("Time-series shape clustering of global lakes (K=8, shape+temp)",
             x = main_x, y = 0.98, hjust = 0, vjust = 1, fontface = "bold", size = 14, color = col_text) +
  draw_label("K-means on z-scored annual STL trend + mean temperature. Clusters ordered by mean temperature. Point size = lake area.",
             x = main_x, y = 0.94, hjust = 0, vjust = 1, size = 9.5, color = col_subtitle) +
  draw_plot(top_plot, x = main_x, y = top_y, width = main_w, height = top_h) +
  draw_plot(main_plot, x = main_x, y = main_y, width = main_w, height = main_h) +
  draw_plot(right_plot, x = right_x, y = main_y, width = right_w, height = main_h) +
  draw_plot(legend_grob, x = legend_x, y = legend_y, width = 0.25, height = 0.22)

save_figure("fig-ts-cluster-map-ridges.png", p, width = figure_width, height = figure_height)
p
``` -->

### Cluster descriptions

**C1: Late-accelerating moderate warming.** This cluster contains 20,444 lakes.
The mean STL-trend anomaly rises slowly during the first half of the record and
more strongly after 2000, reaching about +0.86 °C in 2020 relative to the
1981--1990 baseline.

**C2: Early decline / late rebound.** This cluster contains 12,072 lakes. The
mean trajectory is flat to slightly negative through the 1990s, then turns
upward after the early 2000s and reaches about +0.67 °C by 2020.

**C3: Strong sustained warming.** This is the smallest but strongest warming
cluster, with 8,581 lakes. The mean STL-trend anomaly increases steadily across
the full 40-year record and reaches about +1.15 °C in 2020.

**C4: Weak / near-stable warming.** This is the largest cluster, with 29,657
lakes. The mean response remains close to the baseline throughout the record,
with only weak warming and a 2020 anomaly of about +0.15 °C.

**C5: Early warming then plateau.** This cluster contains 21,491 lakes. The
mean trajectory rises through the 1990s and early 2000s, then flattens or
slightly declines, ending near +0.31 °C in 2020.

### Cluster summary

<!-- ```{r}
#| label: fig-cluster-continent-heatmap
#| fig-cap: "Cluster--continent allocation. Each row sums to 100%. Right column = total lakes per cluster. Bottom row = total lakes per continent."
#| fig-width: 4
#| fig-height: 4

# --- Shared cluster data ---
cluster_data <- prepare_cluster_data()
trend_sub <- cluster_data$trend_sub

# --- Cluster × Continent counts ---
raw_counts <- trend_sub %>%
  count(cluster_label, Continent, name = "n")

row_totals <- raw_counts %>% group_by(cluster_label) %>% summarise(total = sum(n), .groups = "drop")
col_totals <- raw_counts %>% group_by(Continent) %>% summarise(total = sum(n), .groups = "drop")
grand_total <- sum(raw_counts$n)

cluster_levels <- paste0("C", 1:8)
cont_levels <- col_totals %>% arrange(desc(total)) %>% pull(Continent)

# --- Main heatmap cells ---
main_cells <- raw_counts %>%
  left_join(row_totals, by = "cluster_label") %>%
  left_join(col_totals, by = "Continent", suffix = c("_row", "_col")) %>%
  mutate(
    frac_row = n / total_row,  # row-normalized: fraction of cluster across continents
    frac_col = n / total_col,  # col-normalized: fraction of clusters within continent
  ) %>%
  mutate(
    cluster_label = factor(cluster_label, levels = c(cluster_levels, "Total")),
    Continent = factor(Continent, levels = c(as.character(cont_levels), "Total")),
    display = comma(n)
  )

# --- Row totals (right column) ---
row_total_cells <- row_totals %>%
  mutate(
    cluster_label = factor(cluster_label, levels = c(cluster_levels, "Total")),
    Continent = factor("Total", levels = c(as.character(cont_levels), "Total")),
    frac = total / grand_total,
    display = comma(total)
  )

# --- Column totals (bottom row) ---
col_total_cells <- col_totals %>%
  mutate(
    cluster_label = factor("Total", levels = c(cluster_levels, "Total")),
    Continent = factor(Continent, levels = c(as.character(cont_levels), "Total")),
    frac = total / grand_total,
    display = comma(total)
  )

# --- Grand total ---
grand_cell <- tibble(
  cluster_label = factor("Total", levels = c(cluster_levels, "Total")),
  Continent = factor("Total", levels = c(as.character(cont_levels), "Total")),
  frac = 1, display = comma(grand_total)
)

# --- Row bar data (horizontal bars per cluster) ---
row_bars <- main_cells %>%
  mutate(
    x_num = as.numeric(Continent),
    y_num = as.numeric(cluster_label)
  ) %>%
  group_by(cluster_label) %>%
  mutate(
    # Bar width fills cell, height = fraction directly (rows sum to 100%)
    bar_left = x_num - 0.5,
    bar_right = x_num - 0.5 + 1.0,
    bar_bottom = y_num - 0.5,
    bar_top = y_num - 0.5 + 0.8 * frac_row
  ) %>%
  ungroup()

# --- Column bar data (vertical bars per continent) ---
col_bars <- main_cells %>%
  mutate(
    x_num = as.numeric(Continent),
    y_num = as.numeric(cluster_label)
  ) %>%
  group_by(Continent) %>%
  mutate(
    # Bar height fills cell, width = fraction directly (columns sum to 100%)
    bar_left = x_num - 0.5,
    bar_right = x_num - 0.5 + 0.8 * frac_col,
    bar_bottom = y_num - 0.5,
    bar_top = y_num + 0.5
  ) %>%
  ungroup()

# --- Combine all cells ---
all_cells <- bind_rows(
  main_cells %>% mutate(cell_type = "main"),
  row_total_cells %>% mutate(cell_type = "row_total"),
  col_total_cells %>% mutate(cell_type = "col_total"),
  grand_cell %>% mutate(cell_type = "grand_total")
) %>%
  mutate(
    text_color = case_when(
      cell_type == "main" ~ col_text,
      TRUE ~ col_text_soft
    )
  )

# --- Plot ---
p <- ggplot() +
  # Main heatmap tiles (white background)
  geom_tile(
    data = all_cells %>% filter(cell_type == "main"),
    aes(x = Continent, y = cluster_label),
    fill = "white", color = "grey85", linewidth = 0.3
  ) +
  # Total cells: grey background
  geom_tile(
    data = all_cells %>% filter(cell_type != "main"),
    aes(x = Continent, y = cluster_label),
    fill = "grey92", color = "white", linewidth = 0.4
  ) +
  # Row bars (horizontal, from bottom edge)
  geom_rect(
    data = row_bars,
    aes(xmin = bar_left, xmax = bar_right, ymin = bar_bottom, ymax = bar_top),
    fill = alpha(pal_direction[["warming"]], 0.5), color = NA
  ) +
  # Column bars (vertical, from left edge)
  geom_rect(
    data = col_bars,
    aes(xmin = bar_left, xmax = bar_right, ymin = bar_bottom, ymax = bar_top),
    fill = alpha(pal_direction[["cooling"]], 0.5), color = NA
  ) +
  # Text labels
  geom_text(
    data = all_cells,
    aes(x = Continent, y = cluster_label, label = display, color = text_color),
    size = 2.8
  ) +
  scale_fill_identity() +
  scale_color_identity() +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(
    x = NULL, y = "Cluster"
  ) +
  theme_atomic_base() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, size = 8),
    axis.text.y = element_text(size = 8),
    axis.title.y = element_text(size = 9, face = "bold"),
    legend.position = "none",
    plot.margin = margin(6, 6, 4, 4)
  )

save_figure("fig-cluster-continent-heatmap.png", p, width = 4, height = 4)
p
``` -->

| Cluster | n | Temp | Warming | Accel | Dominant | Moran's I |
|---|---:|---:|---:|---:|---|---:|
| C1 | 8,636 | 1.8 °C | +0.36 | +0.13 | NA 97% | 0.65 |
| C2 | 7,855 | 2.8 °C | +0.35 | -1.33 | NA 96% | 0.70 |
| C3 | 16,401 | 3.1 °C | +0.71 | +1.98 | EU 74% | 0.65 |
| C4 | 9,667 | 3.4 °C | +0.29 | -1.02 | NA 99% | 0.93 |
| C5 | 21,493 | 4.3 °C | +0.84 | +0.37 | EU 68% | 0.61 |
| C6 | 13,768 | 5.0 °C | +0.46 | -1.02 | NA 97% | 0.80 |
| C7 | 8,649 | 5.0 °C | +0.57 | +1.83 | EU 88% | 0.72 |
| C8 | 5,776 | 21.9 °C | +0.57 | +0.69 | SA 30% | 0.67 |

Temp is mean STL trend in °C. Warming is Sen slope of annual STL trend in
°C / 40 yr. Accel is Sen slope of first-differenced STL trend in 10^-3^
°C / yr^2^. Dominant is the continent with the highest fraction of lakes.

## Regime-shift timing by response type

Each cluster has its own characteristic regime-shift timing. If the clustering
captures real differences in climate response, lakes within the same cluster
should share similar shift-year patterns.

::: {.callout-note}
## Finding 12
Each cluster shows distinct shift-year peaks. The eight clusters differ not
only in spatial pattern but also in physical characteristics. Temperature
level, warming rate, acceleration, and regime-shift timing all vary
systematically across clusters. STL trend shapes and shift timing are
consistent within clusters and distinct between clusters.
:::

<!-- ```{r}
#| label: fig-cluster-characteristics
#| fig-cap: "Cluster characteristics: distributions of temperature, warming, and acceleration, plus mean STL trend by cluster."
#| fig-width: 10
#| fig-height: 7.2

# --- Shared cluster data ---
cluster_data <- prepare_cluster_data(include_accel = TRUE, include_warming = TRUE)
trend_sub <- cluster_data$trend_sub
trend_mat <- cluster_data$trend_mat
year_cols <- cluster_data$year_cols
cluster_meta <- cluster_data$cluster_meta
label_map <- cluster_data$label_map
cluster_cols <- pal_cluster
direction_fills <- pal_direction_fill
direction_cols_line <- pal_direction
all_clusters <- paste0("C", 1:8)
shifts <- read_csv(file.path(rdata, "shifts_L7_p005.csv"), show_col_types = FALSE)

# --- Shift timing data ---
shifts_cluster <- shifts %>%
  inner_join(trend_sub %>% select(lake_id, cluster_label), by = "lake_id") %>%
  mutate(direction = factor(direction, levels = c("warming", "cooling")))

year_range <- 1988:2020
year_dist <- shifts_cluster %>%
  count(cluster_label, year, direction) %>%
  tidyr::complete(cluster_label = paste0("C", 1:8), year = year_range, direction = c("warming", "cooling"), fill = list(n = 0)) %>%
  filter(year >= 1989) %>%
  mutate(
    signed_n = if_else(direction == "warming", n, -n),
    direction = factor(direction, levels = c("warming", "cooling")),
    cluster_label = factor(cluster_label, levels = paste0("C", 1:8))
  )

# ============================================================
# LEFT COLUMN: 4 panels, compact

# --- Peak detection algorithm ---
# For each cluster+direction: smooth with 3-year rolling mean, find local maxima,
# filter by 75th percentile threshold.
detect_peaks <- function(years, counts) {
  n <- length(counts)
  if (n < 5) return(tibble(year = integer(), peak_count = numeric(), is_range = logical()))

  # 3-year rolling mean
  smoothed <- as.numeric(stats::filter(counts, rep(1/3, 3), sides = 2))
  smoothed[is.na(smoothed)] <- counts[is.na(smoothed)]

  # Threshold: 90th percentile of positive smoothed values
  pos_vals <- smoothed[smoothed > 0]
  if (length(pos_vals) < 3) return(tibble(year = integer(), peak_count = numeric(), is_range = logical()))
  threshold <- quantile(pos_vals, 0.75, na.rm = TRUE)

  # All years above threshold
  above_idx <- which(smoothed >= threshold)
  if (length(above_idx) == 0) return(tibble(year = integer(), peak_count = numeric(), is_range = logical()))

  peak_years <- years[above_idx]
  peak_counts <- smoothed[above_idx]

  # Group consecutive years (gap > 2 = new group)
  groups <- cumsum(c(1, diff(peak_years) > 2))
  result <- list()
  for (g in unique(groups)) {
    idx <- which(groups == g)
    yrs <- peak_years[idx]
    cnts <- peak_counts[idx]
    best <- which.max(cnts)
    if (length(yrs) == 1) {
      result[[length(result) + 1]] <- tibble(year = yrs, peak_count = cnts, is_range = FALSE)
    } else {
      result[[length(result) + 1]] <- tibble(
        year = yrs[best], peak_count = cnts[best],
        is_range = TRUE, range_min = min(yrs), range_max = max(yrs)
      )
    }
  }
  bind_rows(result)
}

# Detect peaks for each cluster+direction
peaks <- year_dist %>%
  group_by(cluster_label, direction) %>%
  group_modify(~ detect_peaks(.x$year, .x$n)) %>%
  ungroup() %>%
  mutate(
    label = if_else(is_range,
      paste0(range_min, "-", substr(range_max, 3, 4)),
      as.character(year)
    ),
    y_label = if_else(direction == "warming", peak_count * 1.15, -peak_count * 1.15)
  )

# Peak ranges for shading
peak_ranges <- peaks %>%
  filter(is_range) %>%
  mutate(
    x_min = range_min,
    x_max = range_max
  ) %>%
  select(cluster_label, direction, x_min, x_max)

# ============================================================

make_violin <- function(data, y_col, y_label, title_text) {
  ggplot(data, aes(x = cluster_label, y = .data[[y_col]])) +
    geom_violin(aes(color = cluster_label), fill = NA, scale = "width", linewidth = 0.35) +
    geom_boxplot(width = 0.10, fill = "white", color = "grey30", linewidth = 0.25, outlier.shape = NA) +
    geom_hline(yintercept = 0, color = "grey50", linewidth = 0.2, linetype = "dashed") +
    scale_color_manual(values = cluster_cols, guide = "none") +
    scale_x_discrete(labels = label_map) +
    labs(x = NULL, y = y_label, title = title_text) +
    theme_atomic_base(grid = FALSE) +
    theme(
      axis.text.x = element_text(angle = 35, hjust = 1, size = 5.5),
      axis.text.y = element_text(size = 6),
      axis.title.y = element_text(size = 7, face = "bold"),
      plot.title = element_text(size = 8, face = "bold"),
      plot.margin = margin(1, 2, 1, 1)
    )
}

p_v1 <- make_violin(trend_sub, "mean_temp", "Mean STL trend (\u00b0C)", "Temperature level")
p_v2 <- make_violin(trend_sub, "warming", "Warming rate (\u00b0C/40yr)", "Warming rate")
p_v3 <- make_violin(trend_sub, "accel", expression("Acceleration (10"^-3 * " °C/yr"^2 * ")"), "Acceleration")

# --- STL trend panels ---
cluster_ts <- trend_sub %>%
  select(cluster_label, all_of(year_cols)) %>%
  group_by(cluster_label) %>%
  summarise(
    across(all_of(year_cols), list(mean = ~ mean(.x, na.rm = TRUE), sd = ~ sd(.x, na.rm = TRUE))),
    .groups = "drop"
  ) %>%
  pivot_longer(-cluster_label, names_to = "col_name", values_to = "value") %>%
  mutate(
    stat = if_else(grepl("_mean$", col_name), "mean", "sd"),
    year = as.integer(gsub("trend_|_(mean|sd)$", "", col_name))
  ) %>%
  select(cluster_label, year, stat, value) %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  filter(!is.na(year))

# C1-C7: y-axis from full mean ± SD range
y_range_1_7 <- cluster_ts %>% filter(cluster_label != "C8") %>%
  { range(c(.$mean - .$sd, .$mean + .$sd), na.rm = TRUE) }
y_pad_1_7 <- diff(y_range_1_7) * 0.03
y_limits_1_7 <- c(y_range_1_7[1] - y_pad_1_7, y_range_1_7[2] + y_pad_1_7)

y_range_all <- range(c(cluster_ts$mean - cluster_ts$sd, cluster_ts$mean + cluster_ts$sd), na.rm = TRUE)
y_pad_all <- diff(y_range_all) * 0.05
y_limits_all <- c(y_range_all[1] - y_pad_all, y_range_all[2] + y_pad_all)

trend_panels <- list()

for (i in 1:8) {
  highlight <- all_clusters[i]
  y_lim <- if (i == 8) y_limits_all else y_limits_1_7
  range_sub <- peak_ranges %>% filter(cluster_label == highlight)
  # Only draw individual peak lines for non-range peaks
  peak_sub <- peaks %>% filter(cluster_label == highlight, !is_range)

  p <- ggplot() +
    # Peak range shading (higher alpha)
    {if (nrow(range_sub) > 0) geom_rect(
      data = range_sub,
      aes(xmin = x_min, xmax = x_max, ymin = -Inf, ymax = Inf, fill = direction),
      inherit.aes = FALSE, alpha = 0.18
    )} +
    # Range boundary lines (dashed, thin, at start and end of each range)
    {if (nrow(range_sub) > 0) geom_vline(
      data = range_sub %>% pivot_longer(c(x_min, x_max), values_to = "x_pos"),
      aes(xintercept = x_pos, color = direction),
      inherit.aes = FALSE, linewidth = 0.3, linetype = "dashed", alpha = 0.5
    )} +
    # Grey lines for other clusters
    geom_line(
      data = cluster_ts %>% filter(cluster_label != highlight),
      aes(x = year, y = mean, group = cluster_label),
      color = "grey65", linewidth = 0.25, alpha = 0.5
    ) +
    # SD ribbon for highlighted cluster
    geom_ribbon(
      data = cluster_ts %>% filter(cluster_label == highlight),
      aes(x = year, ymin = mean - sd, ymax = mean + sd),
      fill = cluster_cols[highlight], alpha = 0.2, color = NA
    ) +
    # Mean line for highlighted cluster
    geom_line(
      data = cluster_ts %>% filter(cluster_label == highlight),
      aes(x = year, y = mean),
      color = cluster_cols[highlight], linewidth = 0.6
    ) +
    # Peak year lines (thin, semi-transparent)
    {if (nrow(peak_sub) > 0) geom_vline(
      data = peak_sub,
      aes(xintercept = year, color = direction),
      linewidth = 0.3, linetype = "dashed", alpha = 0.5
    )} +
    # Peak year labels
    {if (nrow(peak_sub) > 0) geom_text(
      data = peak_sub,
      aes(x = year, y = Inf, label = label, color = direction),
      vjust = 1.5, hjust = 0.5, size = 2, fontface = "bold"
    )} +
    scale_color_manual(values = direction_cols_line, guide = "none") +
    scale_fill_manual(values = direction_fills, guide = "none") +
    scale_x_continuous(breaks = seq(1985, 2020, by = 10)) +
    scale_y_continuous(limits = y_lim, breaks = pretty_breaks(3)) +
    labs(title = highlight) +
    theme_void(base_size = 7) +
    theme(
      plot.title = element_text(size = 7, face = "bold", color = cluster_cols[highlight], hjust = 0.5),
      axis.text.y = element_text(size = 5, color = "grey50"),
      axis.ticks.y = element_line(color = "grey50", linewidth = 0.15),
      axis.line.y = element_line(color = "grey50", linewidth = 0.15),
      plot.margin = margin(1, 1, 1, 1)
    )

  trend_panels[[i]] <- p
}

p_trends <- wrap_plots(trend_panels, ncol = 1) +
  plot_annotation(
    subtitle = "C1-C7: shared y-axis. C8: own y-axis.",
    theme = theme(
      plot.subtitle = element_text(size = 6, color = col_subtitle)
    )
  )

# --- Info panel for bottom-left ---
info_text <- paste0(
  "Cluster summary\n\n",
  paste0(
    cluster_meta$cluster_label, ": ",
    comma(cluster_meta$n), " lakes, ",
    sprintf("%.1f", cluster_meta$mt), "°C",
    collapse = "\n"
  )
)
p_info <- ggplot() +
  annotate("text", x = 0.5, y = 0.95, label = "Cluster summary",
           size = 4.5, fontface = "bold", color = col_text, hjust = 0.5, vjust = 1) +
  annotate("text", x = 0.5, y = 0.82, label = info_text,
           size = 3.0, color = col_text_soft, hjust = 0.5, vjust = 1, family = "mono", lineheight = 1.1) +
  xlim(0, 1) + ylim(0, 1) +
  theme_void() +
  theme(plot.margin = margin(4, 4, 4, 4))

p_left <- p_v1 / p_v2 / p_v3 / p_info +
  plot_layout(heights = c(1, 1, 1, 1))

# ============================================================
# RIGHT COLUMN: 8 shift timing charts
# ============================================================

shift_panels <- list()
for (i in 1:8) {
  cl <- all_clusters[i]
  sub <- year_dist %>% filter(cluster_label == cl)
  # Prepare peak annotations: range = horizontal segment, single = vertical line
  peak_sub <- peaks %>% filter(cluster_label == cl)
  range_sub <- peak_ranges %>% filter(cluster_label == cl)

  # Compute y offset for directed peak marks (small fraction of max count)
  max_n <- max(abs(sub$signed_n), na.rm = TRUE)
  y_off <- max_n * 0.08

  range_marks <- NULL
  single_marks <- NULL

  p <- ggplot(sub, aes(x = year, y = signed_n, fill = direction)) +
    # Low-alpha cluster color background
    annotate("rect", xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf,
             fill = cluster_cols[cl], alpha = 0.15) +
    # Peak range shading
    {if (nrow(range_sub) > 0) geom_rect(
      data = range_sub,
      aes(xmin = x_min, xmax = x_max, ymin = -Inf, ymax = Inf, fill = direction),
      inherit.aes = FALSE, alpha = 0.1
    )} +
    geom_col(width = 0.82, alpha = 0.65, linewidth = 0) +
    geom_col(width = 0.82, fill = NA, aes(color = direction), linewidth = 0.2) +
    geom_hline(yintercept = 0, color = "grey30", linewidth = 0.15) +
    scale_fill_manual(values = c(direction_fills, direction_fills), guide = "none") +
    scale_color_manual(values = direction_cols_line, guide = "none") +
    scale_x_continuous(breaks = seq(1990, 2020, by = 10), limits = c(1988.5, 2020.5), expand = c(0, 0)) +
    scale_y_continuous(labels = label_number(big.mark = ",")) +
    labs(title = cl, x = NULL, y = NULL) +
    theme_atomic_base(grid = FALSE) +
    theme(
      plot.title = element_text(size = 7, face = "bold", color = cluster_cols[cl], hjust = 0.5),
      plot.background = element_rect(fill = "white", color = NA),
      axis.text = element_text(size = 4.5),
      axis.text.x = element_text(angle = 0),
      plot.margin = margin(2, 3, 2, 3)
    )

  shift_panels[[i]] <- p
}

p_right <- wrap_plots(shift_panels, ncol = 1) +
  plot_annotation(
    subtitle = "Positive = warming-direction; negative = cooling-direction.",
    theme = theme(
      plot.subtitle = element_text(size = 6, color = col_subtitle)
    )
  )

# ============================================================
# COMPOSE: characteristics panels
# ============================================================
p_characteristics <- p_left | p_trends +
  plot_layout(widths = c(1, 1.15))

save_figure("fig-cluster-characteristics.png", p_characteristics, width = 10, height = 7.2)
p_characteristics
``` -->

<!-- ```{r}
#| label: fig-cluster-shift-timing
#| fig-cap: "Regime-shift timing by cluster. Positive bars are warming shifts; negative bars are cooling shifts."
#| fig-width: 7.2
#| fig-height: 8.8

save_figure("fig-cluster-shift-timing.png", p_right, width = 7.2, height = 8.8)
p_right
``` -->

## Spatial coherence

The eight clusters exhibit different spatial coherence patterns. Some form
tight geographic clusters; others form latitudinal bands spanning multiple
continents.

::: {.callout-note}
## Finding 13
Five clusters (C2, C3, C5, C6, C7) form tight geographic clusters, with more
than 80% of lakes within ±30° of their centroid. Two clusters (C1, C8) form
latitudinal bands spanning the full longitude range at high latitudes. One
cluster (C4, tropical) is globally dispersed but concentrated in the
subtropical latitude band.
:::
