---
title: "Warming Pattern Decomposition"
---

```{r}
#| label: setup-figures-pca
#| include: false
source("../../../shared/R/figure-style.R")
source("../../../shared/R/descriptive-helpers.R")
source("../../../shared/R/02-warming-patterns-helpers.R")
list2env(prepare_pca_data(), envir = environment())
fmt_pct <- function(x, digits = 1) formatC(100 * x, format = "f", digits = digits)
```

Long-term warming and warming-speed change describe trajectory direction and rate, but not when low-frequency departures occur or whether lakes share their timing. This chapter decomposes baseline-centred annual STL trajectories using spatially balanced PCA.

> 长期增温与增温速度变化描述方向和速率，不能保留低频轨迹何时偏离、湖泊间是否同步。本章对基线中心化的年 STL 轨迹做空间平衡 PCA。

## Representation and spatial estimand

Annual reconstructed LSWT is smoothed with the STL trend (nt=99), then each lake trajectory is centred on its 1981--1990 mean. Lakes are averaged within occupied equal-area spherical cells before PCA: 573 cells represent 92,245 lakes. Thus PCA estimates covariance for a typical represented spatial cell, not a typical sampled lake. Individual lake scores are projections onto fixed cell-PCA axes; no lake-equal PCA is refitted.

> 年重建 LSWT 先取 nt=99 STL 趋势，再按各湖 1981–1990 均值中心化。PCA 前在等面积球面格网内汇总：92,245 湖形成 573 个格网。它描述典型空间格网的协变；湖泊分数仅投影到固定格网 PCA 轴，不重新做湖泊等权 PCA。

Detailed construction, LOCO alignment, and projection boundary are specified in [the PCA stability contract](../prose/pca-stability-contract.qmd) and [Spatially balanced PCA](../../../metrics/algorithms.qmd#spatially-balanced-pca).

> 算法、LOCO 对齐与投影边界见 PCA stability contract 和 Spatially balanced PCA 指标页。

```{r}
#| label: fig-pca-variance
#| fig-cap: "Scree plot for equal-area-cell PCA. PC1--PC5 are highlighted; later components add progressively less variance."
#| fig-width: 6
#| fig-height: 3.5
ggplot(scree_data, aes(pc, explained_variance)) +
  geom_col(aes(fill = is_main), width = .7, show.legend = FALSE) +
  geom_line(aes(y = cumulative_explained_variance), colour = col_inchart_line, linewidth = wid_inchart_line) +
  geom_point(aes(y = cumulative_explained_variance), colour = col_inchart_line, size = 1.5) +
  scale_fill_manual(values = c("TRUE" = pal_direction[["warming"]], "FALSE" = col_grid)) +
  scale_x_continuous(breaks = 1:9, labels = paste0("PC", 1:9)) +
  scale_y_continuous(labels = scales::label_percent(), sec.axis = sec_axis(~., name = "Cumulative", labels = scales::label_percent())) +
  labs(x = "Principal component", y = "Variance explained") + theme_base(grid = TRUE)
```

PC1 explains most spatial-cell trajectory variance; PC1--PC5 form the predeclared descriptive set. The 95% threshold is diagnostic, not a retention rule.

> PC1 解释最多的格网轨迹方差；PC1–PC5 是预先确定的描述集合。95% 仅作诊断，不作为保留阈值。

## Temporal covariance modes

LOCO tests reaggregate retained lakes into equal-area cells before refitting. Secondary modes recur, but their rank can exchange and PC4--PC5 can mix; fixed PC labels are temporal descriptors rather than invariant physical entities.

> LOCO 每次先对保留湖泊重新格网汇总再拟合。次级模态会重复出现，但排序可交换、PC4/PC5 可混合；PC 标签仅是时间描述，不是固定物理实体。

```{r}
#| label: fig-pca-loadings
#| fig-cap: "Equal-area-cell PCA loadings for PC1--PC5. Positive lake scores and positive loadings jointly indicate positive baseline-relative temperature contributions."
#| fig-width: 8
#| fig-height: 9
ggplot(loading_plot_data, aes(year, loading)) +
  geom_hline(yintercept = 0, colour = col_inchart_line, linewidth = wid_grid) +
  geom_col(aes(fill = is_positive), width = .8, show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = pal_direction[["warming"]], "FALSE" = pal_direction[["cooling"]])) +
  facet_wrap(~component, ncol = 1, scales = "free_y") +
  scale_x_continuous(breaks = seq(1980, 2020, 5), expand = c(0, 0)) +
  labs(x = "Year", y = "Loading") + theme_base(grid = TRUE) + theme_facet(base_size = 8)
```

Each loading is a year weight and each projected lake score is the strength with which a lake expresses that temporal contrast. PCA signs are arbitrary: interpret score-loading combinations, not isolated signs.

> 载荷是年份权重，投影分数是湖泊表达该时间对比的强度。PCA 正负可整体翻转；应解释“分数×载荷”组合，而非孤立正负。

## Spatial organisation

```{r}
#| label: fig-pca-score-maps
#| fig-cap: "Projected PC1--PC4 scores, averaged in 1° × 1° cells. Colours express arbitrary PCA signs; panels show spatial organisation of temporal contrasts."
#| fig-width: 12
#| fig-height: 7
make_pca_map <- function(pc_col, title) {
  map_data <- prepare_pca_score_map_data(pc_col)
  ggplot() +
    geom_tile(data = map_data$data, aes(lon_cell + .5, lat_cell + .5, fill = score_clamped), width = 1, height = 1) +
    the_geom_coastline() +
    scale_fill_distiller(palette = "RdBu", limits = c(-map_data$limit, map_data$limit), oob = scales::squish, name = "Projected score") +
    coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 85), expand = FALSE) +
    labs(title = title, x = "Longitude", y = "Latitude") + theme_map_grid(base_size = 9) + theme(legend.position = "bottom")
}
maps <- lapply(1:4, \(k) make_pca_map(paste0("pc", k), paste0("PC", k)))
(maps[[1]] + maps[[2]]) / (maps[[3]] + maps[[4]]) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

The maps show continuous spatial organisation, not continental classes. Different temporal contrasts dominate in different neighbouring regions; this is descriptive evidence for spatial heterogeneity, while maps alone do not identify a forcing mechanism.

> 地图展示连续空间组织，而非大洲类别。不同时间对比在不同、且可能相邻的区域强弱不同；这描述空间异质性，但地图本身不识别驱动机制。

```{r}
#| label: fig-pc1-pc2-density
#| fig-cap: "Joint distribution of projected PC1 and PC2 scores. Contours show dense score-space structure; points show tails below the 5th local-density percentile."
#| fig-width: 7
#| fig-height: 5
pc_density <- split_kde_core(pc_scatter_data |> transmute(x = pc1, y = pc2), core_quantile = .05, contour_n = 7)
ggplot(pc_density$points, aes(x, y)) +
  geom_point(data = pc_density$outliers, aes(alpha = density), colour = col_text_soft, size = .35) +
  geom_path(data = pc_density$contours, aes(x, y, group = group), inherit.aes = FALSE, colour = col_inchart_line, linewidth = wid_inchart_line) +
  scale_alpha_continuous(range = c(.08, .55), guide = "none") +
  labs(x = "PC1", y = "PC2") + theme_base(grid = TRUE)
```

@fig-pc1-pc2-density retains continuous score geometry. It neither defines discrete lake types nor supports a regional causal claim.

> @fig-pc1-pc2-density 保留连续分数几何；它不定义离散湖泊类型，也不支持区域因果结论。

## Descriptive trajectory typology

K-means partitions projected PC1--PC5 scores only for readable trajectory summaries. K=5 is a display resolution, not an estimated number of natural response types.

> K-means 仅切分 projected PC1–PC5 分数，以便展示轨迹摘要。K=5 是展示分辨率，不是自然类型数。

```{r}
#| label: fig-cluster-profiles-k5
#| fig-cap: "Median baseline-centred trajectories for descriptive K=5 clustering of projected PC1--PC5 scores. Ribbons show within-cluster interquartile ranges."
#| fig-width: 8
#| fig-height: 4.5
ggplot(cluster_profiles_k5, aes(year, anomaly_median, colour = cluster, fill = cluster)) +
  geom_ribbon(aes(ymin = anomaly_q25, ymax = anomaly_q75), alpha = .14, colour = NA) +
  geom_line(linewidth = wid_inchart_line) +
  scale_colour_manual(values = pal_cluster) + scale_fill_manual(values = pal_cluster) +
  labs(x = NULL, y = "Baseline-centred STL trajectory (°C)", colour = NULL, fill = NULL) +
  theme_base(grid = TRUE) + theme(legend.position = "bottom")
```

```{r}
#| label: fig-clustering-sensitivity
#| fig-cap: "Within-cluster cost across K=4--8 for K-means on projected PC1--PC5 scores. No unique elbow is treated as a natural category count."
#| fig-width: 5.5
#| fig-height: 3.4
ggplot(k_selection_metrics, aes(k, cost_per_lake)) +
  geom_line(colour = col_inchart_line, linewidth = wid_inchart_line) + geom_point(size = 1.8) +
  scale_x_continuous(breaks = 4:8) + labs(x = "K", y = "Within-cluster cost / lake") + theme_base(grid = TRUE)
```

@fig-clustering-sensitivity has no uniquely privileged K. The partition supplies readable profiles; PCA scores and maps remain primary evidence.

> @fig-clustering-sensitivity 没有唯一优选 K。分组提供可读轨迹摘要；主证据仍是 PCA 分数和空间图。

## Association boundary

No predictor model is reported here. A future association analysis must use continuous geographic and climate-background predictors, spatially blocked evaluation, and explicit treatment of spatial autocorrelation; continent dummy variables are not admissible primary explanatory variables. Such a model can identify associations and candidate mechanisms, not causal attribution.

> 本章不报告预测因子模型。未来关联分析必须使用连续地理与气候背景变量、空间分块验证，并显式处理空间自相关；大洲虚拟变量不能作为主要解释变量。模型只能识别关联和候选机制，不能作因果归因。

Requirements are specified in [the association contract](../prose/era5-association-scope.qmd). Chapter 3 remains inactive until inputs, temporal alignment, confounding structure, and held-out evaluation are implemented.

> 具体要求见 association contract。Ch3 在输入、时间对齐、混杂结构和留出评估实现前仍不启动。
