---
title: "Low-Frequency Warming Pattern Decomposition"
---

```{r}
#| include: false
source("../../../shared/R/figure-style.R")
source("../../../shared/R/descriptive-helpers.R")
source("../../../shared/R/02-warming-patterns-helpers.R")
list2env(prepare_pca_data(), envir = environment())
```

This chapter decomposes baseline-centred annual STL trajectories after equal-area spatial aggregation. STL with `nt=99` is PCA-only low-frequency preprocessing; raw annual LSWT remains the primary warming and local-speed representation.

> 本章在等面积空间汇总后分解基线中心化年 STL 轨迹。`nt=99` STL 只用于 PCA 前低频处理；raw 年均 LSWT 仍是主要增温与局部速度表征。

## Spatially balanced trajectory space

PCA is fitted to 573 occupied equal-area cells representing 92,245 lakes. It estimates covariance for a represented spatial cell rather than for a typical sampled lake; lake scores are projections onto fixed cell axes.

> PCA 拟合于 573 个被占据的等面积格网，代表 92,245 湖。它描述代表性空间格网协变，不描述典型抽样湖泊；湖泊分数仅投影到固定格网轴。

```{r}
#| label: fig-pattern-scree
#| fig-cap: "Explained and cumulative variance of spatially balanced PCA."
ggplot(scree_data, aes(pc, explained_variance)) +
  geom_col(aes(fill = is_main), show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = pal_direction[["warming"]], "FALSE" = col_grid)) +
  labs(x = "Principal component", y = "Variance explained") + theme_base(grid = TRUE)
```

PC1 is the robust common low-frequency background. PC2--PC3 form a recurring secondary subspace whose rank can exchange; PC4--PC5 are retained as lower-prominence, partly mixed descriptive modes.

> PC1 是稳健共同低频背景。PC2–PC3 构成可重复、但排序可交换的次级子空间；PC4–PC5 保留为呈现优先级较低、部分混合的描述模态。

## Temporal and spatial modes

Loadings identify the temporal contrast expressed by a score. PCA signs are arbitrary; interpretation always concerns the score-loading combination. PC1 captures the dominant common late-period contrast. PC2 and PC3 describe secondary changes in the timing and persistence of low-frequency warming, rather than two fixed physical mechanisms.

> loading 给出分数所表达的时间对比。PCA 符号可整体翻转，必须解释“分数—loading”组合。PC1 捕捉主要共同后期对比；PC2、PC3 描述低频增温时间和持续性的次级差异，不是两个固定物理机制。

```{r}
#| label: fig-pattern-loadings
#| fig-cap: "Equal-area PCA loadings for PC1--PC5."
#| fig-width: 8
#| fig-height: 9
ggplot(loading_plot_data, aes(year, loading)) +
  geom_hline(yintercept = 0, colour = col_grid, linewidth = wid_grid) +
  geom_col(aes(fill = is_positive), width = .8, show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = pal_direction[["warming"]], "FALSE" = pal_direction[["cooling"]])) +
  facet_wrap(~component, ncol = 1, scales = "free_y") +
  labs(x = "Year", y = "Loading") + theme_base(grid = TRUE) + theme_facet(base_size = 8)
```

```{r}
#| label: fig-pattern-score-maps
#| fig-cap: "Spatial organisation of projected PC scores, averaged in 1° cells."
#| fig-width: 12
#| fig-height: 10
make_pca_map <- function(pc_col, title) {
  map_data <- prepare_pca_score_map_data(pc_col)
  ggplot() +
    geom_tile(data = map_data$data, aes(lon_cell + .5, lat_cell + .5, fill = score_clamped), width = 1, height = 1) +
    the_geom_coastline() +
    scale_fill_distiller(palette = "RdBu", limits = c(-map_data$limit, map_data$limit), oob = scales::squish, name = "Score") +
    coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 85), expand = FALSE) +
    labs(title = title, x = "Longitude", y = "Latitude") + theme_map_grid(base_size = 8)
}
wrap_plots(lapply(1:5, \(k) make_pca_map(paste0("pc", k), paste0("PC", k))), ncol = 2)
```

The score maps show continuous spatial organisation. They do not define continental types: adjacent regions can differ, and distant regions can express similar score combinations. This is the spatial form of heterogeneous warming trajectories.

> 分数地图显示连续空间组织，不定义大洲类型：相邻区域可不同，远距区域也可有相似分数组合。这就是异质增温轨迹的空间形式。

## Reproducibility and result hierarchy

At the reference grid PC1--PC5 explain 84.6% of cell-trajectory variance. PC2 and PC3 recur under leave-one-continent-out refits, but can exchange order; PC4--PC5 recur more weakly and partly mix. We therefore retain PC1 as the common background, PC2--PC3 as the main secondary descriptive subspace, and PC4--PC5 as lower-prominence detail.

> 参考格网中 PC1–PC5 共解释 84.6% 格网轨迹方差。PC2、PC3 在 LOCO 重拟中重复但可交换排序；PC4、PC5 重复更弱且部分混合。因此 PC1 为共同背景，PC2–PC3 为主要次级描述子空间，PC4–PC5 为较低优先级细节。

Detailed loading interpretation, score-pole composites, and LOCO matching are retained in [PCA Stability Contract](../prose/pca-stability-contract.qmd). PCA identifies reproducible covariance structure; it does not identify a forcing mechanism by itself.

> loading 解读、score-pole 复合和 LOCO 匹配详见 PCA Stability Contract。PCA 识别可重复协变结构，本身不识别驱动机制。

Construction, loading interpretation, maps, poles, and LOCO diagnostics remain in [PCA Stability Contract](../prose/pca-stability-contract.qmd) and the earlier detailed [PCA exploration](02-warming-patterns.qmd).

> 构建、loading 解读、地图、极端轨迹与 LOCO 诊断见 PCA Stability Contract 和既有 PCA 详细探索页。
