---
title: "Spatially Heterogeneous Teleconnection Sensitivity"
---

```{r}
#| include: false
source("../../../shared/R/figure-style.R")
source("../../../shared/R/pca-seasonal-teleconnection-helpers.R")
source("../../../shared/R/pca-selected-seasonal-teleconnection-helpers.R")
tele_spatial <- prepare_selected_tele_spatial_fields()
tele_neighbors <- prepare_selected_tele_neighbor_diagnostics(tele_spatial$primary)
tele_composites <- prepare_selected_tele_trajectory_composites(spatial = tele_spatial)
```

This chapter asks where detrended lake-temperature sensitivity to predeclared circulation indices differs, and whether that sensitivity field shares spatial structure with PCA scores. It is an association analysis, not teleconnection attribution.

> 本章考察湖温对预先定义环流指数的去趋势敏感性在哪里不同，以及该敏感性场是否与 PCA 分数共享空间结构。这是关联分析，不是遥相关归因。

## Retained NAO/AO result

The strongest retained family pairs prior-summer NAO/AO with current-summer JJA LSWT. PC1 adds little held-out prediction; PC2 and the PC2--PC3 subspace add stable spatial skill across grids, continent omissions, and decade omissions.

> 最稳的关联族是前年夏季 NAO/AO 与当年夏季 JJA LSWT。PC1 几乎不增加空间留出预测；PC2 与 PC2–PC3 子空间在格网、大陆删除、年代删除中保持稳定增益。

```{r}
#| label: fig-tele-primary-fields
#| fig-cap: "Signed Fisher-z association fields for JJA NAO/AO in year t−1 and JJA LSWT in year t."
ggplot(tele_spatial$primary) +
  geom_rect(aes(xmin = lon_min, xmax = lon_max, ymin = lat_min, ymax = lat_max, fill = tele_fisher_z), colour = NA) +
  the_geom_coastline() + facet_wrap(~tele_index) +
  scale_fill_distiller(palette = "RdBu", direction = -1, name = "Fisher-z") +
  coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 85), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") + theme_map_grid(base_size = 8)
```

The fields are spatially continuous; most visible sign changes are weak-field zero crossings. Positive and negative association poles differ in PC2 position and late local warming-speed evolution. PDO, MAM NAO, and DJF PDO remain bounded regional or coverage-limited comparisons.

> 关联场总体空间连续；大部分可见异号是弱场跨零。正负关联端在 PC2 位置和后期局部增温速度演变上不同。PDO、MAM NAO、DJF PDO 仅保留为区域敏感或覆盖受限比较。

## Spatial continuity and local boundaries

Across 908 east/north adjacent equal-area cell pairs, the NAO and AO fields have neighbour correlations of 0.71 and 0.73. Strong opposite-sign neighbours are rare: at `|Fisher-z| >= 0.15` on both sides, NAO has 5 of 213 pairs and AO has 3 of 193. The field is therefore spatially smooth with weak zero crossings, not a patchwork of locally opposite strong responses.

> 在 908 个东西/南北邻接等面积格网对中，NAO、AO 场的邻接相关分别为 0.71、0.73。两侧均满足 `|Fisher-z| >= 0.15` 的强异号邻接对很少：NAO 为 213 对中的 5 对，AO 为 193 对中的 3 对。因此该场总体连续，局部异号主要是弱场跨零，不是强反向响应的拼贴。

## Trajectory bridge

Positive-association and negative-association poles are display ends of a continuous signed field, not lake types or sensitivity magnitudes. For NAO, their mean PC2 scores are 0.392 and −0.342; for AO they are 0.267 and −0.238. The separation is clearest in late local speed: by 2020 the positive-association pole has annual 10-year speed near 0.069 °C yr⁻¹ for both indices, while the negative pole is near −0.014 (NAO) or −0.018 (AO).

> 正负关联端只是连续有符号场的展示端，不是湖泊类型，也不表示敏感性大小。NAO 两端平均 PC2 为 0.392、−0.342；AO 为 0.267、−0.238。差异在后期局部速度最清楚：2020 年正关联端两个指数的年 10 年速度均约 0.069 °C yr⁻¹，负关联端分别约 −0.014、−0.018。

```{r}
#| label: fig-tele-speed-poles
#| fig-cap: "Trailing 10-year annual and JJA Sen-speed composites for negative and positive NAO/AO association poles."
#| fig-width: 12
#| fig-height: 7
ggplot(tele_composites$speeds, aes(year, speed_mean, colour = sensitivity_pole, fill = sensitivity_pole)) +
  geom_hline(yintercept = 0, colour = col_grid, linewidth = wid_grid) +
  geom_ribbon(aes(ymin = speed_q25, ymax = speed_q75), alpha = .16, colour = NA) +
  geom_line(linewidth = wid_inchart_line) +
  facet_grid(representation ~ tele_index, scales = "free_y") +
  labs(x = "Endpoint year", y = "Trailing 10-year Sen speed (°C yr⁻¹)", colour = NULL, fill = NULL) +
  theme_facet() + theme(legend.position = "bottom")
```

This is a spatial co-location result. It does not show NAO/AO causes PC2, and it does not identify a circulation pathway. Candidate selection followed the broad lag screen; grid, LOCO, and LODO tests assess stability rather than independent discovery.

> 这是空间共定位结果，不表示 NAO/AO 造成 PC2，也不识别环流路径。候选来自宽 lag 筛查之后；格网、LOCO、LODO 检验稳定性，不构成独立发现。

Full validation, neighbour diagnostics, and trajectory composites are in [Selected Seasonal Teleconnection Candidates](../prose/pca-selected-seasonal-teleconnection.qmd).

> 完整验证、邻接诊断和轨迹复合见 Selected Seasonal Teleconnection Candidates。
