# Descriptive overview {#sec-descriptive}

```{r}
#| label: setup-figures
#| include: false

source("../../../../shared/R/figure-style.R")
```

## Sample coverage

GLAST: 92,245 lakes, 1981--2020.

37 lakes have no data. The least complete year still has valid observations for
96.6% of lakes, with 3,145 lakes missing.

```{r}
#| label: fig-data-availability
#| fig-cap: "Annual data availability."
#| fig-width: 8.0
#| fig-height: 4.2

availability_line_col <- "#00a63e"
availability_ribbon_col <- "#b9f8cf"

raw_annual <- read_csv(
  file.path(rdata, "raw_annual_summary.csv"),
  show_col_types = FALSE
)
avail <- raw_annual |>
  mutate(valid_fraction = n_lakes / 92245, missing_n = 92245 - n_lakes)
write_derived_csv(avail, file.path(rdata, "annual_data_availability.csv"))

theme_pub <- theme_atomic_base()

p <- ggplot(avail, aes(x = year, y = valid_fraction)) +
  geom_hline(yintercept = 1, color = "grey70", linewidth = 0.25) +
  geom_ribbon(
    aes(ymax = valid_fraction, ymin = 0),
    fill = availability_ribbon_col, alpha = 0.35
  ) +
  geom_line(color = availability_line_col, linewidth = 0.75) +
  geom_point(color = availability_line_col, size = 1.15) +
  scale_y_continuous(labels = percent) +
  coord_cartesian(ylim = c(0.96, 1.00)) +
  scale_x_continuous(breaks = seq(1981, 2020, by = 5)) +
  labs(
    x = "Year",
    y = "Valid raw annual mean fraction"
  ) +
  theme_pub

save_figure("fig-data-availability.png", p, width = 8.0, height = 4.2)
p
```

## Long-term warming

::: {.callout-note}
## Finding 1
79,178 / 92,208 lakes (85.9%) are warming by Sen's slope of annual mean;
13,030 are cooling.
:::

::: {.callout-note}
## Finding 2
Mean warming from Sen's slope of annual mean is 0.730 ± 0.748 °C / 40 yr.
Median warming is 0.69 °C / 40 yr, with Q1--Q3 of 0.26--1.15 °C / 40 yr.
:::

::: {.callout-note}
## Finding 3
Mean warming from Sen's slope of STL trend is 0.573 ± 0.320 °C / 40 yr.
Median warming is 0.52 °C / 40 yr, with Q1--Q3 of 0.34--0.75 °C / 40 yr.
:::

## Long-term acceleration

::: {.callout-note}
## Finding 4
50,614 / 92,193 lakes (54.9%) show positive acceleration, measured as Sen's
slope of first-differenced STL trend. 41,579 show negative acceleration.

Here "acceleration" means faster and faster warming, slower and slower cooling,
or a transition toward warming.
:::

::: {.callout-note}
## Finding 5
Mean acceleration is 0.294 ± 1.486 × 10^-3^ °C / yr^2^. Median acceleration is
0.20 × 10^-3^ °C / yr^2^, with Q1--Q3 of -0.84--1.31 × 10^-3^ °C / yr^2^.
:::

## Joint warming and acceleration states

| State | n | % |
|---|---:|---:|
| warming + accelerating | 44,857 | 48.66% |
| warming + decelerating | 34,312 | 37.22% |
| cooling + accelerating (slower and slower cooling) | 5,757 | 6.24% |
| cooling + decelerating | 7,267 | 7.88% |

```{r}
#| label: fig-global-state-scatter
#| fig-cap: "Joint warming--acceleration density. Cells show log10 lake counts; marginal curves summarise one-dimensional distributions and reference quartiles."
#| fig-width: 7.8
#| fig-height: 5.9

metrics <- prepare_warming_accel_data()

scatter <- metrics |>
  filter(is.finite(raw_sen_warming_40yr), is.finite(stl_accel_sen_slope_1e3)) |>
  mutate(
    quadrant = case_when(
      raw_sen_warming_40yr > 0 & stl_accel_sen_slope_1e3 > 0
      ~ "Warming + accelerating",
      raw_sen_warming_40yr > 0 & stl_accel_sen_slope_1e3 <= 0
      ~ "Warming + decelerating",
      raw_sen_warming_40yr <= 0 & stl_accel_sen_slope_1e3 > 0
      ~ "Cooling + accelerating",
      TRUE ~ "Cooling + decelerating"
    ),
    quadrant = factor(quadrant, levels = c(
      "Warming + accelerating",
      "Warming + decelerating",
      "Cooling + accelerating",
      "Cooling + decelerating"
    ))
  )

quad <- scatter |>
  count(quadrant, name = "n") |>
  mutate(
    prop = n / sum(n),
    label = paste0(percent(prop, accuracy = 0.1), "\n", comma(n))
  )
write_derived_csv(scatter, file.path(rdata, "global_warming_accel_scatter.csv"))
write_derived_csv(quad, file.path(rdata, "global_warming_accel_quadrants.csv"))

axis_source <- scatter
xlim <- quantile(
  axis_source$raw_sen_warming_40yr, c(0.001, 0.999),
  na.rm = TRUE
)
ylim <- quantile(
  axis_source$stl_accel_sen_slope_1e3, c(0.001, 0.999),
  na.rm = TRUE
)
xlim[1] <- floor(min(xlim[1], -0.05) * 4) / 4
xlim[2] <- ceiling(max(xlim[2], 2.0) * 4) / 4
yabs <- ceiling(max(abs(ylim)) / 1) * 1
ylim <- c(-yabs, yabs)

x_step <- 0.25
tile_width_norm <- x_step / diff(xlim)
y_step <- tile_width_norm * diff(ylim)

plot_df <- scatter |>
  filter(
    raw_sen_warming_40yr >= xlim[1], raw_sen_warming_40yr <= xlim[2],
    stl_accel_sen_slope_1e3 >= ylim[1], stl_accel_sen_slope_1e3 <= ylim[2]
  ) |>
  mutate(
    x_norm = (raw_sen_warming_40yr - xlim[1]) / diff(xlim),
    y_norm = (stl_accel_sen_slope_1e3 - ylim[1]) / diff(ylim),
    x_bin = floor(raw_sen_warming_40yr / x_step),
    y_bin = floor(stl_accel_sen_slope_1e3 / y_step)
  )

tile_df <- plot_df |>
  count(x_bin, y_bin, name = "n") |>
  mutate(
    x = (x_bin * x_step + x_step / 2 - xlim[1]) / diff(xlim),
    y = (y_bin * y_step + y_step / 2 - ylim[1]) / diff(ylim)
  )

label_df <- tibble(
  x = c(0.72, 0.72, 0.14, 0.14),
  y = c(0.86, 0.14, 0.86, 0.14),
  quadrant = factor(levels(scatter$quadrant), levels = levels(scatter$quadrant))
) |> left_join(quad, by = "quadrant")

x_density <- density(
  plot_df$raw_sen_warming_40yr,
  from = xlim[1], to = xlim[2], n = 512, adjust = 1.05
)
y_density <- density(
  plot_df$stl_accel_sen_slope_1e3,
  from = ylim[1], to = ylim[2], n = 512, adjust = 1.05
)
side_density_scale <- 0.58
side_headroom <- 2.25
side_fill_col <- alpha("#bfdbfe", 0.32)
x_density_df <- tibble(
  x = (x_density$x - xlim[1]) / diff(xlim),
  density = x_density$y * side_density_scale
)
y_density_df <- tibble(
  y = (y_density$x - ylim[1]) / diff(ylim),
  density = y_density$y * side_density_scale
)
x_density_bar_width <- median(diff(x_density_df$x), na.rm = TRUE) * 1.25
y_density_bar_width <- median(diff(y_density_df$y), na.rm = TRUE) * 1.25
x_density_top <- max(x_density_df$density, na.rm = TRUE) * side_headroom
y_density_top <- max(y_density_df$density, na.rm = TRUE) * side_headroom

x_stats <- tibble(
  stat = factor(
    c("P25", "Median", "P75", "Mean"),
    levels = c("P25", "Median", "P75", "Mean")
  ),
  value = c(
    quantile(plot_df$raw_sen_warming_40yr, 0.25, na.rm = TRUE),
    median(plot_df$raw_sen_warming_40yr, na.rm = TRUE),
    quantile(plot_df$raw_sen_warming_40yr, 0.75, na.rm = TRUE),
    mean(plot_df$raw_sen_warming_40yr, na.rm = TRUE)
  )
) |>
  mutate(
    value_norm = (value - xlim[1]) / diff(xlim),
    density = approx(
      x_density_df$x, x_density_df$density,
      xout = value_norm, rule = 2
    )$y
  )
y_stats <- tibble(
  stat = factor(
    c("P25", "Median", "P75", "Mean"),
    levels = c("P25", "Median", "P75", "Mean")
  ),
  value = c(
    quantile(plot_df$stl_accel_sen_slope_1e3, 0.25, na.rm = TRUE),
    median(plot_df$stl_accel_sen_slope_1e3, na.rm = TRUE),
    quantile(plot_df$stl_accel_sen_slope_1e3, 0.75, na.rm = TRUE),
    mean(plot_df$stl_accel_sen_slope_1e3, na.rm = TRUE)
  )
) |>
  mutate(
    value_norm = (value - ylim[1]) / diff(ylim),
    density = approx(
      y_density_df$y, y_density_df$density,
      xout = value_norm, rule = 2
    )$y
  )
x_quantile_stats <- x_stats |>
  filter(stat != "Mean") |>
  mutate(reference = "Quartile")
x_mean_stats <- x_stats |>
  filter(stat == "Mean") |>
  mutate(density = x_density_top, reference = "Mean")
y_quantile_stats <- y_stats |>
  filter(stat != "Mean") |>
  mutate(reference = "Quartile")
y_mean_stats <- y_stats |>
  filter(stat == "Mean") |>
  mutate(density = y_density_top, reference = "Mean")

reference_levels <- c("Mean", "Quartile")
reference_line_col <- "#0284c7"
reference_linetypes <- c(
  "Mean" = "solid",
  "Quartile" = "dashed"
)
count_fill_cols <- paletteer_c("grDevices::Turku", n = 256)

x_breaks <- seq(ceiling(xlim[1]), floor(xlim[2]), by = 1)
y_breaks <- seq(ceiling(ylim[1]), floor(ylim[2]), by = 2)
x_breaks <- x_breaks[x_breaks >= xlim[1] & x_breaks <= xlim[2]]
y_breaks <- y_breaks[y_breaks >= ylim[1] & y_breaks <= ylim[2]]
x_labels_pos <- (x_breaks - xlim[1]) / diff(xlim)
y_labels_pos <- (y_breaks - ylim[1]) / diff(ylim)
x0 <- (0 - xlim[1]) / diff(xlim)
y0 <- (0 - ylim[1]) / diff(ylim)
pad <- tile_width_norm / 2

base_theme <- theme_atomic_base(grid = TRUE) + theme(
  ggside.panel.scale = 0.26,
  ggside.panel.border = element_blank(),
  ggside.panel.background = element_blank(),
  ggside.panel.grid = element_blank(),
  ggside.axis.text = element_blank(),
  ggside.axis.ticks = element_blank(),
  ggside.axis.title = element_blank()
)

p <- ggplot(
  plot_df,
  aes(x = x_norm, y = y_norm)
) +
  geom_tile(
    data = tile_df,
    aes(x = x, y = y, fill = n),
    inherit.aes = FALSE,
    width = tile_width_norm,
    height = tile_width_norm,
    color = "white",
    linewidth = 0.04,
    alpha = 0.96
  ) +
  geom_hline(yintercept = y0, color = "grey12", linewidth = 0.44) +
  geom_vline(xintercept = x0, color = "grey12", linewidth = 0.44) +
  geom_text(
    data = label_df,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    size = 3.2,
    fontface = "bold",
    lineheight = 0.92,
    color = "grey12"
  ) +
  geom_xsidecol(
    data = x_density_df,
    aes(x = x, y = density),
    inherit.aes = FALSE,
    width = x_density_bar_width,
    fill = side_fill_col,
    color = NA
  ) +
  geom_ysidecol(
    data = y_density_df,
    aes(x = density, y = y),
    inherit.aes = FALSE,
    width = y_density_bar_width,
    fill = side_fill_col,
    color = NA
  ) +
  geom_xsideline(
    data = x_density_df,
    aes(x = x, y = density),
    inherit.aes = FALSE,
    color = "#0284c7",
    linewidth = 0.42
  ) +
  geom_ysideline(
    data = y_density_df,
    aes(x = density, y = y),
    inherit.aes = FALSE,
    color = "#0284c7",
    linewidth = 0.42
  ) +
  geom_xsidepoint(
    data = tibble(x = 0, y = x_density_top),
    aes(x = x, y = y),
    inherit.aes = FALSE,
    alpha = 0
  ) +
  geom_ysidepoint(
    data = tibble(x = y_density_top, y = 0),
    aes(x = x, y = y),
    inherit.aes = FALSE,
    alpha = 0
  ) +
  geom_xsidesegment(
    data = x_quantile_stats,
    aes(
      x = value_norm,
      xend = value_norm,
      y = 0,
      yend = density,
      linetype = reference
    ),
    inherit.aes = FALSE,
    color = reference_line_col,
    linewidth = 0.42,
    alpha = 0.9
  ) +
  geom_xsidesegment(
    data = x_mean_stats,
    aes(
      x = value_norm,
      xend = value_norm,
      y = 0,
      yend = density,
      linetype = reference
    ),
    inherit.aes = FALSE,
    color = reference_line_col,
    linewidth = 0.62,
    alpha = 0.96
  ) +
  geom_ysidesegment(
    data = y_quantile_stats,
    aes(
      x = 0,
      xend = density,
      y = value_norm,
      yend = value_norm,
      linetype = reference
    ),
    inherit.aes = FALSE,
    color = reference_line_col,
    linewidth = 0.42,
    alpha = 0.9
  ) +
  geom_ysidesegment(
    data = y_mean_stats,
    aes(
      x = 0,
      xend = density,
      y = value_norm,
      yend = value_norm,
      linetype = reference
    ),
    inherit.aes = FALSE,
    color = reference_line_col,
    linewidth = 0.62,
    alpha = 0.96
  ) +
  scale_fill_gradientn(
    colors = count_fill_cols,
    trans = "log10",
    labels = comma,
    name = "Lake count\nlog10 scale"
  ) +
  scale_linetype_manual(
    values = reference_linetypes,
    breaks = reference_levels,
    name = "Marginal\nreference"
  ) +
  scale_x_continuous(
    breaks = x_labels_pos,
    labels = number(x_breaks, accuracy = 0.1),
    limits = c(-pad, 1 + pad),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    breaks = y_labels_pos,
    labels = number(y_breaks, accuracy = 1),
    limits = c(-pad, 1 + pad),
    expand = c(0, 0)
  ) +
  coord_fixed(ratio = 1, clip = "off") +
  labs(
    x = "Raw annual Sen warming, 1981-2020 (°C/40yr)",
    y = expression("Sen acceleration of STL trend diff (10"^-3 * " °C/yr²)")
  ) +
  guides(
    fill = guide_colorbar(
      order = 1, barheight = unit(36, "pt"), barwidth = unit(7, "pt")
    ),
    linetype = guide_legend(order = 2, override.aes = list(linewidth = 0.8, color = reference_line_col))
  ) +
  base_theme +
  theme(
    aspect.ratio = 1,
    legend.position = "right",
    legend.box = "vertical"
  )

save_figure("fig-global-state-scatter.png", p, width = 7.8, height = 5.9)
p
```

```{r}
#| label: fig-grid-median-temp
#| fig-cap: "Median lake temperature in warming--acceleration space. Marginal profiles show binned median temperature with interquartile ribbons and linear fits."
#| fig-width: 9.6
#| fig-height: 7.2

metrics <- prepare_warming_accel_data()
raw <- read_csv("processed/to-monthly-average/C data-monthly-average.csv", show_col_types = FALSE, progress = FALSE)
stopifnot(nrow(metrics) == nrow(raw), all(metrics$lake_id == raw$lake_id))

raw_months <- as.matrix(raw[, 4:ncol(raw)])
mean_temp_C <- rowMeans(raw_months, na.rm = TRUE) - 273.15
mean_temp_C[is.nan(mean_temp_C)] <- NA_real_

axis_source <- metrics %>%
  filter(is.finite(raw_sen_warming_40yr), is.finite(stl_accel_sen_slope_1e3))

plot_df <- metrics %>%
  mutate(mean_temp_C = mean_temp_C) %>%
  filter(is.finite(raw_sen_warming_40yr), is.finite(stl_accel_sen_slope_1e3), is.finite(mean_temp_C))

xlim <- quantile(axis_source$raw_sen_warming_40yr, c(0.001, 0.999), na.rm = TRUE)
ylim <- quantile(axis_source$stl_accel_sen_slope_1e3, c(0.001, 0.999), na.rm = TRUE)
xlim[1] <- floor(min(xlim[1], -0.05) * 4) / 4
xlim[2] <- ceiling(max(xlim[2], 2.0) * 4) / 4
yabs <- ceiling(max(abs(ylim)) / 1) * 1
ylim <- c(-yabs, yabs)

# Build square-looking tiles in normalized coordinates. The x bins are anchored
# at 0, so grid boundaries align with 0, 1, and 2 on the x axis.
x_step <- 0.25
tile_width_norm <- x_step / diff(xlim)
y_step <- tile_width_norm * diff(ylim)

plot_df <- plot_df %>%
  filter(
    raw_sen_warming_40yr >= xlim[1], raw_sen_warming_40yr <= xlim[2],
    stl_accel_sen_slope_1e3 >= ylim[1], stl_accel_sen_slope_1e3 <= ylim[2]
  ) %>%
  mutate(
    x_bin = floor(raw_sen_warming_40yr / x_step),
    y_bin = floor(stl_accel_sen_slope_1e3 / y_step)
  )

tile_df <- plot_df %>%
  group_by(x_bin, y_bin) %>%
  summarise(mean_temp_C = median(mean_temp_C, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(
    x = (x_bin * x_step + x_step / 2 - xlim[1]) / diff(xlim),
    y = (y_bin * y_step + y_step / 2 - ylim[1]) / diff(ylim)
  )

top_profile <- plot_df %>%
  group_by(x_bin) %>%
  summarise(
    temp_q25 = quantile(mean_temp_C, 0.25, na.rm = TRUE),
    temp_med = median(mean_temp_C, na.rm = TRUE),
    temp_q75 = quantile(mean_temp_C, 0.75, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  filter(n >= 20) %>%
  mutate(
    x_raw = x_bin * x_step + x_step / 2,
    x = (x_raw - xlim[1]) / diff(xlim)
  )

right_profile <- plot_df %>%
  group_by(y_bin) %>%
  summarise(
    temp_q25 = quantile(mean_temp_C, 0.25, na.rm = TRUE),
    temp_med = median(mean_temp_C, na.rm = TRUE),
    temp_q75 = quantile(mean_temp_C, 0.75, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  filter(n >= 20) %>%
  mutate(
    y_raw = y_bin * y_step + y_step / 2,
    y = (y_raw - ylim[1]) / diff(ylim)
  )

temp_profile_lim <- range(c(top_profile$temp_q25, top_profile$temp_q75, right_profile$temp_q25, right_profile$temp_q75), na.rm = TRUE)
temp_profile_lim <- c(floor(temp_profile_lim[1] / 5) * 5, ceiling(temp_profile_lim[2] / 5) * 5)

top_fit <- lm(temp_med ~ x_raw, data = top_profile)
right_fit <- lm(temp_med ~ y_raw, data = right_profile)
top_fit_summary <- summary(top_fit)
right_fit_summary <- summary(right_fit)
top_profile <- top_profile %>%
  mutate(temp_fit = predict(top_fit, newdata = top_profile))
right_profile <- right_profile %>%
  mutate(temp_fit = predict(right_fit, newdata = right_profile))

top_fit_label <- sprintf(
  "fit: T = %.1f %+.1f W\nR² = %.2f",
  coef(top_fit)[[1]], coef(top_fit)[[2]], top_fit_summary$r.squared
)
right_fit_label <- sprintf(
  "fit: T = %.1f %+.2f A\nR² = %.2f",
  coef(right_fit)[[1]], coef(right_fit)[[2]], right_fit_summary$r.squared
)

fit_stats <- tibble(
  relation = c("mean_temp_vs_warming", "mean_temp_vs_acceleration"),
  response = "binned median mean_temp_C",
  predictor = c("raw_sen_warming_40yr", "stl_accel_sen_slope_1e3"),
  intercept = c(coef(top_fit)[[1]], coef(right_fit)[[1]]),
  slope = c(coef(top_fit)[[2]], coef(right_fit)[[2]]),
  r_squared = c(top_fit_summary$r.squared, right_fit_summary$r.squared),
  adj_r_squared = c(top_fit_summary$adj.r.squared, right_fit_summary$adj.r.squared),
  n_bins = c(nrow(top_profile), nrow(right_profile))
)
write_derived_csv(fit_stats, file.path(rdata, "fig03_temperature_profile_linear_fits.csv"))

x_breaks <- seq(ceiling(xlim[1]), floor(xlim[2]), by = 1)
y_breaks <- seq(ceiling(ylim[1]), floor(ylim[2]), by = 2)
x_breaks <- x_breaks[x_breaks >= xlim[1] & x_breaks <= xlim[2]]
y_breaks <- y_breaks[y_breaks >= ylim[1] & y_breaks <= ylim[2]]
x_labels_pos <- (x_breaks - xlim[1]) / diff(xlim)
y_labels_pos <- (y_breaks - ylim[1]) / diff(ylim)

x0 <- (0 - xlim[1]) / diff(xlim)
y0 <- (0 - ylim[1]) / diff(ylim)

# Expanded limits prevent boundary cells from being cut by the panel.
pad <- tile_width_norm / 2

theme_pub <- theme_atomic_base(grid = TRUE)
profile_line_col <- "#075985"
profile_fit_col <- "#b45309"
profile_fill_col <- alpha("#bae6fd", 0.42)
profile_grid_col <- alpha(col_grid, 0.55)

main_plot <- ggplot(tile_df, aes(x = x, y = y, fill = mean_temp_C)) +
  geom_tile(width = tile_width_norm, height = tile_width_norm, color = "white", linewidth = 0.08) +
  geom_hline(yintercept = y0, color = "grey15", linewidth = 0.38) +
  geom_vline(xintercept = x0, color = "grey15", linewidth = 0.38) +
  scale_fill_gradientn(
    # colors = c("#08306b", "#2171b5", "#41b6c4", "#ffffbf", "#fdae61", "#d73027", "#7f0000"),
    colors = paletteer_c("grDevices::Temps", n = 256),
    name = "Median temp\n(°C)"
  ) +
  scale_x_continuous(breaks = x_labels_pos, labels = number(x_breaks, accuracy = 0.1), limits = c(-pad, 1 + pad), expand = c(0, 0)) +
  scale_y_continuous(breaks = y_labels_pos, labels = number(y_breaks, accuracy = 1), limits = c(-pad, 1 + pad), expand = c(0, 0)) +
  coord_fixed(ratio = 1, clip = "off") +
  labs(
    x = "Raw annual Sen warming, 1981-2020 (°C/40yr)",
    y = expression("Sen acceleration of STL trend diff (10"^-3 * " °C/yr²)")
  ) +
  theme_pub +
  theme(
    legend.position = "right",
    aspect.ratio = 1
  )

top_plot <- ggplot(top_profile, aes(x = x)) +
  geom_ribbon(aes(ymin = temp_q25, ymax = temp_q75), fill = profile_fill_col, color = NA) +
  geom_line(aes(y = temp_med), color = profile_line_col, linewidth = 0.62) +
  geom_line(aes(y = temp_fit), color = profile_fit_col, linewidth = 0.58, linetype = "longdash") +
  annotate(
    "text",
    x = 0.98,
    y = temp_profile_lim[2] - diff(temp_profile_lim) * 0.08,
    label = top_fit_label,
    hjust = 1,
    vjust = 1,
    size = 2.7,
    lineheight = 0.92,
    color = profile_fit_col
  ) +
  scale_x_continuous(breaks = x_labels_pos, labels = NULL, limits = c(-pad, 1 + pad), expand = c(0, 0)) +
  scale_y_continuous(limits = temp_profile_lim, expand = expansion(mult = c(0.04, 0.08))) +
  labs(y = "Mean\ntemp (°C)", x = NULL) +
  theme_pub +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(size = 8, color = col_text_soft),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(color = profile_grid_col, linewidth = 0.25),
    panel.border = element_blank(),
    axis.line.y.left = element_line(color = col_axis_text, linewidth = 0.35),
    axis.line.x = element_blank(),
    plot.margin = margin(4, 4, 1, 4),
    legend.position = "none"
  )

right_plot <- ggplot(right_profile, aes(y = y)) +
  geom_ribbon(aes(xmin = temp_q25, xmax = temp_q75), orientation = "y", fill = profile_fill_col, color = NA) +
  geom_line(aes(x = temp_med), color = profile_line_col, linewidth = 0.62, orientation = "y") +
  geom_line(aes(x = temp_fit), color = profile_fit_col, linewidth = 0.58, linetype = "longdash", orientation = "y") +
  annotate(
    "text",
    x = temp_profile_lim[2] - diff(temp_profile_lim) * 0.04,
    y = 0.98,
    label = right_fit_label,
    hjust = 1,
    vjust = 1,
    size = 2.7,
    lineheight = 0.92,
    color = profile_fit_col
  ) +
  scale_x_continuous(limits = temp_profile_lim, expand = expansion(mult = c(0.04, 0.08))) +
  scale_y_continuous(breaks = y_labels_pos, labels = NULL, limits = c(-pad, 1 + pad), expand = c(0, 0)) +
  labs(x = "Mean\ntemp (°C)", y = NULL) +
  theme_pub +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.x = element_text(size = 8, color = col_text_soft),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = profile_grid_col, linewidth = 0.25),
    panel.border = element_blank(),
    axis.line.x.bottom = element_line(color = col_axis_text, linewidth = 0.35),
    axis.line.y = element_blank(),
    plot.margin = margin(4, 4, 4, 1),
    legend.position = "none"
  )

figure_width <- 9.6
figure_height <- 7.2
main_x <- 0.08
main_y <- 0.09
main_h <- 0.62
main_w <- main_h * figure_height / figure_width
profile_h <- 0.18
profile_w <- profile_h * figure_height / figure_width * 1.10
gap <- 0.025
right_x <- main_x + main_w + gap
top_y <- main_y + main_h + gap
legend_x <- right_x + profile_w + 0.035

legend_grob <- get_legend(main_plot + theme(legend.position = "right"))
main_plot <- main_plot + theme(legend.position = "none")

p <- ggdraw() +
  draw_label(
    "Grid-cell median 40-year mean temperature",
    x = main_x,
    y = 0.965,
    hjust = 0,
    vjust = 1,
    fontface = "bold",
    size = 13,
    color = col_text
  ) +
  draw_plot(top_plot, x = main_x, y = top_y, width = main_w, height = profile_h) +
  draw_plot(main_plot, x = main_x, y = main_y, width = main_w, height = main_h) +
  draw_plot(right_plot, x = right_x, y = main_y, width = profile_w, height = main_h) +
  draw_plot(legend_grob, x = legend_x, y = main_y + 0.19, width = 0.18, height = 0.30)

save_figure("fig-grid-median-temp.png", p, width = figure_width, height = figure_height)
p
```

## Spatial pattern

```{r}
#| label: fig-map-hex-warming-accel
#| fig-cap: "Hex-binned maps of lake warming and acceleration. Each hexagon summarises lakes within an approximately 5° spatial bin."
#| fig-width: 12
#| fig-height: 8.4

map_cols <- pal_diverging

metrics <- prepare_warming_accel_data()
coast <- read_csv(file.path(rdata, "coastline.csv"), show_col_types = FALSE)

# Flat-top regular hexagons in lon/lat plane.
# Hex height is approximately 5 degrees.
hex_height_deg <- 5
s <- hex_height_deg / sqrt(3) # circumradius / side length in degrees
hex_width_deg <- 2 * s
center_dx <- 1.5 * s
center_dy <- sqrt(3) * s

lon_min <- -180
lon_max <- 180
lat_min <- -90
lat_max <- 90

# Convert lon/lat to flat-top axial hex coordinate and round to nearest hex.
hex_round <- function(q, r) {
  x <- q
  z <- r
  y <- -x - z
  rx <- round(x)
  ry <- round(y)
  rz <- round(z)
  x_diff <- abs(rx - x)
  y_diff <- abs(ry - y)
  z_diff <- abs(rz - z)
  choose_x <- x_diff > y_diff & x_diff > z_diff
  choose_y <- !choose_x & y_diff > z_diff
  rx2 <- ifelse(choose_x, -ry - rz, rx)
  ry2 <- ifelse(choose_y, -rx2 - rz, ry)
  rz2 <- ifelse(!choose_x & !choose_y, -rx2 - ry2, rz)
  list(q = rx2, r = rz2)
}

pts <- metrics %>%
  filter(
    is.finite(raw_sen_warming_40yr), is.finite(stl_accel_sen_slope_1e3),
    is.finite(lon), is.finite(lat)
  ) %>%
  mutate(
    # Shift origin to lon_min/lat_min for stable positive coordinates.
    x = lon - lon_min,
    y = lat - lat_min,
    q_float = (2 / 3 * x) / s,
    r_float = (-1 / 3 * x + sqrt(3) / 3 * y) / s
  )

rounded <- hex_round(pts$q_float, pts$r_float)
pts$q_hex <- rounded$q
pts$r_hex <- rounded$r

hex_summary <- pts %>%
  group_by(q_hex, r_hex) %>%
  summarise(
    n = n(),
    warming = mean(raw_sen_warming_40yr, na.rm = TRUE),
    accel = mean(stl_accel_sen_slope_1e3, na.rm = TRUE),
    lon_c = lon_min + s * 3 / 2 * first(q_hex),
    lat_c = lat_min + s * sqrt(3) * (first(r_hex) + first(q_hex) / 2),
    .groups = "drop"
  ) %>%
  filter(
    lon_c >= lon_min - hex_width_deg, lon_c <= lon_max + hex_width_deg,
    lat_c >= lat_min - hex_height_deg, lat_c <= lat_max + hex_height_deg
  )

make_hex <- function(cx, cy, s, id) {
  # flat-top hex vertices: 0, 60, ..., 300 degrees
  theta <- pi / 180 * seq(0, 300, by = 60)
  tibble(
    id = id,
    lon = cx + s * cos(theta),
    lat = cy + s * sin(theta)
  )
}

hex_poly <- bind_rows(lapply(seq_len(nrow(hex_summary)), function(i) {
  make_hex(hex_summary$lon_c[i], hex_summary$lat_c[i], s, i)
})) %>%
  left_join(hex_summary %>% mutate(id = row_number()), by = "id")

warm_lim <- quantile(abs(hex_summary$warming), 0.98, na.rm = TRUE)
warm_lim <- max(warm_lim, 0.5)
acc_lim <- quantile(abs(hex_summary$accel), 0.98, na.rm = TRUE)
acc_lim <- max(acc_lim, 0.2)

theme_map <- theme_atomic_map()

map_base_layers <- list(
  geom_path(
    data = coast,
    aes(x = lon, y = lat, group = segment),
    color = "grey25",
    linewidth = 0.18
  ),
  coord_fixed(
    ratio = 1,
    xlim = c(lon_min, lon_max),
    ylim = c(lat_min, lat_max),
    expand = FALSE
  ),
  labs(x = "Longitude", y = "Latitude"),
  theme_map,
  theme(
    legend.position = "none",
    plot.title = element_blank(),
    plot.subtitle = element_blank(),
    plot.margin = margin(0, 0, 0, 0)
  )
)

hex_layer <- function(fill_var) {
  geom_polygon(
    data = hex_poly,
    aes(x = lon, y = lat, group = id, fill = {{ fill_var }}),
    color = "white",
    linewidth = 0.12
  )
}

p_warming <- ggplot() +
  hex_layer(warming) +
  map_base_layers +
  scale_fill_gradient2(
    low = map_cols[["low"]],
    mid = map_cols[["mid"]],
    high = map_cols[["high"]],
    midpoint = 0,
    limits = c(-warm_lim, warm_lim),
    oob = squish
  ) +
  annotate(
    "label",
    x = -174,
    y = 82,
    label = "A  Mean warming",
    hjust = 0,
    vjust = 1,
    fill = alpha("white", 0.78),
    color = col_text,
    fontface = "bold",
    size = 3.2
  )

p_accel <- ggplot() +
  hex_layer(accel) +
  map_base_layers +
  scale_fill_gradient2(
    low = map_cols[["low"]],
    mid = map_cols[["mid"]],
    high = map_cols[["high"]],
    midpoint = 0,
    limits = c(-acc_lim, acc_lim),
    oob = squish
  ) +
  annotate(
    "label",
    x = -174,
    y = 82,
    label = "B  Mean acceleration",
    hjust = 0,
    vjust = 1,
    fill = alpha("white", 0.78),
    color = col_text,
    fontface = "bold",
    size = 3.2
  )

legend_breaks <- c(-1, -0.5, 0, 0.5, 1)
legend_df <- tibble(
  y = seq(-1, 1, length.out = 360),
  x = 1,
  value = y
)
legend_plot <- ggplot(legend_df, aes(x = x, y = y, fill = value)) +
  geom_tile(width = 0.22, height = 2 / 360) +
  scale_fill_gradient2(
    low = map_cols[["low"]],
    mid = map_cols[["mid"]],
    high = map_cols[["high"]],
    midpoint = 0,
    limits = c(-1, 1),
    guide = "none"
  ) +
  scale_y_continuous(
    breaks = legend_breaks,
    labels = number(legend_breaks * warm_lim, accuracy = 0.1),
    limits = c(-1.01, 1.01),
    sec.axis = dup_axis(
      breaks = legend_breaks,
      labels = number(legend_breaks * acc_lim, accuracy = 0.1),
      name = expression("Mean acceleration (10"^-3 * " °C/yr²)")
    ),
    name = "Mean warming (°C/40yr)"
  ) +
  scale_x_continuous(limits = c(0.82, 1.18), expand = c(0, 0)) +
  theme_void(base_size = 9) +
  theme(
    axis.title.y.left = element_text(angle = 90, color = col_text, size = 8, margin = margin(r = 5)),
    axis.title.y.right = element_text(angle = 90, color = col_text, size = 8, margin = margin(l = 5)),
    axis.text.y.left = element_text(color = col_text, size = 8),
    axis.text.y.right = element_text(color = col_text, size = 8),
    axis.ticks.y.left = element_line(color = col_axis_text, linewidth = 0.25),
    axis.ticks.y.right = element_line(color = col_axis_text, linewidth = 0.25),
    axis.line.y.left = element_line(color = col_axis_text, linewidth = 0.3),
    axis.line.y.right = element_line(color = col_axis_text, linewidth = 0.3),
    plot.margin = margin(4, 4, 4, 4)
  )

figure_width <- 12
figure_height <- 8.4
map_x <- 0.055
map_w <- 0.62
map_h <- map_w * figure_width / 2 / figure_height
map_gap <- 0.012
bottom_y <- 0.065
top_y <- bottom_y + map_h + map_gap
legend_x <- map_x + map_w + 0.055
legend_y <- bottom_y + 0.08
legend_h <- map_h * 1.55

p <- ggdraw() +
  draw_label(
    "Spatial fields of lake warming and acceleration",
    x = map_x,
    y = 1,
    hjust = 0,
    vjust = 1,
    fontface = "bold",
    size = 14,
    color = col_text
  ) +
  draw_plot(p_warming, x = map_x, y = top_y, width = map_w, height = map_h) +
  draw_plot(p_accel, x = map_x, y = bottom_y, width = map_w, height = map_h) +
  draw_plot(legend_plot, x = legend_x, y = legend_y, width = 0.24, height = legend_h)

save_figure("fig-map-hex-warming-accel.png", p, width = figure_width, height = figure_height)

write_derived_csv(hex_summary, file.path(rdata, "hex_world_warming_accel_5deg.csv"))
p
```

```{r}
#| label: fig-map-state-points-ridgelines
#| fig-cap: "Geographic distribution of the four warming--acceleration states, with marginal longitude and latitude ridgelines and global state proportions."
#| fig-width: 12
#| fig-height: 7.4

state_levels <- c(
  "warming + accelerating",
  "warming + decelerating",
  "cooling + accelerating",
  "cooling + decelerating"
)
state_cols <- pal_state
state_labels <- c(
  "warming + accelerating" = "Warming + accelerating",
  "warming + decelerating" = "Warming + decelerating",
  "cooling + accelerating" = "Cooling + accelerating",
  "cooling + decelerating" = "Cooling + decelerating"
)

metrics <- prepare_warming_accel_data()
attr <- read_csv("processed/lake-attributes/data-lake-attributes.csv", show_col_types = FALSE)
coast <- read_csv(file.path(rdata, "coastline.csv"), show_col_types = FALSE)
stopifnot(nrow(metrics) == nrow(attr), all(metrics$lake_id == attr$lake_id))

area_km2 <- attr$Lake_area / 1e6
area_clip <- quantile(area_km2, c(0.05, 0.95), na.rm = TRUE)

plot_df <- metrics %>%
  mutate(
    accel = stl_accel_sen_slope_1e3,
    area_km2 = area_km2,
    area_km2_clipped = pmin(pmax(area_km2, area_clip[[1]]), area_clip[[2]]),
    point_area = rescale(log10(area_km2_clipped), to = c(0.1, 2), from = log10(area_clip)),
    point_size = sqrt(point_area)
  ) %>%
  filter(
    is.finite(raw_sen_warming_40yr),
    is.finite(accel),
    is.finite(lon),
    is.finite(lat),
    is.finite(point_size)
  ) %>%
  mutate(
    state = case_when(
      raw_sen_warming_40yr > 0 & accel > 0 ~ state_levels[1],
      raw_sen_warming_40yr > 0 & accel <= 0 ~ state_levels[2],
      raw_sen_warming_40yr <= 0 & accel > 0 ~ state_levels[3],
      raw_sen_warming_40yr <= 0 & accel <= 0 ~ state_levels[4],
      TRUE ~ NA_character_
    ),
    state = factor(state, levels = state_levels)
  ) %>%
  filter(!is.na(state))

lon_min <- -180
lon_max <- 180
lat_min <- -90
lat_max <- 90
bin_width <- 2
ridge_height <- 0.92
ridge_gap <- 0.54
smooth_count <- function(x) {
  smoothed <- as.numeric(stats::filter(x, rep(1 / 3, 3), sides = 2))
  ifelse(is.na(smoothed), x, smoothed)
}
ridge_levels_top_to_bottom <- c(
  "warming + accelerating",
  "warming + decelerating",
  "cooling + accelerating",
  "cooling + decelerating"
)
ridge_base <- rev(seq_along(ridge_levels_top_to_bottom)) * ridge_gap
names(ridge_base) <- ridge_levels_top_to_bottom

lon_centers <- seq(lon_min + bin_width / 2, lon_max - bin_width / 2, by = bin_width)
lat_centers <- seq(lat_min + bin_width / 2, lat_max - bin_width / 2, by = bin_width)

lon_counts <- plot_df %>%
  mutate(lon_bin = floor((lon - lon_min) / bin_width), lon_center = lon_min + lon_bin * bin_width + bin_width / 2) %>%
  filter(lon_center >= min(lon_centers), lon_center <= max(lon_centers)) %>%
  count(state, lon_center, name = "n") %>%
  complete(state = factor(state_levels, levels = state_levels), lon_center = lon_centers, fill = list(n = 0)) %>%
  group_by(state) %>%
  mutate(
    n_sqrt = sqrt(n),
    n_smooth = smooth_count(n_sqrt),
    n_scaled = n_smooth,
    baseline = ridge_base[as.character(state)],
    y = baseline + n_scaled * ridge_height
  ) %>%
  ungroup()

lat_counts <- plot_df %>%
  mutate(lat_bin = floor((lat - lat_min) / bin_width), lat_center = lat_min + lat_bin * bin_width + bin_width / 2) %>%
  filter(lat_center >= min(lat_centers), lat_center <= max(lat_centers)) %>%
  count(state, lat_center, name = "n") %>%
  complete(state = factor(state_levels, levels = state_levels), lat_center = lat_centers, fill = list(n = 0)) %>%
  group_by(state) %>%
  mutate(
    n_sqrt = sqrt(n),
    n_smooth = smooth_count(n_sqrt),
    n_scaled = n_smooth,
    baseline = ridge_base[as.character(state)],
    x = baseline + n_scaled * ridge_height
  ) %>%
  ungroup()

lon_global_max <- max(lon_counts$n_smooth, na.rm = TRUE)
lat_global_max <- max(lat_counts$n_smooth, na.rm = TRUE)
lon_counts <- lon_counts %>%
  mutate(
    n_scaled = n_smooth / lon_global_max,
    y = baseline + n_scaled * ridge_height,
    ridge_order = match(as.character(state), ridge_levels_top_to_bottom)
  ) %>%
  arrange(desc(baseline), lon_center)
lat_counts <- lat_counts %>%
  mutate(
    n_scaled = n_smooth / lat_global_max,
    x = baseline + n_scaled * ridge_height,
    ridge_order = match(as.character(state), ridge_levels_top_to_bottom)
  ) %>%
  arrange(desc(baseline), lat_center)

state_summary <- plot_df %>%
  count(state, name = "n") %>%
  mutate(prop = n / sum(n)) %>%
  arrange(factor(state, levels = state_levels))
write_derived_csv(state_summary, file.path(rdata, "fig05_world_state_counts.csv"))
write_derived_csv(lon_counts, file.path(rdata, "fig05_world_state_longitude_ridges.csv"))
write_derived_csv(lat_counts, file.path(rdata, "fig05_world_state_latitude_ridges.csv"))

theme_map <- theme_atomic_map()

main_plot <- ggplot() +
  geom_path(
    data = coast,
    aes(x = lon, y = lat, group = segment),
    color = "grey35",
    linewidth = 0.18
  ) +
  geom_point(
    data = plot_df,
    aes(x = lon, y = lat, colour = state, size = point_size),
    alpha = 0.3,
    stroke = 0.1,
  ) +
  scale_color_manual(values = state_cols, labels = state_labels, name = NULL) +
  scale_size_identity(guide = "none") +
  coord_fixed(ratio = 1, xlim = c(lon_min, lon_max), ylim = c(lat_min, lat_max), expand = FALSE) +
  labs(x = NULL, y = NULL) +
  theme_map +
  theme(
    legend.position = "right",
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = margin(0, 0, 0, 0)
  )

ridge_theme <- theme_void(base_size = 8) +
  theme(
    plot.margin = margin(0, 0, 0, 0),
    legend.position = "none"
  )

top_plot <- ggplot(lon_counts, aes(x = lon_center, group = state, color = state, fill = state)) +
  geom_ribbon(aes(ymin = baseline, ymax = y), alpha = 0.24, color = NA) +
  geom_line(aes(y = y), linewidth = 0.52, alpha = 0.96) +
  geom_hline(
    data = tibble(state = factor(ridge_levels_top_to_bottom, levels = state_levels), baseline = ridge_base),
    aes(yintercept = baseline, color = state),
    linewidth = 0.18,
    alpha = 0.40
  ) +
  scale_color_manual(values = state_cols, guide = "none") +
  scale_fill_manual(values = state_cols, guide = "none") +
  scale_x_continuous(limits = c(lon_min, lon_max), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0.45, max(ridge_base) + ridge_height + 0.28), expand = c(0, 0)) +
  ridge_theme

right_plot <- ggplot(lat_counts, aes(y = lat_center, group = state, color = state, fill = state)) +
  geom_ribbon(aes(xmin = baseline, xmax = x), orientation = "y", alpha = 0.24, color = NA) +
  geom_line(aes(x = x), linewidth = 0.52, alpha = 0.96, orientation = "y") +
  geom_vline(
    data = tibble(state = factor(ridge_levels_top_to_bottom, levels = state_levels), baseline = ridge_base),
    aes(xintercept = baseline, color = state),
    linewidth = 0.18,
    alpha = 0.40
  ) +
  scale_color_manual(values = state_cols, guide = "none") +
  scale_fill_manual(values = state_cols, guide = "none") +
  scale_y_continuous(limits = c(lat_min, lat_max), expand = c(0, 0)) +
  scale_x_continuous(limits = c(0.45, max(ridge_base) + ridge_height + 0.28), expand = c(0, 0)) +
  ridge_theme

main_plot <- main_plot + theme(legend.position = "none")

make_annular_sector <- function(theta0, theta1, r0, r1, id, n = 48) {
  theta_outer <- seq(theta0, theta1, length.out = n)
  theta_inner <- seq(theta1, theta0, length.out = n)
  tibble(
    id = id,
    x = c(r1 * cos(theta_outer), r0 * cos(theta_inner)),
    y = c(r1 * sin(theta_outer), r0 * sin(theta_inner))
  )
}

quarter_bar_width <- pi / 8 # 1/16 circle = 22.5°
inner_radius <- 0.34
outer_radius_max <- 1.02
bar_gap <- pi / 180 * 1.8
radial_df <- state_summary %>%
  mutate(
    id = row_number(),
    theta0 = (id - 1) * quarter_bar_width + bar_gap,
    theta1 = id * quarter_bar_width - bar_gap,
    r0 = inner_radius,
    r1 = inner_radius + (outer_radius_max - inner_radius) * prop / max(prop),
    theta_mid = (theta0 + theta1) / 2,
    label = paste0(case_when(
      state == state_levels[1] ~ "W+A",
      state == state_levels[2] ~ "W+D",
      state == state_levels[3] ~ "C+A",
      TRUE ~ "C+D"
    ), "\n", percent(prop, accuracy = 0.1)),
    label_r = pmin(r1 + 0.13, 1.12),
    label_x = label_r * cos(theta_mid),
    label_y = label_r * sin(theta_mid)
  )
radial_poly <- bind_rows(lapply(seq_len(nrow(radial_df)), function(i) {
  make_annular_sector(radial_df$theta0[i], radial_df$theta1[i], radial_df$r0[i], radial_df$r1[i], radial_df$id[i])
})) %>%
  left_join(radial_df %>% select(id, state), by = "id")

radial_plot <- ggplot() +
  geom_polygon(
    data = radial_poly,
    aes(x = x, y = y, group = id, fill = state),
    color = "white",
    linewidth = 0.35
  ) +
  # geom_path(
  #   data = tibble(
  #     theta = seq(0, pi / 2, length.out = 120),
  #     x = outer_radius_max * cos(theta),
  #     y = outer_radius_max * sin(theta)
  #   ),
  #   aes(x = x, y = y),
  #   color = alpha(col_text_soft, 0.45),
  #   linewidth = 0.25
  # ) +
  # geom_segment(aes(x = 0, y = 0, xend = outer_radius_max, yend = 0), color = alpha(col_text_soft, 0.55), linewidth = 0.28) +
  # geom_segment(aes(x = 0, y = 0, xend = 0, yend = outer_radius_max), color = alpha(col_text_soft, 0.55), linewidth = 0.28) +
  geom_text(
    data = radial_df,
    aes(x = label_x, y = label_y, label = label, color = state),
    size = 2.25,
    lineheight = 0.85,
    fontface = "bold"
  ) +
  annotate("text", x = 0.03, y = 0.08, label = "State\nshare", hjust = 0, vjust = 0, size = 2.55, color = col_text, fontface = "bold", lineheight = 0.88) +
  scale_fill_manual(values = state_cols, guide = "none") +
  scale_color_manual(values = state_cols, guide = "none") +
  coord_fixed(xlim = c(0, 1.68), ylim = c(0, 1.18), expand = FALSE, clip = "off") +
  theme_void() +
  theme(plot.margin = margin(0, 0, 0, 0))

figure_width <- 12
figure_height <- 7.4
main_x <- 0.07
main_y <- 0.12
main_w <- 0.68
main_h <- main_w * figure_width / 2 / figure_height
top_h <- 0.20
right_w <- top_h * figure_height / figure_width * 1.15
gap <- 0.025
top_y <- main_y + main_h + gap
right_x <- main_x + main_w + gap
radial_x <- right_x - 0.004
radial_y <- top_y + 0.008

p <- ggdraw() +
  draw_label(
    "Geographic distribution of warming-acceleration states",
    x = main_x,
    y = 0.98,
    hjust = 0,
    vjust = 1,
    fontface = "bold",
    size = 14,
    color = col_text
  ) +
  draw_label(
    paste0("Lake points coloured by four states; marginal ridgelines show 2° binned ",
           "longitude and latitude counts, square-root scaled and normalized separately for longitude and latitude ridges."),
    x = main_x,
    y = 0.94,
    hjust = 0,
    vjust = 1,
    size = 9.5,
    color = col_subtitle
  ) +
  draw_plot(top_plot, x = main_x, y = top_y, width = main_w, height = top_h) +
  draw_plot(main_plot, x = main_x, y = main_y, width = main_w, height = main_h) +
  draw_plot(right_plot, x = right_x, y = main_y, width = right_w, height = main_h) +
  draw_plot(radial_plot, x = radial_x, y = radial_y, width = right_w * 1.32, height = top_h * 0.96)

save_figure("fig-map-state-points-ridgelines.png", p, width = figure_width, height = figure_height)
p
```

## RSI intensity

::: {.callout-note}
## Finding 6
Warming-direction shifts have higher RSI than cooling-direction shifts:
median 1.25 vs 1.11. Cooling shifts are more concentrated at lower intensities.
:::

## First summary of detected regime shifts

STL trend -> annual mean -> first-differenced -> STARS ($L = 7$, $p = 0.05$).

::: {.callout-note}
## Finding 7
273,216 regime shifts are detected, or 2.96 per lake. Warming-direction shifts
account for 143,323 (52.5%); cooling-direction shifts account for 129,893
(47.5%).
:::

```{r}
#| label: fig-stars-breakpoint-count-bar
#| fig-cap: "Distribution of STARS-detected breakpoint counts per lake. Counts are grouped as 0, 1, 2, 3, and at least 4 breakpoints."
#| fig-width: 6.8
#| fig-height: 3.8

breakpoint_bar_cols <- c(dominant = "#b45309", other = "#d6d3d1")

bp <- read_csv(file.path(rdata, "bp_count_L7_p005.csv"), show_col_types = FALSE)

plot_df <- bp %>%
  mutate(group = case_when(
    n_bp == 0 ~ "0 shift",
    n_bp == 1 ~ "1 shift",
    n_bp == 2 ~ "2 shifts",
    n_bp == 3 ~ "3 shifts",
    n_bp >= 4 ~ "4+ shifts"
  )) %>%
  count(group, name = "n") %>%
  mutate(
    group = factor(group, levels = rev(c("0 shift", "1 shift", "2 shifts", "3 shifts", "4+ shifts")))
  ) %>%
  arrange(group) %>%
  mutate(
    prop = n / sum(n),
    label = paste0(comma(n), "  ", percent(prop, accuracy = 0.1)),
    emphasis = if_else(group == "3 shifts", "dominant", "other")
  )

max_n <- max(plot_df$n)

p <- ggplot(plot_df, aes(x = n, y = group)) +
  geom_col(aes(fill = emphasis), width = 0.62, color = col_text) +
  geom_text(aes(label = label), hjust = -0.08, size = 3.2, color = col_text) +
  scale_fill_manual(values = breakpoint_bar_cols, guide = "none") +
  scale_x_continuous(
    labels = comma,
    expand = expansion(mult = c(0, 0.20)),
    breaks = pretty_breaks(4)
  ) +
  labs(
    x = "Number of lakes",
    y = NULL
  ) +
  coord_cartesian(xlim = c(0, max_n * 1.18), clip = "off") +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = col_grid, linewidth = 0.28),
    axis.title.x = element_text(face = "bold", size = 9, color = "#292524", margin = margin(t = 8)),
    axis.text.x = element_text(size = 7.5, color = col_text_soft),
    axis.text.y = element_text(size = 9, color = "#292524", face = "bold"),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.margin = margin(8, 28, 8, 8)
  )

save_figure("fig-stars-breakpoint-count-bar.png", p, width = 6.8, height = 3.8)
p
```

::: {.callout-note}
## Finding 8
The 273,216 regime shifts divide the 92,245 lake time series into 365,425
regimes. The median regime length is 8 years; Q1--Q3 is 7--12 years. The most
common regime length is 7 years, matching the STARS window length $L = 7$.
:::

```{r}
#| label: fig-regime-length-distribution
#| fig-cap: "Distribution of regime lengths between STARS-detected shift points. Dashed lines mark the mean and median."
#| fig-width: 8
#| fig-height: 4.5

# --- Load precomputed data ---
regime_df <- read_csv(file.path(rdata, "regime_level_stats.csv"), show_col_types = FALSE) %>%
  mutate(length = as.integer(length))

# --- Stats ---
q_vals <- quantile(regime_df$length, c(0.05, 0.25, 0.50, 0.75, 0.95), na.rm = TRUE)
mean_val <- mean(regime_df$length, na.rm = TRUE)

# --- Main histogram ---
hist_data <- regime_df %>% count(length)

# Colors: muted teal bar, warm brown median, steel blue mean
col_bar <- "#5b8a8a"
col_median <- "#8b5e3c"
col_mean <- "#4a6fa5"

p <- ggplot(hist_data, aes(x = length, y = n)) +
  geom_col(fill = col_bar, color = "white", linewidth = 0.2, width = 0.85) +
  geom_vline(xintercept = q_vals[3], color = col_median, linewidth = 0.55, linetype = "dashed") +
  geom_vline(xintercept = mean_val, color = col_mean, linewidth = 0.55, linetype = "dashed") +
  annotate("text", x = q_vals[3] + 0.5, y = Inf, vjust = 1.8, hjust = 0,
           label = paste0("median = ", q_vals[3]), size = 3, color = col_median, fontface = "bold") +
  annotate("text", x = mean_val + 0.5, y = Inf, vjust = 3.2, hjust = 0,
           label = paste0("mean = ", sprintf("%.1f", mean_val)), size = 3, color = col_mean, fontface = "bold") +
  scale_x_continuous(breaks = seq(0, 40, by = 5), limits = c(0.5, 40.5), expand = c(0, 0)) +
  scale_y_continuous(labels = label_number(big.mark = ","), expand = expansion(mult = c(0, 0.08))) +
  labs(x = "Regime length (years)", y = "Count") +
  theme_atomic_base(grid = FALSE) +
  theme(plot.margin = margin(6, 10, 6, 4))

save_figure("fig-regime-length-distribution.png", p, width = 8, height = 4.5)
p
```

::: {.callout-note}
## Finding 9
The largest shift-year counts are 1999 (16,918 shifts) and 2014 (16,852
shifts), followed by 2006, 2013, and 1990.
:::

```{r}
#| label: fig-shifts-L7-rsi
#| fig-cap: "STARS-detected regime shifts by year ($L = 7$, $p = 0.05$). Red = warming-direction shift; blue = cooling-direction shift. Black line = mean signed RSI. Grey band = mean ± SD."
#| fig-width: 10.5
#| fig-height: 4.5

direction_cols <- pal_direction
direction_fills <- pal_direction_fill
direction_fills_alpha <- alpha(pal_direction_fill, 0.55)

shifts <- read_csv(
  file.path(rdata, "shifts_L7_p005.csv"),
  show_col_types = FALSE
)
df <- shifts %>% mutate(signed_rsi = if_else(direction == "warming", rsi, -rsi))
counts <- df %>%
  count(year, direction) %>%
  mutate(n_signed = if_else(direction == "warming", n, -n))
rsi_year <- df %>%
  group_by(year) %>%
  summarise(
    mean_signed_rsi = mean(signed_rsi),
    sd_signed_rsi = sd(signed_rsi),
    .groups = "drop"
  ) %>%
  mutate(
    rsi_lo = mean_signed_rsi - sd_signed_rsi,
    rsi_hi = mean_signed_rsi + sd_signed_rsi
  )

count_lim <- max(abs(counts$n_signed), na.rm = TRUE)
rsi_lim <- max(abs(c(rsi_year$rsi_lo, rsi_year$rsi_hi)), na.rm = TRUE)
scale_factor <- count_lim / rsi_lim

# Peak years for annotation
peak_years <- counts %>%
  group_by(year) %>%
  summarise(total = sum(abs(n_signed)), .groups = "drop") %>%
  slice_max(total, n = 3) %>%
  pull(year)

p <- ggplot() +
  # Bars: filled with alpha, solid border
  geom_col(
    data = counts,
    aes(x = year, y = n_signed, fill = direction),
    width = 0.82,
    alpha = 0.65,
    linewidth = 0
  ) +
  geom_col(
    data = counts,
    aes(x = year, y = n_signed, color = direction),
    width = 0.82,
    fill = NA,
    linewidth = 0.35
  ) +
  # RSI band and line
  geom_ribbon(
    data = rsi_year,
    aes(x = year, ymin = rsi_lo * scale_factor, ymax = rsi_hi * scale_factor),
    fill = "#1a1a1a",
    alpha = 0.10
  ) +
  geom_line(
    data = rsi_year,
    aes(x = year, y = mean_signed_rsi * scale_factor),
    color = "#1a1a1a",
    linewidth = 0.48
  ) +
  geom_point(
    data = rsi_year,
    aes(x = year, y = mean_signed_rsi * scale_factor),
    color = "#1a1a1a",
    size = 1.2,
    shape = 16
  ) +
  # Peak year vertical lines
  geom_vline(
    xintercept = peak_years,
    color = alpha(col_text_soft, 0.35),
    linewidth = 0.22,
    linetype = "dashed"
  ) +
  # Baseline
  geom_hline(yintercept = 0, color = "grey30", linewidth = 0.35) +
  # Scales
  scale_fill_manual(values = direction_fills, name = NULL) +
  scale_color_manual(values = direction_cols, guide = "none") +
  scale_x_continuous(
    breaks = seq(1985, 2020, 5),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_y_continuous(
    labels = label_number(big.mark = ","),
    name = "Shift count",
    sec.axis = sec_axis(~ . / scale_factor, name = "Mean signed RSI ± SD")
  ) +
  # Peak year labels
  geom_text(
    data = tibble(year = peak_years, y = rep(0, length(peak_years))),
    aes(x = year, y = y, label = year),
    vjust = 1.3,
    size = 2.8,
    color = col_text_soft,
    fontface = "bold"
  ) +
  # Labels
  labs(
    x = "Year"
  ) +
  theme_atomic_base(grid = FALSE) +
  theme(
    legend.position = c(0.01, 0.97),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = alpha("white", 0.85), color = NA),
    legend.key = element_rect(fill = "white", color = NA),
    legend.key.size = unit(0.55, "cm"),
    legend.spacing.x = unit(0.3, "cm"),
    axis.title.y.right = element_text(color = "#1a1a1a", face = "bold", size = 8.5),
    axis.text.y.right = element_text(color = "#1a1a1a", size = 7),
    panel.spacing = unit(0.5, "cm"),
    plot.margin = margin(6, 10, 4, 4)
  )

save_figure("fig-shifts-L7-rsi.png", p, width = 10.5, height = 4.5)
p
```
