# Regional decomposition {#sec-regional}

```{r}
#| label: setup-figures-regional
#| include: false

source("../../../../shared/R/figure-style.R")
```

## Continental overview

| Continent | n | Warming | Accel | Warming % | Accel % |
|---|---:|---:|---:|---:|---:|
| North America | 49,589 | 0.59 ± 0.64 | -0.53 ± 1.19 | 83.4% | 28.6% |
| Europe | 35,140 | 0.98 ± 0.87 | 1.38 ± 1.21 | 88.0% | 86.8% |
| Asia | 3,980 | 0.59 ± 0.54 | 0.58 ± 0.98 | 89.9% | 71.9% |
| South America | 2,406 | 0.39 ± 0.26 | 0.91 ± 0.53 | 92.6% | 96.4% |
| Africa | 712 | 0.50 ± 0.27 | 0.27 ± 0.64 | 99.9% | 66.6% |
| Oceania | 366 | 0.57 ± 0.29 | 0.77 ± 1.00 | 94.3% | 73.8% |

Warming is raw annual Sen warming in °C / 40 yr. Accel is Sen acceleration of
STL trend first difference in 10^-3^ °C / yr^2^.

::: {.callout-note}
## Finding 10
North America is the only continent with negative mean acceleration (-0.53 ×
10^-3^ °C / yr^2^), meaning most lakes there are warming but decelerating.
Europe has the strongest acceleration (+1.38).
:::

Continental labels are convenient for description but do not correspond to
natural units of climate response. A lake in southern North America may share
more with a lake in northern South America than with a lake in northern Canada.
We therefore move from continental description to climate-based classification.

## Temporal response types

We classify lakes by the shape of their 40-year STL trend, combined with their
mean temperature. Each lake's annual STL trend is z-scored to remove mean and
scale, then combined with the normalised mean STL temperature. The resulting
feature vector is classified by $k$-means clustering ($K = 8$).

The clusters describe temporal response types: lakes with similar warming
trajectories and temperature regimes are grouped together regardless of
continental labels.

::: {.callout-note}
## Finding 11
Eight temporal response types are identified. Most clusters are geographically
coherent and physically interpretable: cold northwestern North America (C3,
C7), eastern North America (C6), warm and cold European regimes (C2, C8), and
a tropical belt (C4).
:::

```{r}
#| label: fig-ts-cluster-map-ridges
#| fig-cap: "Spatial distribution of the eight time-series response clusters. Marginal ridgelines show longitude and latitude density by cluster."
#| fig-width: 12
#| fig-height: 7.4

# --- Shared cluster data ---
cluster_data <- prepare_cluster_data(include_area = TRUE)
trend_sub <- cluster_data$trend_sub
trend_mat <- cluster_data$trend_mat
year_cols <- cluster_data$year_cols
cluster_meta <- trend_sub %>%
  group_by(cluster_label) %>%
  summarise(
    n = n(),
    mean_temp = mean(mean_temp),
    mean_warming = mean(raw_sen_warming_40yr),
    .groups = "drop"
  )
cluster_order <- cluster_data$cluster_order$cluster
new_labels <- paste0("C", 1:8)
names(new_labels) <- as.character(cluster_order)
cluster_cols <- pal_cluster
cluster_labels <- paste0(
  new_labels[as.character(cluster_order)], ": ",
  comma(cluster_meta$n[order(cluster_meta$mean_temp)]), " lakes, ",
  sprintf("%.1f", cluster_meta$mean_temp[order(cluster_meta$mean_temp)]), "°C"
)
names(cluster_labels) <- new_labels[as.character(cluster_order)]
coast <- read_csv(file.path(rdata, "coastline.csv"), show_col_types = FALSE)


# --- Ridge parameters ---
ridge_levels <- new_labels[as.character(cluster_order)]
ridge_height <- 0.92
ridge_gap <- 0.54
smooth_count <- function(x) {
  smoothed <- as.numeric(stats::filter(x, rep(1 / 3, 3), sides = 2))
  ifelse(is.na(smoothed), x, smoothed)
}
ridge_base <- rev(seq_along(ridge_levels)) * ridge_gap
names(ridge_base) <- ridge_levels

lon_min <- -180; lon_max <- 180; lat_min <- -90; lat_max <- 90
bin_width <- 2
lon_centers <- seq(lon_min + bin_width / 2, lon_max - bin_width / 2, by = bin_width)
lat_centers <- seq(lat_min + bin_width / 2, lat_max - bin_width / 2, by = bin_width)

# --- Ridge data ---
lon_counts <- trend_sub %>%
  mutate(lon_center = lon_min + floor((lon - lon_min) / bin_width) * bin_width + bin_width / 2) %>%
  filter(lon_center >= min(lon_centers), lon_center <= max(lon_centers)) %>%
  count(cluster_label, lon_center, name = "n") %>%
  tidyr::complete(cluster_label = ridge_levels, lon_center = lon_centers, fill = list(n = 0)) %>%
  group_by(cluster_label) %>%
  mutate(n_smooth = smooth_count(sqrt(n)), baseline = ridge_base[cluster_label]) %>%
  ungroup()

lon_global_max <- max(lon_counts$n_smooth, na.rm = TRUE)
lon_counts <- lon_counts %>%
  mutate(n_scaled = n_smooth / lon_global_max, y = baseline + n_scaled * ridge_height) %>%
  arrange(desc(baseline), lon_center)

lat_counts <- trend_sub %>%
  mutate(lat_center = lat_min + floor((lat - lat_min) / bin_width) * bin_width + bin_width / 2) %>%
  filter(lat_center >= min(lat_centers), lat_center <= max(lat_centers)) %>%
  count(cluster_label, lat_center, name = "n") %>%
  tidyr::complete(cluster_label = ridge_levels, lat_center = lat_centers, fill = list(n = 0)) %>%
  group_by(cluster_label) %>%
  mutate(n_smooth = smooth_count(sqrt(n)), baseline = ridge_base[cluster_label]) %>%
  ungroup()

lat_global_max <- max(lat_counts$n_smooth, na.rm = TRUE)
lat_counts <- lat_counts %>%
  mutate(n_scaled = n_smooth / lat_global_max, x = baseline + n_scaled * ridge_height) %>%
  arrange(desc(baseline), lat_center)

ridge_base_df <- tibble(cluster_label = ridge_levels, baseline = ridge_base)

# --- Build plots ---
theme_map <- theme_atomic_map()

main_plot <- ggplot() +
  geom_path(data = coast, aes(x = lon, y = lat, group = segment), color = "grey35", linewidth = 0.18) +
  geom_point(data = trend_sub, aes(x = lon, y = lat, colour = cluster_label, size = point_size), alpha = 0.48, stroke = 0.1) +
  scale_color_manual(values = cluster_cols, labels = cluster_labels, name = NULL) +
  scale_size_identity(guide = "none") +
  coord_fixed(ratio = 1, xlim = c(lon_min, lon_max), ylim = c(lat_min, lat_max), expand = FALSE) +
  labs(x = NULL, y = NULL) +
  theme_map +
  theme(
    legend.position = "right",
    axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(),
    plot.margin = margin(0, 0, 0, 0)
  )

ridge_theme <- theme_void(base_size = 8) + theme(plot.margin = margin(0, 0, 0, 0), legend.position = "none")

top_plot <- ggplot(lon_counts, aes(x = lon_center, group = cluster_label, color = cluster_label, fill = cluster_label)) +
  geom_ribbon(aes(ymin = baseline, ymax = y), alpha = 0.24, color = NA) +
  geom_line(aes(y = y), linewidth = 0.52, alpha = 0.96) +
  geom_hline(data = ridge_base_df, aes(yintercept = baseline, color = cluster_label), linewidth = 0.18, alpha = 0.40) +
  scale_color_manual(values = cluster_cols, guide = "none") +
  scale_fill_manual(values = cluster_cols, guide = "none") +
  scale_x_continuous(limits = c(lon_min, lon_max), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0.45, max(ridge_base) + ridge_height + 0.28), expand = c(0, 0)) +
  ridge_theme

right_plot <- ggplot(lat_counts, aes(y = lat_center, group = cluster_label, color = cluster_label, fill = cluster_label)) +
  geom_ribbon(aes(xmin = baseline, xmax = x), orientation = "y", alpha = 0.24, color = NA) +
  geom_line(aes(x = x), linewidth = 0.52, alpha = 0.96, orientation = "y") +
  geom_vline(data = ridge_base_df, aes(xintercept = baseline, color = cluster_label), linewidth = 0.18, alpha = 0.40) +
  scale_color_manual(values = cluster_cols, guide = "none") +
  scale_fill_manual(values = cluster_cols, guide = "none") +
  scale_y_continuous(limits = c(lat_min, lat_max), expand = c(0, 0)) +
  scale_x_continuous(limits = c(0.45, max(ridge_base) + ridge_height + 0.28), expand = c(0, 0)) +
  ridge_theme

# --- Extract legend ---
legend_grob <- get_legend(
  main_plot +
    guides(color = guide_legend(override.aes = list(size = 2.6, alpha = 0.9), ncol = 1)) +
    theme(legend.position = "right")
)
main_plot <- main_plot + theme(legend.position = "none")

# --- Layout ---
figure_width <- 12
figure_height <- 7.4
main_x <- 0.07; main_y <- 0.12
main_w <- 0.68; main_h <- main_w * figure_width / 2 / figure_height
top_h <- 0.20
right_w <- top_h * figure_height / figure_width * 1.15
gap <- 0.025
top_y <- main_y + main_h + gap
right_x <- main_x + main_w + gap
legend_x <- right_x - 0.002
legend_y <- top_y + 0.035

p <- ggdraw() +
  draw_label("Time-series shape clustering of global lakes (K=8, shape+temp)",
             x = main_x, y = 0.98, hjust = 0, vjust = 1, fontface = "bold", size = 14, color = col_text) +
  draw_label("K-means on z-scored annual STL trend + mean temperature. Clusters ordered by mean temperature. Point size = lake area.",
             x = main_x, y = 0.94, hjust = 0, vjust = 1, size = 9.5, color = col_subtitle) +
  draw_plot(top_plot, x = main_x, y = top_y, width = main_w, height = top_h) +
  draw_plot(main_plot, x = main_x, y = main_y, width = main_w, height = main_h) +
  draw_plot(right_plot, x = right_x, y = main_y, width = right_w, height = main_h) +
  draw_plot(legend_grob, x = legend_x, y = legend_y, width = 0.25, height = 0.22)

save_figure("fig-ts-cluster-map-ridges.png", p, width = figure_width, height = figure_height)
p
```

### Cluster descriptions

**C1: Cold high-latitude North America.** This is the coldest cluster, with a
mean STL trend of 1.8 °C. Warming is moderate (+0.36 °C / 40 yr, 99% positive)
and acceleration is weakly positive (+0.13 × 10^-3^ °C / yr^2^, 60% positive).
It is concentrated in northern North America (97%) with high spatial coherence
(Moran's I = 0.65). Regime-shift peaks occur in 1990--91 and 2006 (warming),
and 1997--99 (cooling).

**C2: Decelerating high-latitude North America.** Mean temperature is 2.8 °C.
Warming is moderate (+0.35 °C / 40 yr, 99% positive), but acceleration is
strongly negative (-1.33 × 10^-3^ °C / yr^2^, only 1% positive), meaning most
lakes in this cluster are warming but decelerating. It is concentrated in
northern North America (96%) with Moran's I = 0.70. Regime-shift peaks occur in
2003--05 (warming), 1997 (cooling), and 2010--12 (cooling).

**C3: Strongly accelerating Europe.** Mean temperature is 3.1 °C. Warming is
strong (+0.71 °C / 40 yr, 99% positive), and acceleration is the highest among
all clusters (+1.98 × 10^-3^ °C / yr^2^, 98% positive). It is dominated by
Europe (74%) with Moran's I = 0.65. Regime-shift peaks occur in 1989 and
2013--15 (warming), and 1989--93 and 2006 (cooling).

**C4: Cold decelerating North America.** Mean temperature is 3.4 °C. Warming is
weak (+0.29 °C / 40 yr, 99% positive), and acceleration is strongly negative
(-1.02 × 10^-3^ °C / yr^2^, only 5% positive). It is the most spatially coherent
cluster (Moran's I = 0.93), concentrated almost entirely in North America
(99%). Regime-shift peaks occur in 2005--07 (warming), 1998 (cooling), and
2013--14 (cooling).

**C5: Warming Europe.** This is the largest cluster (21,493 lakes). Mean
temperature is 4.3 °C. Warming is strong (+0.84 °C / 40 yr, 100% positive), and
acceleration is weakly positive (+0.37 × 10^-3^ °C / yr^2^, 65% positive). It is
dominated by Europe (68%) with Moran's I = 0.61. Regime-shift peaks occur in
1998--99 and 2013--14 (warming), and 1992 and 2005--07 (cooling).

**C6: Decelerating eastern North America.** Mean temperature is 5.0 °C. Warming
is moderate (+0.46 °C / 40 yr, 100% positive), but acceleration is strongly
negative (-1.02 × 10^-3^ °C / yr^2^, only 12% positive). It is concentrated in
North America (97%) with Moran's I = 0.80. Regime-shift peaks occur in 1992--95
(warming) and 1999--02 (cooling).

**C7: Strongly accelerating Europe.** Mean temperature is 5.0 °C. Warming is
moderate (+0.57 °C / 40 yr, 99% positive), and acceleration is strongly
positive (+1.83 × 10^-3^ °C / yr^2^, 95% positive). It is dominated by Europe
(88%) with Moran's I = 0.72. Regime-shift peaks occur in 1997--2000 (warming)
and 2005--08 (cooling).

**C8: Warm tropical/subtropical lakes.** This is the warmest cluster (mean
21.9 °C). Warming is moderate (+0.57 °C / 40 yr, 99% positive), and
acceleration is positive (+0.69 × 10^-3^ °C / yr^2^, 85% positive). It is
geographically dispersed but concentrated in the subtropical latitude band
(Moran's I = 0.67). Regime-shift peaks occur in 2009--12 (warming), 1989--91
(cooling), and 2000 (cooling).

### Cluster summary

```{r}
#| label: fig-cluster-continent-heatmap
#| fig-cap: "Cluster--continent allocation. Each row sums to 100%. Right column = total lakes per cluster. Bottom row = total lakes per continent."
#| fig-width: 4
#| fig-height: 4

# --- Shared cluster data ---
cluster_data <- prepare_cluster_data()
trend_sub <- cluster_data$trend_sub

# --- Cluster × Continent counts ---
raw_counts <- trend_sub %>%
  count(cluster_label, Continent, name = "n")

row_totals <- raw_counts %>% group_by(cluster_label) %>% summarise(total = sum(n), .groups = "drop")
col_totals <- raw_counts %>% group_by(Continent) %>% summarise(total = sum(n), .groups = "drop")
grand_total <- sum(raw_counts$n)

cluster_levels <- paste0("C", 1:8)
cont_levels <- col_totals %>% arrange(desc(total)) %>% pull(Continent)

# --- Main heatmap cells ---
main_cells <- raw_counts %>%
  left_join(row_totals, by = "cluster_label") %>%
  left_join(col_totals, by = "Continent", suffix = c("_row", "_col")) %>%
  mutate(
    frac_row = n / total_row,  # row-normalized: fraction of cluster across continents
    frac_col = n / total_col,  # col-normalized: fraction of clusters within continent
  ) %>%
  mutate(
    cluster_label = factor(cluster_label, levels = c(cluster_levels, "Total")),
    Continent = factor(Continent, levels = c(as.character(cont_levels), "Total")),
    display = comma(n)
  )

# --- Row totals (right column) ---
row_total_cells <- row_totals %>%
  mutate(
    cluster_label = factor(cluster_label, levels = c(cluster_levels, "Total")),
    Continent = factor("Total", levels = c(as.character(cont_levels), "Total")),
    frac = total / grand_total,
    display = comma(total)
  )

# --- Column totals (bottom row) ---
col_total_cells <- col_totals %>%
  mutate(
    cluster_label = factor("Total", levels = c(cluster_levels, "Total")),
    Continent = factor(Continent, levels = c(as.character(cont_levels), "Total")),
    frac = total / grand_total,
    display = comma(total)
  )

# --- Grand total ---
grand_cell <- tibble(
  cluster_label = factor("Total", levels = c(cluster_levels, "Total")),
  Continent = factor("Total", levels = c(as.character(cont_levels), "Total")),
  frac = 1, display = comma(grand_total)
)

# --- Row bar data (horizontal bars per cluster) ---
row_bars <- main_cells %>%
  mutate(
    x_num = as.numeric(Continent),
    y_num = as.numeric(cluster_label)
  ) %>%
  group_by(cluster_label) %>%
  mutate(
    # Bar width fills cell, height = fraction directly (rows sum to 100%)
    bar_left = x_num - 0.5,
    bar_right = x_num - 0.5 + 1.0,
    bar_bottom = y_num - 0.5,
    bar_top = y_num - 0.5 + 0.8 * frac_row
  ) %>%
  ungroup()

# --- Column bar data (vertical bars per continent) ---
col_bars <- main_cells %>%
  mutate(
    x_num = as.numeric(Continent),
    y_num = as.numeric(cluster_label)
  ) %>%
  group_by(Continent) %>%
  mutate(
    # Bar height fills cell, width = fraction directly (columns sum to 100%)
    bar_left = x_num - 0.5,
    bar_right = x_num - 0.5 + 0.8 * frac_col,
    bar_bottom = y_num - 0.5,
    bar_top = y_num + 0.5
  ) %>%
  ungroup()

# --- Combine all cells ---
all_cells <- bind_rows(
  main_cells %>% mutate(cell_type = "main"),
  row_total_cells %>% mutate(cell_type = "row_total"),
  col_total_cells %>% mutate(cell_type = "col_total"),
  grand_cell %>% mutate(cell_type = "grand_total")
) %>%
  mutate(
    text_color = case_when(
      cell_type == "main" ~ col_text,
      TRUE ~ col_text_soft
    )
  )

# --- Plot ---
p <- ggplot() +
  # Main heatmap tiles (white background)
  geom_tile(
    data = all_cells %>% filter(cell_type == "main"),
    aes(x = Continent, y = cluster_label),
    fill = "white", color = "grey85", linewidth = 0.3
  ) +
  # Total cells: grey background
  geom_tile(
    data = all_cells %>% filter(cell_type != "main"),
    aes(x = Continent, y = cluster_label),
    fill = "grey92", color = "white", linewidth = 0.4
  ) +
  # Row bars (horizontal, from bottom edge)
  geom_rect(
    data = row_bars,
    aes(xmin = bar_left, xmax = bar_right, ymin = bar_bottom, ymax = bar_top),
    fill = alpha(pal_direction[["warming"]], 0.5), color = NA
  ) +
  # Column bars (vertical, from left edge)
  geom_rect(
    data = col_bars,
    aes(xmin = bar_left, xmax = bar_right, ymin = bar_bottom, ymax = bar_top),
    fill = alpha(pal_direction[["cooling"]], 0.5), color = NA
  ) +
  # Text labels
  geom_text(
    data = all_cells,
    aes(x = Continent, y = cluster_label, label = display, color = text_color),
    size = 2.8
  ) +
  scale_fill_identity() +
  scale_color_identity() +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(
    x = NULL, y = "Cluster"
  ) +
  theme_atomic_base() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, size = 8),
    axis.text.y = element_text(size = 8),
    axis.title.y = element_text(size = 9, face = "bold"),
    legend.position = "none",
    plot.margin = margin(6, 6, 4, 4)
  )

save_figure("fig-cluster-continent-heatmap.png", p, width = 4, height = 4)
p
```

| Cluster | n | Temp | Warming | Accel | Dominant | Moran's I |
|---|---:|---:|---:|---:|---|---:|
| C1 | 8,636 | 1.8 °C | +0.36 | +0.13 | NA 97% | 0.65 |
| C2 | 7,855 | 2.8 °C | +0.35 | -1.33 | NA 96% | 0.70 |
| C3 | 16,401 | 3.1 °C | +0.71 | +1.98 | EU 74% | 0.65 |
| C4 | 9,667 | 3.4 °C | +0.29 | -1.02 | NA 99% | 0.93 |
| C5 | 21,493 | 4.3 °C | +0.84 | +0.37 | EU 68% | 0.61 |
| C6 | 13,768 | 5.0 °C | +0.46 | -1.02 | NA 97% | 0.80 |
| C7 | 8,649 | 5.0 °C | +0.57 | +1.83 | EU 88% | 0.72 |
| C8 | 5,776 | 21.9 °C | +0.57 | +0.69 | SA 30% | 0.67 |

Temp is mean STL trend in °C. Warming is Sen slope of annual STL trend in
°C / 40 yr. Accel is Sen slope of first-differenced STL trend in 10^-3^
°C / yr^2^. Dominant is the continent with the highest fraction of lakes.

## Regime-shift timing by response type

Each cluster has its own characteristic regime-shift timing. If the clustering
captures real differences in climate response, lakes within the same cluster
should share similar shift-year patterns.

::: {.callout-note}
## Finding 12
Each cluster shows distinct shift-year peaks. The eight clusters differ not
only in spatial pattern but also in physical characteristics. Temperature
level, warming rate, acceleration, and regime-shift timing all vary
systematically across clusters. STL trend shapes and shift timing are
consistent within clusters and distinct between clusters.
:::

```{r}
#| label: fig-cluster-characteristics
#| fig-cap: "Cluster characteristics: distributions of temperature, warming, and acceleration, plus mean STL trend by cluster."
#| fig-width: 10
#| fig-height: 7.2

# --- Shared cluster data ---
cluster_data <- prepare_cluster_data(include_accel = TRUE, include_warming = TRUE)
trend_sub <- cluster_data$trend_sub
trend_mat <- cluster_data$trend_mat
year_cols <- cluster_data$year_cols
cluster_meta <- cluster_data$cluster_meta
label_map <- cluster_data$label_map
cluster_cols <- pal_cluster
direction_fills <- pal_direction_fill
direction_cols_line <- pal_direction
all_clusters <- paste0("C", 1:8)
shifts <- read_csv(file.path(rdata, "shifts_L7_p005.csv"), show_col_types = FALSE)

# --- Shift timing data ---
shifts_cluster <- shifts %>%
  inner_join(trend_sub %>% select(lake_id, cluster_label), by = "lake_id") %>%
  mutate(direction = factor(direction, levels = c("warming", "cooling")))

year_range <- 1988:2020
year_dist <- shifts_cluster %>%
  count(cluster_label, year, direction) %>%
  tidyr::complete(cluster_label = paste0("C", 1:8), year = year_range, direction = c("warming", "cooling"), fill = list(n = 0)) %>%
  filter(year >= 1989) %>%
  mutate(
    signed_n = if_else(direction == "warming", n, -n),
    direction = factor(direction, levels = c("warming", "cooling")),
    cluster_label = factor(cluster_label, levels = paste0("C", 1:8))
  )

# ============================================================
# LEFT COLUMN: 4 panels, compact

# --- Peak detection algorithm ---
# For each cluster+direction: smooth with 3-year rolling mean, find local maxima,
# filter by 75th percentile threshold.
detect_peaks <- function(years, counts) {
  n <- length(counts)
  if (n < 5) return(tibble(year = integer(), peak_count = numeric(), is_range = logical()))
  
  # 3-year rolling mean
  smoothed <- as.numeric(stats::filter(counts, rep(1/3, 3), sides = 2))
  smoothed[is.na(smoothed)] <- counts[is.na(smoothed)]
  
  # Threshold: 90th percentile of positive smoothed values
  pos_vals <- smoothed[smoothed > 0]
  if (length(pos_vals) < 3) return(tibble(year = integer(), peak_count = numeric(), is_range = logical()))
  threshold <- quantile(pos_vals, 0.75, na.rm = TRUE)
  
  # All years above threshold
  above_idx <- which(smoothed >= threshold)
  if (length(above_idx) == 0) return(tibble(year = integer(), peak_count = numeric(), is_range = logical()))
  
  peak_years <- years[above_idx]
  peak_counts <- smoothed[above_idx]
  
  # Group consecutive years (gap > 2 = new group)
  groups <- cumsum(c(1, diff(peak_years) > 2))
  result <- list()
  for (g in unique(groups)) {
    idx <- which(groups == g)
    yrs <- peak_years[idx]
    cnts <- peak_counts[idx]
    best <- which.max(cnts)
    if (length(yrs) == 1) {
      result[[length(result) + 1]] <- tibble(year = yrs, peak_count = cnts, is_range = FALSE)
    } else {
      result[[length(result) + 1]] <- tibble(
        year = yrs[best], peak_count = cnts[best],
        is_range = TRUE, range_min = min(yrs), range_max = max(yrs)
      )
    }
  }
  bind_rows(result)
}

# Detect peaks for each cluster+direction
peaks <- year_dist %>%
  group_by(cluster_label, direction) %>%
  group_modify(~ detect_peaks(.x$year, .x$n)) %>%
  ungroup() %>%
  mutate(
    label = if_else(is_range,
      paste0(range_min, "-", substr(range_max, 3, 4)),
      as.character(year)
    ),
    y_label = if_else(direction == "warming", peak_count * 1.15, -peak_count * 1.15)
  )

# Peak ranges for shading
peak_ranges <- peaks %>%
  filter(is_range) %>%
  mutate(
    x_min = range_min,
    x_max = range_max
  ) %>%
  select(cluster_label, direction, x_min, x_max)

# ============================================================

make_violin <- function(data, y_col, y_label, title_text) {
  ggplot(data, aes(x = cluster_label, y = .data[[y_col]])) +
    geom_violin(aes(color = cluster_label), fill = NA, scale = "width", linewidth = 0.35) +
    geom_boxplot(width = 0.10, fill = "white", color = "grey30", linewidth = 0.25, outlier.shape = NA) +
    geom_hline(yintercept = 0, color = "grey50", linewidth = 0.2, linetype = "dashed") +
    scale_color_manual(values = cluster_cols, guide = "none") +
    scale_x_discrete(labels = label_map) +
    labs(x = NULL, y = y_label, title = title_text) +
    theme_atomic_base(grid = FALSE) +
    theme(
      axis.text.x = element_text(angle = 35, hjust = 1, size = 5.5),
      axis.text.y = element_text(size = 6),
      axis.title.y = element_text(size = 7, face = "bold"),
      plot.title = element_text(size = 8, face = "bold"),
      plot.margin = margin(1, 2, 1, 1)
    )
}

p_v1 <- make_violin(trend_sub, "mean_temp", "Mean STL trend (\u00b0C)", "Temperature level")
p_v2 <- make_violin(trend_sub, "warming", "Warming rate (\u00b0C/40yr)", "Warming rate")
p_v3 <- make_violin(trend_sub, "accel", expression("Acceleration (10"^-3 * " °C/yr"^2 * ")"), "Acceleration")

# --- STL trend panels ---
cluster_ts <- trend_sub %>%
  select(cluster_label, all_of(year_cols)) %>%
  group_by(cluster_label) %>%
  summarise(
    across(all_of(year_cols), list(mean = ~ mean(.x, na.rm = TRUE), sd = ~ sd(.x, na.rm = TRUE))),
    .groups = "drop"
  ) %>%
  pivot_longer(-cluster_label, names_to = "col_name", values_to = "value") %>%
  mutate(
    stat = if_else(grepl("_mean$", col_name), "mean", "sd"),
    year = as.integer(gsub("trend_|_(mean|sd)$", "", col_name))
  ) %>%
  select(cluster_label, year, stat, value) %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  filter(!is.na(year))

# C1-C7: y-axis from full mean ± SD range
y_range_1_7 <- cluster_ts %>% filter(cluster_label != "C8") %>%
  { range(c(.$mean - .$sd, .$mean + .$sd), na.rm = TRUE) }
y_pad_1_7 <- diff(y_range_1_7) * 0.03
y_limits_1_7 <- c(y_range_1_7[1] - y_pad_1_7, y_range_1_7[2] + y_pad_1_7)

y_range_all <- range(c(cluster_ts$mean - cluster_ts$sd, cluster_ts$mean + cluster_ts$sd), na.rm = TRUE)
y_pad_all <- diff(y_range_all) * 0.05
y_limits_all <- c(y_range_all[1] - y_pad_all, y_range_all[2] + y_pad_all)

trend_panels <- list()

for (i in 1:8) {
  highlight <- all_clusters[i]
  y_lim <- if (i == 8) y_limits_all else y_limits_1_7
  range_sub <- peak_ranges %>% filter(cluster_label == highlight)
  # Only draw individual peak lines for non-range peaks
  peak_sub <- peaks %>% filter(cluster_label == highlight, !is_range)
  
  p <- ggplot() +
    # Peak range shading (higher alpha)
    {if (nrow(range_sub) > 0) geom_rect(
      data = range_sub,
      aes(xmin = x_min, xmax = x_max, ymin = -Inf, ymax = Inf, fill = direction),
      inherit.aes = FALSE, alpha = 0.18
    )} +
    # Range boundary lines (dashed, thin, at start and end of each range)
    {if (nrow(range_sub) > 0) geom_vline(
      data = range_sub %>% pivot_longer(c(x_min, x_max), values_to = "x_pos"),
      aes(xintercept = x_pos, color = direction),
      inherit.aes = FALSE, linewidth = 0.3, linetype = "dashed", alpha = 0.5
    )} +
    # Grey lines for other clusters
    geom_line(
      data = cluster_ts %>% filter(cluster_label != highlight),
      aes(x = year, y = mean, group = cluster_label),
      color = "grey65", linewidth = 0.25, alpha = 0.5
    ) +
    # SD ribbon for highlighted cluster
    geom_ribbon(
      data = cluster_ts %>% filter(cluster_label == highlight),
      aes(x = year, ymin = mean - sd, ymax = mean + sd),
      fill = cluster_cols[highlight], alpha = 0.2, color = NA
    ) +
    # Mean line for highlighted cluster
    geom_line(
      data = cluster_ts %>% filter(cluster_label == highlight),
      aes(x = year, y = mean),
      color = cluster_cols[highlight], linewidth = 0.6
    ) +
    # Peak year lines (thin, semi-transparent)
    {if (nrow(peak_sub) > 0) geom_vline(
      data = peak_sub,
      aes(xintercept = year, color = direction),
      linewidth = 0.3, linetype = "dashed", alpha = 0.5
    )} +
    # Peak year labels
    {if (nrow(peak_sub) > 0) geom_text(
      data = peak_sub,
      aes(x = year, y = Inf, label = label, color = direction),
      vjust = 1.5, hjust = 0.5, size = 2, fontface = "bold"
    )} +
    scale_color_manual(values = direction_cols_line, guide = "none") +
    scale_fill_manual(values = direction_fills, guide = "none") +
    scale_x_continuous(breaks = seq(1985, 2020, by = 10)) +
    scale_y_continuous(limits = y_lim, breaks = pretty_breaks(3)) +
    labs(title = highlight) +
    theme_void(base_size = 7) +
    theme(
      plot.title = element_text(size = 7, face = "bold", color = cluster_cols[highlight], hjust = 0.5),
      axis.text.y = element_text(size = 5, color = "grey50"),
      axis.ticks.y = element_line(color = "grey50", linewidth = 0.15),
      axis.line.y = element_line(color = "grey50", linewidth = 0.15),
      plot.margin = margin(1, 1, 1, 1)
    )
  
  trend_panels[[i]] <- p
}

p_trends <- wrap_plots(trend_panels, ncol = 1) +
  plot_annotation(
    subtitle = "C1-C7: shared y-axis. C8: own y-axis.",
    theme = theme(
      plot.subtitle = element_text(size = 6, color = col_subtitle)
    )
  )

# --- Info panel for bottom-left ---
info_text <- paste0(
  "Cluster summary\n\n",
  paste0(
    cluster_meta$cluster_label, ": ",
    comma(cluster_meta$n), " lakes, ",
    sprintf("%.1f", cluster_meta$mt), "°C",
    collapse = "\n"
  )
)
p_info <- ggplot() +
  annotate("text", x = 0.5, y = 0.95, label = "Cluster summary",
           size = 4.5, fontface = "bold", color = col_text, hjust = 0.5, vjust = 1) +
  annotate("text", x = 0.5, y = 0.82, label = info_text,
           size = 3.0, color = col_text_soft, hjust = 0.5, vjust = 1, family = "mono", lineheight = 1.1) +
  xlim(0, 1) + ylim(0, 1) +
  theme_void() +
  theme(plot.margin = margin(4, 4, 4, 4))

p_left <- p_v1 / p_v2 / p_v3 / p_info +
  plot_layout(heights = c(1, 1, 1, 1))

# ============================================================
# RIGHT COLUMN: 8 shift timing charts
# ============================================================

shift_panels <- list()
for (i in 1:8) {
  cl <- all_clusters[i]
  sub <- year_dist %>% filter(cluster_label == cl)
  # Prepare peak annotations: range = horizontal segment, single = vertical line
  peak_sub <- peaks %>% filter(cluster_label == cl)
  range_sub <- peak_ranges %>% filter(cluster_label == cl)
  
  # Compute y offset for directed peak marks (small fraction of max count)
  max_n <- max(abs(sub$signed_n), na.rm = TRUE)
  y_off <- max_n * 0.08
  
  range_marks <- NULL
  single_marks <- NULL
  
  p <- ggplot(sub, aes(x = year, y = signed_n, fill = direction)) +
    # Low-alpha cluster color background
    annotate("rect", xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf,
             fill = cluster_cols[cl], alpha = 0.15) +
    # Peak range shading
    {if (nrow(range_sub) > 0) geom_rect(
      data = range_sub,
      aes(xmin = x_min, xmax = x_max, ymin = -Inf, ymax = Inf, fill = direction),
      inherit.aes = FALSE, alpha = 0.1
    )} +
    geom_col(width = 0.82, alpha = 0.65, linewidth = 0) +
    geom_col(width = 0.82, fill = NA, aes(color = direction), linewidth = 0.2) +
    geom_hline(yintercept = 0, color = "grey30", linewidth = 0.15) +
    scale_fill_manual(values = c(direction_fills, direction_fills), guide = "none") +
    scale_color_manual(values = direction_cols_line, guide = "none") +
    scale_x_continuous(breaks = seq(1990, 2020, by = 10), limits = c(1988.5, 2020.5), expand = c(0, 0)) +
    scale_y_continuous(labels = label_number(big.mark = ",")) +
    labs(title = cl, x = NULL, y = NULL) +
    theme_atomic_base(grid = FALSE) +
    theme(
      plot.title = element_text(size = 7, face = "bold", color = cluster_cols[cl], hjust = 0.5),
      plot.background = element_rect(fill = "white", color = NA),
      axis.text = element_text(size = 4.5),
      axis.text.x = element_text(angle = 0),
      plot.margin = margin(2, 3, 2, 3)
    )
  
  shift_panels[[i]] <- p
}

p_right <- wrap_plots(shift_panels, ncol = 1) +
  plot_annotation(
    subtitle = "Positive = warming-direction; negative = cooling-direction.",
    theme = theme(
      plot.subtitle = element_text(size = 6, color = col_subtitle)
    )
  )

# ============================================================
# COMPOSE: characteristics panels
# ============================================================
p_characteristics <- p_left | p_trends +
  plot_layout(widths = c(1, 1.15))

save_figure("fig-cluster-characteristics.png", p_characteristics, width = 10, height = 7.2)
p_characteristics
```

```{r}
#| label: fig-cluster-shift-timing
#| fig-cap: "Regime-shift timing by cluster. Positive bars are warming shifts; negative bars are cooling shifts."
#| fig-width: 7.2
#| fig-height: 8.8

save_figure("fig-cluster-shift-timing.png", p_right, width = 7.2, height = 8.8)
p_right
```

## Spatial coherence

The eight clusters exhibit different spatial coherence patterns. Some form
tight geographic clusters; others form latitudinal bands spanning multiple
continents.

::: {.callout-note}
## Finding 13
Five clusters (C2, C3, C5, C6, C7) form tight geographic clusters, with more
than 80% of lakes within ±30° of their centroid. Two clusters (C1, C8) form
latitudinal bands spanning the full longitude range at high latitudes. One
cluster (C4, tropical) is globally dispersed but concentrated in the
subtropical latitude band.
:::
