# Teleconnection attribution {#sec-teleconnection}

```{r}
#| label: setup-figures-teleconnection
#| include: false

source("../../../../shared/R/figure-style.R")
```

## Background

Large-scale climate modes influence lake thermal regimes through atmospheric and
oceanic teleconnections. We evaluate how five major teleconnection indices--ENSO
(Nino3.4), PDO, AMO, NAO, and AO--contribute to the observed STL trend and its
first difference across the eight temporal response types.

The analysis proceeds in two levels: (1) regression of the annual STL trend on
the five indices, and (2) regression of the first-differenced STL trend on the
same indices. Level 1 captures which indices drive the long-term trend level;
Level 2 captures which indices drive year-to-year changes in trend slope.

## Level 1: Trend attribution

::: {.callout-note}
## Finding 14
AMO is the dominant driver of STL trend across all clusters. The global mean
standardised coefficient is $\beta = 0.56$, with a partial $R^2$ of 0.34. PDO
is the second strongest driver ($\beta = -0.18$, partial $R^2 = 0.05$). ENSO,
NAO, and AO have weak independent effects.
:::

::: {.callout-note}
## Finding 15
AMO dominance varies by cluster. C5 (warming Europe) has the highest AMO
coefficient ($\beta = 0.69$, partial $R^2 = 0.49$). C4 (decelerating North
America) has the lowest AMO effect but the strongest PDO effect
($\beta = -0.31$).
:::

```{r}
#| label: fig-teleconnection-attribution
#| fig-cap: "Teleconnection attribution by cluster: standardised coefficients, partial $R^2$, and annual teleconnection index histories."
#| fig-width: 11
#| fig-height: 8

# --- Shared cluster data ---
cluster_data <- prepare_cluster_data()
trend_sub <- cluster_data$trend_sub
trend_mat <- cluster_data$trend_mat
year_cols <- cluster_data$year_cols

tc <- read_csv(file.path(rdata, "teleconnection_indices.csv"), show_col_types = FALSE)
tc_names <- c("Nino34", "PDO", "AMO", "NAO", "AO")

tc_annual <- tc %>%
  mutate(year = ceiling(month / 12)) %>%
  group_by(year) %>%
  summarise(across(all_of(tc_names), ~ mean(.x, na.rm = TRUE)), .groups = "drop")

annual_trend <- trend_mat
n_lakes <- nrow(annual_trend)
n_sample <- min(10000, n_lakes)
set.seed(123)
sample_idx <- sample(n_lakes, n_sample)

# --- Level 1: Standardized coefficients ---
results_l1 <- data.frame(
  cluster_label = trend_sub$cluster_label[sample_idx],
  R2 = NA_real_,
  coef_Nino34 = NA_real_, coef_PDO = NA_real_, coef_AMO = NA_real_,
  coef_NAO = NA_real_, coef_AO = NA_real_
)

for (k in seq_along(sample_idx)) {
  i <- sample_idx[k]
  lake_y <- annual_trend[i, ]
  if (sum(is.finite(lake_y)) < 20) next
  df_reg <- data.frame(y = lake_y, Nino34 = tc_annual$Nino34, PDO = tc_annual$PDO,
                       AMO = tc_annual$AMO, NAO = tc_annual$NAO, AO = tc_annual$AO) %>%
    filter(is.finite(y))
  if (nrow(df_reg) < 20) next
  df_scaled <- as.data.frame(scale(df_reg))
  fit <- lm(y ~ Nino34 + PDO + AMO + NAO + AO, data = df_scaled)
  results_l1$R2[k] <- summary(lm(y ~ Nino34 + PDO + AMO + NAO + AO, data = df_reg))$r.squared
  coefs <- coef(fit)
  results_l1$coef_Nino34[k] <- coefs["Nino34"]
  results_l1$coef_PDO[k] <- coefs["PDO"]
  results_l1$coef_AMO[k] <- coefs["AMO"]
  results_l1$coef_NAO[k] <- coefs["NAO"]
  results_l1$coef_AO[k] <- coefs["AO"]
}

# --- Level 2: First-differenced ---
diff_trend <- t(diff(t(annual_trend)))

results_l2 <- data.frame(
  cluster_label = trend_sub$cluster_label[sample_idx],
  R2 = NA_real_,
  coef_Nino34 = NA_real_, coef_PDO = NA_real_, coef_AMO = NA_real_,
  coef_NAO = NA_real_, coef_AO = NA_real_
)

for (k in seq_along(sample_idx)) {
  i <- sample_idx[k]
  lake_y <- diff_trend[i, ]
  if (sum(is.finite(lake_y)) < 15) next
  df_reg <- data.frame(y = lake_y, Nino34 = tc_annual$Nino34[2:40], PDO = tc_annual$PDO[2:40],
                       AMO = tc_annual$AMO[2:40], NAO = tc_annual$NAO[2:40], AO = tc_annual$AO[2:40]) %>%
    filter(is.finite(y))
  if (nrow(df_reg) < 15) next
  df_scaled <- as.data.frame(scale(df_reg))
  fit <- lm(y ~ Nino34 + PDO + AMO + NAO + AO, data = df_scaled)
  results_l2$R2[k] <- summary(lm(y ~ Nino34 + PDO + AMO + NAO + AO, data = df_reg))$r.squared
  coefs <- coef(fit)
  results_l2$coef_Nino34[k] <- coefs["Nino34"]
  results_l2$coef_PDO[k] <- coefs["PDO"]
  results_l2$coef_AMO[k] <- coefs["AMO"]
  results_l2$coef_NAO[k] <- coefs["NAO"]
  results_l2$coef_AO[k] <- coefs["AO"]
}

# --- Heatmap 1: Level 1 standardized β ---
coef_l1 <- results_l1 %>%
  group_by(cluster_label) %>%
  summarise(across(starts_with("coef_"), ~ mean(.x, na.rm = TRUE)), .groups = "drop") %>%
  tidyr::pivot_longer(-cluster_label, names_to = "index", values_to = "beta") %>%
  mutate(index = gsub("coef_", "", index), index = factor(index, levels = tc_names),
         label = sprintf("%+.2f", beta))

p1 <- ggplot(coef_l1, aes(x = cluster_label, y = index, fill = beta)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(aes(label = label), size = 2.8, color = ifelse(abs(coef_l1$beta) > 0.15, "white", col_text)) +
  scale_atomic_diverging(name = expression(beta)) +
  scale_x_discrete(expand = c(0, 0)) + scale_y_discrete(expand = c(0, 0)) +
  labs(x = NULL, y = NULL, title = "Level 1: STL trend") +
  theme_atomic_base() +
  theme(axis.text.x = element_text(size = 7), axis.text.y = element_text(size = 7),
        plot.title = element_text(size = 9), legend.position = "right",
        legend.key.width = unit(0.35, "cm"), legend.key.height = unit(1.2, "cm"))

# --- Heatmap 2: Level 2 standardized β ---
coef_l2 <- results_l2 %>%
  group_by(cluster_label) %>%
  summarise(across(starts_with("coef_"), ~ mean(.x, na.rm = TRUE)), .groups = "drop") %>%
  tidyr::pivot_longer(-cluster_label, names_to = "index", values_to = "beta") %>%
  mutate(index = gsub("coef_", "", index), index = factor(index, levels = tc_names),
         label = sprintf("%+.2f", beta))

p2 <- ggplot(coef_l2, aes(x = cluster_label, y = index, fill = beta)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(aes(label = label), size = 2.8, color = ifelse(abs(coef_l2$beta) > 0.15, "white", col_text)) +
  scale_atomic_diverging(name = expression(beta)) +
  scale_x_discrete(expand = c(0, 0)) + scale_y_discrete(expand = c(0, 0)) +
  labs(x = NULL, y = NULL, title = "Level 2: diff(STL trend)") +
  theme_atomic_base() +
  theme(axis.text.x = element_text(size = 7), axis.text.y = element_text(size = 7),
        plot.title = element_text(size = 9), legend.position = "right",
        legend.key.width = unit(0.35, "cm"), legend.key.height = unit(1.2, "cm"))

# --- Heatmap 3: Level 1 partial R² ---
# Compute partial R²
results_partial <- data.frame(
  cluster_label = trend_sub$cluster_label[sample_idx],
  partial_R2_Nino34 = NA_real_, partial_R2_PDO = NA_real_,
  partial_R2_AMO = NA_real_, partial_R2_NAO = NA_real_, partial_R2_AO = NA_real_
)

for (k in seq_along(sample_idx)) {
  i <- sample_idx[k]
  lake_y <- annual_trend[i, ]
  if (sum(is.finite(lake_y)) < 20) next
  df_reg <- data.frame(y = lake_y, Nino34 = tc_annual$Nino34, PDO = tc_annual$PDO,
                       AMO = tc_annual$AMO, NAO = tc_annual$NAO, AO = tc_annual$AO) %>%
    filter(is.finite(y))
  if (nrow(df_reg) < 20) next
  fit_full <- lm(y ~ Nino34 + PDO + AMO + NAO + AO, data = df_reg)
  for (tc_name in tc_names) {
    reduced <- as.formula(paste("y ~", paste(setdiff(tc_names, tc_name), collapse = " + ")))
    fit_red <- lm(reduced, data = df_reg)
    ss_red <- sum(residuals(fit_red)^2)
    ss_full <- sum(residuals(fit_full)^2)
    results_partial[[paste0("partial_R2_", tc_name)]][k] <- (ss_red - ss_full) / ss_red
  }
}

partial_l1 <- results_partial %>%
  group_by(cluster_label) %>%
  summarise(across(starts_with("partial_R2_"), ~ mean(.x, na.rm = TRUE)), .groups = "drop") %>%
  tidyr::pivot_longer(-cluster_label, names_to = "index", values_to = "partial_R2") %>%
  mutate(index = gsub("partial_R2_", "", index), index = factor(index, levels = tc_names),
         label = sprintf("%.2f", partial_R2))

p3 <- ggplot(partial_l1, aes(x = cluster_label, y = index, fill = partial_R2)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(aes(label = label), size = 2.8, color = ifelse(partial_l1$partial_R2 > 0.08, "white", col_text)) +
  scale_fill_gradient(low = "#f7f7f7", high = pal_direction[["cooling"]], name = expression("Partial " * R^2), labels = percent_format(accuracy = 1)) +
  scale_x_discrete(expand = c(0, 0)) + scale_y_discrete(expand = c(0, 0)) +
  labs(x = NULL, y = NULL, title = "Partial R²") +
  theme_atomic_base() +
  theme(axis.text.x = element_text(size = 7), axis.text.y = element_text(size = 7),
        plot.title = element_text(size = 9), legend.position = "right",
        legend.key.width = unit(0.35, "cm"), legend.key.height = unit(1.2, "cm"))

# --- Panel 4: Teleconnection time series (5 subplots, bar chart) ---
tc_long <- tc_annual %>%
  tidyr::pivot_longer(-year, names_to = "index", values_to = "value") %>%
  mutate(
    index = factor(index, levels = tc_names),
    sign = ifelse(value >= 0, "positive", "negative")
  )

p4_list <- list()
tc_labels_vec <- c(Nino34 = "ENSO", PDO = "PDO", AMO = "AMO", NAO = "NAO", AO = "AO")

for (idx_name in tc_names) {
  sub <- tc_long %>% filter(index == idx_name)
  p <- ggplot(sub, aes(x = year, y = value, fill = sign)) +
    geom_col(width = 0.8, alpha = 0.7) +
    geom_hline(yintercept = 0, color = "grey30", linewidth = 0.25) +
    scale_fill_manual(values = c(positive = pal_direction[["warming"]], negative = pal_direction[["cooling"]]), guide = "none") +
    scale_x_continuous(breaks = seq(1985, 2020, by = 10), expand = c(0, 0)) +
    labs(title = tc_labels_vec[idx_name], x = NULL, y = NULL) +
    theme_atomic_base(grid = FALSE) +
    theme(
      plot.title = element_text(size = 8, face = "bold", hjust = 0.5),
      axis.text = element_text(size = 5.5),
      plot.margin = margin(1, 2, 1, 2)
    )
  p4_list[[idx_name]] <- p
}

p4 <- wrap_plots(p4_list, ncol = 1) +
  plot_annotation(
    title = "Teleconnection indices (annual means)",
    theme = theme(plot.title = element_text(size = 9, face = "bold", color = col_text))
  )

# --- Compose 2×2 ---
p_all <- (p1 | p2) / (p3 | p4) +
  plot_layout(heights = c(1, 1.3))

save_figure("fig-teleconnection-attribution.png", p_all, width = 11, height = 8)
p_all
```

## Level 2: Acceleration attribution

::: {.callout-note}
## Finding 16
The five indices explain less variance in the first-differenced STL trend
(mean $R^2 = 0.26$) than in the trend level (mean $R^2 = 0.43$). This suggests
that teleconnection indices are better at explaining long-term trend levels
than year-to-year changes in trend slope.
:::

::: {.callout-note}
## Finding 17
The standardised coefficients for the first-differenced model are uniformly
small ($|\beta| < 0.1$ at the global level). This indicates that no single
teleconnection index strongly drives the acceleration or deceleration of lake
warming at the global scale.
:::

## Interpretation

The strong AMO signal across all clusters is consistent with the known role of
Atlantic multidecadal variability in driving Northern Hemisphere temperature
trends. The weak ENSO signal is expected: ENSO is an interannual phenomenon
with limited persistence on the 40-year timescale of the STL trend.

The cluster-level variation in teleconnection response confirms that the eight
temporal response types capture genuinely different climate sensitivities, not
just statistical noise. Lakes in C5 (warming Europe) are most strongly
influenced by AMO, while lakes in C4 (decelerating North America) are most
strongly influenced by PDO.
