---
title: "PC2--PC3 Score-Space Density"
aliases:
  - "/explorations/warming-acceleration/investigations/pca-score-density.html"
---

```{r}
#| label: setup-pca-score-density
#| include: false
source("../../../R/core/style.R")
source("../../../R/figures/investigations/pca-score-density.R")
list2env(prepare_pca_score_density_data(), envir = environment())
```

## Lake count versus represented spatial cells

PCA is fitted to equal-area cell means, but each lake receives a projected
score on the fixed cell-PCA axes. The left panel therefore shows where sampled
**lakes** are concentrated in PC2--PC3 score space. The right panel shows the
573 represented equal-area cells, each with one vote in PCA fitting. Lake-count
density is not an estimate of the fraction of global lake area or geography.

> PCA 拟合于等面积格网均值，但每个湖都投影到固定格网 PCA 轴。左图显示抽样**湖泊**在 PC2–PC3 score 空间中的集中位置；右图为 573 个参与 PCA 拟合的等面积格网，每格一票。湖泊计数密度不代表全球湖面积或地理面积占比。

```{r}
#| label: fig-pc23-score-density
#| fig-cap: "PC2--PC3 score-space distribution of projected lakes and equal-area PCA cells. Lake bins encode log10 lake count; cell point area encodes the number of lakes represented by each cell. Axes are shared." 
#| fig-width: 10
#| fig-height: 4.5

lake_density <- ggplot(lake_scores, aes(pc2, pc3)) +
  stat_bin2d(bins = 55) +
  scale_fill_viridis_c(trans = "log10", name = "Lake count") +
  coord_cartesian(xlim = c(-limit, limit), ylim = c(-limit, limit), expand = FALSE) +
  labs(title = "Projected lakes", x = "PC2 score", y = "PC3 score") +
  theme_base(grid = TRUE) + theme_facet(base_size = 9) + theme(legend.position = "bottom")

cell_density <- ggplot(cell_scores, aes(pc2, pc3)) +
  geom_hline(yintercept = 0, colour = col_grid, linewidth = wid_grid) +
  geom_vline(xintercept = 0, colour = col_grid, linewidth = wid_grid) +
  geom_point(aes(size = n_lakes), colour = col_inchart_line, alpha = .62) +
  scale_size_continuous(trans = "sqrt", range = c(1, 5), name = "Lakes per cell") +
  coord_cartesian(xlim = c(-limit, limit), ylim = c(-limit, limit), expand = FALSE) +
  labs(title = "Equal-area PCA cells", x = "PC2 score", y = "PC3 score") +
  theme_base(grid = TRUE) + theme_facet(base_size = 9) + theme(legend.position = "bottom")

lake_density | cell_density
```

The figure is a distribution diagnostic. It can reveal dense cores, sparse
tails, and whether lake abundance is concentrated in a limited part of the
secondary score plane. It does not by itself define trajectory groups, select a
cluster count, or identify a physical process.

> 此图是分布诊断：可观察致密核心、稀疏尾部，以及湖泊数量是否集中在次级 score 平面的局部。它本身不定义轨迹类别、不选择 cluster 数，也不识别物理过程。

## Density-topology check

At increasingly broad KDE bandwidths, projected lakes retain two density peaks:
the secondary peak remains about `r sprintf("%.0f", 100 * (kde_peak_summary |> filter(source == "Projected lakes", bandwidth_factor == 1.1) |> pull(secondary_peak_relative)))`% of the primary peak at the middle bandwidth.
But this is not a stable two-group result. The equal-area cell distribution has
`r kde_peak_summary |> filter(source == "Equal-area cells", bandwidth_factor == .7) |> pull(n_peaks)` local peaks at a narrow bandwidth, `r kde_peak_summary |> filter(source == "Equal-area cells", bandwidth_factor == 1.1) |> pull(n_peaks)` at the middle bandwidth, and only `r kde_peak_summary |> filter(source == "Equal-area cells", bandwidth_factor == 1.6) |> pull(n_peaks)` at the broad bandwidth. Thus apparent islands depend strongly on smoothing once each represented spatial cell has equal weight.

> KDE 带宽加大后，投影湖泊仍保留两个密度峰，中等带宽下次峰约为主峰的 `r sprintf("%.0f", 100 * (kde_peak_summary |> filter(source == "Projected lakes", bandwidth_factor == 1.1) |> pull(secondary_peak_relative)))`%。但这不构成稳定的两组：等面积格网在窄／中／宽带宽下的局部峰数依次为 `r kde_peak_summary |> filter(source == "Equal-area cells", bandwidth_factor == .7) |> pull(n_peaks)`、`r kde_peak_summary |> filter(source == "Equal-area cells", bandwidth_factor == 1.1) |> pull(n_peaks)`、`r kde_peak_summary |> filter(source == "Equal-area cells", bandwidth_factor == 1.6) |> pull(n_peaks)`。每个空间格网等权后，表观岛屿强烈依赖平滑尺度。

This argues against restoring K-means or treating the two lake-density humps as
natural response types. They may still be useful display regions for a later
continuous score-plane analysis, but their boundary is not data-defined.

> 因此不应恢复 K-means，也不应将两个湖泊密度隆起视为自然响应类型。它们可作为后续连续 score 平面分析的展示区域，但边界不是数据定义的。

## Three-dimensional density surface

Both surfaces use the middle KDE bandwidth and scale their own maximum to one.
Height therefore compares shape within a distribution, not lake density against
cell density. Drag to rotate; the PC2 and PC3 axes are shared with @fig-pc23-score-density.

> 两个曲面均使用中等 KDE 带宽，并将各自最大值缩放为一。高度比较的是各自分布内部的形状，不能比较湖泊密度与格网密度的绝对大小。可拖拽旋转；PC2、PC3 坐标轴与 @fig-pc23-score-density 相同。

```{r}
#| label: fig-pc23-score-density-3d
#| fig-cap: "Interactive PC2--PC3 KDE surfaces. Lake and equal-area-cell KDE maxima are separately normalised to one; surfaces show score-space shape only."
#| fig-width: 12
#| fig-height: 5

make_surface <- function(surface, title) {
  plotly::plot_ly(
    x = surface$x, y = surface$y, z = surface$z,
    type = "surface",
    colorscale = list(c(0, "#440154"), c(.5, "#21908C"), c(1, "#FDE725")),
    showscale = FALSE
  ) |>
    plotly::layout(
      title = title,
      scene = list(
        xaxis = list(title = "PC2", range = c(-limit, limit)),
        yaxis = list(title = "PC3", range = c(-limit, limit)),
        zaxis = list(title = "Relative KDE density", range = c(0, 1))
      )
    )
}

plotly::subplot(
  make_surface(lake_kde_surface, "Projected lakes"),
  make_surface(cell_kde_surface, "Equal-area PCA cells"),
  nrows = 1, margin = .04, titleX = TRUE, titleY = TRUE
)
```

### Equal-area-cell surface alone

This full-width view isolates the score-space density of the 573 represented
cells. Black points are cell scores on the zero-density plane; their size is
the number of lakes in a cell, shown only to expose sampling concentration.

> 此全宽视图单独展示 573 个代表性格网的 score 空间密度。黑点为零密度平面的格网 score；点大小为格网湖泊数，仅用于显示抽样集中程度。

```{r}
#| label: fig-pc23-cell-density-3d
#| fig-cap: "Interactive equal-area-cell KDE surface in PC2--PC3 space. Surface height is relative cell density; base-plane point size shows lakes per cell and does not weight the density estimate."
#| fig-width: 10
#| fig-height: 7

plotly::plot_ly(
  x = cell_kde_surface$x, y = cell_kde_surface$y, z = cell_kde_surface$z,
  type = "surface",
  colorscale = list(c(0, "#440154"), c(.5, "#21908C"), c(1, "#FDE725")),
  opacity = .86,
  showscale = TRUE,
  colorbar = list(title = "Relative KDE density")
) |>
  plotly::add_markers(
    data = cell_scores,
    x = ~pc2, y = ~pc3, z = 0,
    marker = list(color = "black", opacity = .55),
    size = ~sqrt(n_lakes), sizes = c(3, 18),
    inherit = FALSE,
    name = "Equal-area cells"
  ) |>
  plotly::layout(
    scene = list(
      xaxis = list(title = "PC2", range = c(-limit, limit)),
      yaxis = list(title = "PC3", range = c(-limit, limit)),
      zaxis = list(title = "Relative cell KDE density", range = c(0, 1))
    ),
    showlegend = FALSE
  )
```

## Geographic provenance of lake-density cores

The two lake-density maxima are at approximately PC2/PC3 =
(`r sprintf("%.2f", lake_density_peaks$pc2[[1]])`, `r sprintf("%.2f", lake_density_peaks$pc3[[1]])`) and
(`r sprintf("%.2f", lake_density_peaks$pc2[[2]])`, `r sprintf("%.2f", lake_density_peaks$pc3[[2]])`). For display only, lakes above the KDE saddle
(`r sprintf("%.2f", density_saddle)` relative density) are assigned to their nearest maximum. This identifies geographic provenance of two dense score-space cores; it does **not** make them clusters or assign all lakes to a type.

> 两个湖泊密度峰约位于 PC2/PC3 = (`r sprintf("%.2f", lake_density_peaks$pc2[[1]])`, `r sprintf("%.2f", lake_density_peaks$pc3[[1]])`) 与 (`r sprintf("%.2f", lake_density_peaks$pc2[[2]])`, `r sprintf("%.2f", lake_density_peaks$pc3[[2]])`)。仅为展示，将高于 KDE 鞍点（相对密度 `r sprintf("%.2f", density_saddle)`）的湖泊归到最近峰。它识别两个高密度 score 核心的地理来源，**不**把它们变成 cluster，也不为所有湖泊赋类型。

```{r}
#| label: fig-pc23-density-core-map
#| fig-cap: "Geographic provenance of the two displayed lake-density cores in PC2--PC3 space. Core membership is a KDE-saddle display device, not a clustering result."
#| fig-width: 10
#| fig-height: 4.5

ggplot() +
  geom_tile(data = lake_core_map, aes(lon_cell + .5, lat_cell + .5, fill = n_lakes), width = 1, height = 1) +
  the_geom_coastline() +
  facet_wrap(~core, ncol = 2) +
  scale_fill_viridis_c(trans = "log10", name = "Lakes in core") +
  coord_cartesian(xlim = c(-180, 180), ylim = c(-60, 85), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") +
  theme_map_grid(base_size = 8) + theme(legend.position = "bottom")
```

The geographic map is the required next check. If both cores are dominated by
one lake-dense region, their two peaks are mainly a sampling-density feature. If
each core recurs in several separated regions, it becomes a descriptive clue
about recurring score combinations. Either outcome remains separate from a
claim of discrete response types.

> 地图是下一步必要检查。若两核心均由单一湖泊密集区主导，双峰主要是抽样密度特征；若每个核心在多个远距区域重复，才可作为可重复 score 组合的描述线索。两种结果都不等于离散响应类型。
